---
title: "Two API Styles: Direct and Generator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two API Styles: Direct and Generator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(rxsim)
set.seed(42)
```

## Introduction

rxsim supports two equivalent ways to build the same trial.

- **Style A: Direct instantiation** builds `Population`, `Timer`,
  `Condition`, and `Trial` objects yourself.
- **Style B: Generator API** gives `replicate_trial()` the population and
  analysis generators, then lets rxsim build each `Trial` for you.

The two styles are equivalent when the design itself is the same: same arms,
endpoint distributions, enrollment, dropout, and trigger logic. The difference
is where you want control.

Prefer the direct style when you want to inspect or edit the `Timer`, work with
fixed schedules, or debug one trial step by step. Prefer the generator style
when you want many stochastic replicates, quick scenario grids, or minimal
boilerplate.

The sections below use the same two-arm design in both styles:

- placebo (`pbo`) versus treatment (`trt`)
- `sample_size = 40` with `allocation = c(1, 1)`
- stochastic enrollment from `rexp(n, rate = 1)`
- stochastic dropout from `rexp(n, rate = 0.05)`
- continuous endpoint with mean `0` for placebo and `0.5` for treatment
- one final analysis at full enrollment
- `100` replicates for the operating characteristics

```{r shared-design}
sample_size <- 40L
arms <- c("pbo", "trt")
allocation <- c(1, 1)
delta <- 0.5
n_reps <- 100L

enrollment <- function(n) rexp(n, rate = 1)
dropout <- function(n) rexp(n, rate = 0.05)

make_outcome_data <- function(n, mean_shift) {
  data.frame(
    id = seq_len(n),
    outcome = rnorm(n, mean = mean_shift, sd = 1),
    readout_time = 0,
    stringsAsFactors = FALSE
  )
}

final_analysis <- function(df, current_time) {
  enrolled <- subset(df, !is.na(enroll_time))
  fit <- stats::t.test(outcome ~ arm, data = enrolled)

  data.frame(
    n = nrow(enrolled),
    mean_pbo = mean(enrolled$outcome[enrolled$arm == "pbo"]),
    mean_trt = mean(enrolled$outcome[enrolled$arm == "trt"]),
    p_value = unname(fit$p.value),
    stringsAsFactors = FALSE
  )
}

make_final_condition <- function() {
  Condition$new(
    where = enroll_trigger(1.0, sample_size),
    analysis = final_analysis,
    name = "final"
  )
}
```

## Style A: Direct instantiation

Use the direct style when you want to see every moving part. You create the
plan, build the timer, size each arm, define the trigger, and assemble the
trial yourself.

```{r direct-single}
# 1. Draw one stochastic enrollment and dropout plan.
direct_plan <- gen_plan(
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment,
  dropout = dropout
)

# 2. Build the timer from that plan.
direct_timer <- Timer$new(name = "direct_timer")
add_timepoints(direct_timer, direct_plan)

# 3. Compute the planned arm sizes from the timer input.
direct_n_by_arm <- vapply(
  arms,
  function(arm_name) {
    as.integer(sum(direct_plan$enroller[direct_plan$arm == arm_name]))
  },
  integer(1)
)

# 4. Instantiate each arm population explicitly.
pop_pbo <- Population$new(
  name = "pbo",
  data = make_outcome_data(direct_n_by_arm[["pbo"]], mean_shift = 0)
)
pop_trt <- Population$new(
  name = "trt",
  data = make_outcome_data(direct_n_by_arm[["trt"]], mean_shift = delta)
)

# 5. Define the final analysis with an rxsim_trigger object.
direct_final <- make_final_condition()

# 6. Assemble the Trial object.
direct_trial <- Trial$new(
  name = "direct_trial",
  timer = direct_timer,
  population = list(pop_pbo, pop_trt),
  conditions = list(direct_final)
)
```

In this style, the timer is a first-class object. You can inspect the time grid,
modify individual timepoints, or replace the whole schedule before running.

```{r direct-inspect}
head(direct_plan)
direct_timer$get_end_timepoint()
```

```{r direct-run, results=FALSE}
# 7. Run the single direct trial.
direct_trial$run()
```

```{r direct-one-result}
collect_results(direct_trial)
```

For many stochastic replicates, the direct pattern is usually wrapped in a small
constructor function.

```{r direct-builder}
build_direct_trial <- function(name) {
  # 1. Generate a fresh stochastic plan for this replicate.
  plan <- gen_plan(
    sample_size = sample_size,
    arms = arms,
    allocation = allocation,
    enrollment = enrollment,
    dropout = dropout
  )

  # 2. Turn that plan into a timer.
  timer <- Timer$new(name = paste0(name, "_timer"))
  add_timepoints(timer, plan)

  # 3. Size each arm from the generated plan.
  n_by_arm <- vapply(
    arms,
    function(arm_name) {
      as.integer(sum(plan$enroller[plan$arm == arm_name], na.rm = TRUE))
    },
    integer(1)
  )

  # 4. Instantiate populations explicitly.
  populations <- list(
    Population$new(
      name = "pbo",
      data = make_outcome_data(n_by_arm[["pbo"]], mean_shift = 0)
    ),
    Population$new(
      name = "trt",
      data = make_outcome_data(n_by_arm[["trt"]], mean_shift = delta)
    )
  )

  # 5. Create a fresh condition object for this replicate.
  conditions <- list(make_final_condition())

  # 6. Return the assembled trial.
  Trial$new(
    name = name,
    timer = timer,
    population = populations,
    conditions = conditions
  )
}
```

```{r direct-replicates, results=FALSE}
direct_trials <- lapply(seq_len(n_reps), function(i) {
  build_direct_trial(paste0("direct_", i))
})

run_trials(direct_trials)
```

```{r direct-results}
direct_results <- collect_results(direct_trials)
head(direct_results)
```

## Style B: Generator API

Use the generator API when you want the same design with less setup. You define
functions for the arm populations and analyses, then let `replicate_trial()`
build the timers, conditions, and trials internally.

```{r generator-spec}
# 1. Define one population generator per arm.
population_generators <- list(
  pbo = function(n) make_outcome_data(n, mean_shift = 0),
  trt = function(n) make_outcome_data(n, mean_shift = delta)
)

# 2. Define the analysis generator with an rxsim_trigger object.
analysis_generators <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = final_analysis
  )
)
```

The trigger must be an `rxsim_trigger` object such as
`enroll_trigger(1.0, sample_size)`. There is no need for `rlang::exprs()` here.

```{r generator-replicates, results=FALSE}
# 3. Ask rxsim to build 100 independent trials.
generator_trials <- replicate_trial(
  trial_name = "generator",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment,
  dropout = dropout,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = n_reps
)

# 4. Run all generated trials.
run_trials(generator_trials)
```

```{r generator-results}
generator_results <- collect_results(generator_trials)
head(generator_results)
```

This style is shorter because `replicate_trial()` handles the timer
construction, population sizing, and condition creation for every replicate.

## Side-by-side results comparison

Because the data-generating assumptions are the same, both APIs should produce
matching operating-characteristic distributions. The rows are not identical
replicate by replicate, but the summaries should line up.

```{r comparison-summary}
comparison_results <- rbind(
  transform(direct_results, style = "Direct"),
  transform(generator_results, style = "Generator")
)
comparison_results$effect_hat <-
  comparison_results$mean_trt - comparison_results$mean_pbo

comparison_summary <- do.call(
  rbind,
  lapply(split(comparison_results, comparison_results$style), function(x) {
    data.frame(
      style = unique(x$style),
      mean_effect = mean(x$effect_hat),
      sd_effect = sd(x$effect_hat),
      mean_p_value = mean(x$p_value),
      q25_p_value = unname(stats::quantile(x$p_value, 0.25)),
      q50_p_value = unname(stats::quantile(x$p_value, 0.50)),
      q75_p_value = unname(stats::quantile(x$p_value, 0.75)),
      stringsAsFactors = FALSE
    )
  })
)

comparison_summary
```

```{r comparison-plots, fig.width=8, fig.height=4}
op <- par(mfrow = c(1, 2))
boxplot(effect_hat ~ style, data = comparison_results,
        main = "Estimated treatment effect")
boxplot(p_value ~ style, data = comparison_results,
        main = "Final p-value")
par(op)
```

The two summaries should be very close. That is the main practical point:
choose the style that fits how much control you need before the simulation runs.

## When to use Style A

Prefer direct instantiation when you want to work directly with the underlying
objects.

- Manually manipulate the `Timer` before running.
- Use deterministic schedules from `gen_timepoints()` or a hand-built plan.
- Debug a single trial by inspecting `timer`, `population`, and `locked_data`.
- Build multi-stage designs where the schedule is easier to express directly,
  such as the seamless pattern in [Example 8](example-8.html).

## When to use Style B

Prefer the generator API when the design follows the standard workflow that
`replicate_trial()` already knows how to build.

- Run scenario grids generated with `expand_grid()`.
- Simulate standard stochastic enrollment and dropout with minimal setup.
- Write scripts that need many replicates and little boilerplate.
- Keep design changes local to a few small generator functions.

## Mixing the styles

A common workflow is to start with the direct style for one trial, inspect it,
then scale out with `run_trials()` once you are happy with the setup.

```{r mixing-styles, results=FALSE}
# 1. Build one trial directly so you can inspect it.
mixed_trial <- build_direct_trial("mixed_1")
head(do.call(rbind, mixed_trial$timer$timelist))

# 2. Extend the same direct pattern to more replicates.
mixed_trials <- c(
  list(mixed_trial),
  lapply(2:5, function(i) build_direct_trial(paste0("mixed_", i)))
)

# 3. Run them together with the usual batch helper.
run_trials(mixed_trials)
```

If you intentionally want the same fixed timer and populations in each
replicate, `clone_trial()` is also useful. That pattern is most natural for
deterministic schedules, not for stochastic enrollment.

## Next steps

- Add interims and multiple analyses: [Example 2](example-2.html)
- Learn more about trigger objects and analyses:
  [Condition](class-condition.html)
- Inspect the simulation loop and result structure:
  [Trial](class-trial.html)
