---
title: "Population: Subject-Level Data and Enrollment Tracking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Population: Subject-Level Data and Enrollment Tracking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rxsim)
```

## What is a Population?

A `Population` is the subject-level state for one trial arm. It owns the arm's
endpoint data and two tracking vectors: one for enrollment times and one for
dropout times. In practice, it is the object that answers three questions at
once: who belongs to this arm, what data does each subject contribute, and what
has happened to each subject so far?

The example below starts with a single-readout arm. No subject has enrolled or
dropped yet, so both state vectors start as `NA`.

```{r mental-model}
set.seed(1001)

subject_data <- data.frame(
  id = 1:4,
  readout_time = 0,
  response = c(11.2, 9.8, 10.5, 12.1)
)

pop <- Population$new(name = "active", data = subject_data)

list(
  arm = pop$name,
  subjects = pop$n,
  readouts_per_subject = pop$n_readouts,
  enrolled = pop$enrolled,
  dropped = pop$dropped
)
```

## Data frame format

`Population$data` must be a data frame with three structural columns plus at
least one endpoint column.

| Column | Type | Meaning |
|--------|------|---------|
| `id` | integer | Unique subject identifier within the arm |
| `arm` | character | Arm label; auto-filled from `name` if absent |
| `readout_time` | numeric | Time from enrollment to the readout |
| endpoint | any | At least one endpoint column such as `response` |

`readout_time` is relative to enrollment, not calendar time. A subject can have
one row or several rows, depending on how many scheduled readouts you need to
store.

Here `subject_data` omitted `arm`, so `Population$new()` filled it in from the
object name.

```{r data-format}
pop$data
```

## Creating a Population

You can construct a `Population` directly from a data frame when you already
have subject-level rows. For quick single-endpoint examples,
`as_population_data()` is a compact shortcut that creates `id`, `data`, and
`readout_time` so that `Population$new()` can fill the arm label.

The first object reuses the explicit data frame from above. The second shows the
vector shortcut for a placebo arm.

```{r creating}
pop_df <- Population$new(name = "active", data = subject_data)

pop_vector <- Population$new(
  name = "placebo",
  data = as_population_data(c(8.7, 9.1, 8.9, 9.4))
)

pop_df$data
pop_vector$data
```

## Enrollment and dropout tracking

The `enrolled` and `dropped` fields are numeric vectors with one entry per
subject. `NA` means the event has not happened yet. `set_enrolled(n, time)`
randomly chooses `n` currently unenrolled subjects and stamps them with an
enrollment time. `set_dropped(n, time)` randomly chooses from subjects who are
enrolled and not yet dropped.

A step-by-step example makes the bookkeeping visible.

```{r enrollment-dropout}
set.seed(2024)

state_0 <- data.frame(
  id = seq_len(pop$n),
  enrolled = pop$enrolled,
  dropped = pop$dropped
)

pop$set_enrolled(n = 2, time = 1)
state_1 <- data.frame(
  id = seq_len(pop$n),
  enrolled = pop$enrolled,
  dropped = pop$dropped
)

pop$set_enrolled(n = 1, time = 3)
state_2 <- data.frame(
  id = seq_len(pop$n),
  enrolled = pop$enrolled,
  dropped = pop$dropped
)

pop$set_dropped(n = 1, time = 5)
state_3 <- data.frame(
  id = seq_len(pop$n),
  enrolled = pop$enrolled,
  dropped = pop$dropped
)

state_0
state_1
state_2
state_3
```

## Multiple readout times

When a subject contributes more than one scheduled measurement, store multiple
rows for the same `id`. `Population$n` still counts unique subjects, while
`Population$n_readouts` records how many rows belong to each subject.

The data below uses two readouts per subject: one at enrollment and one eight
units later.

```{r multiple-readouts}
multi_readout_data <- data.frame(
  id = rep(1:4, each = 2),
  readout_time = rep(c(0, 8), times = 4),
  response = c(11.2, 10.8, 9.8, 9.4, 10.5, 10.0, 12.1, 11.6)
)

multi_pop <- Population$new(name = "active", data = multi_readout_data)

multi_pop$n
multi_pop$n_readouts
multi_pop$data
```

## Endpoint types

`Population` itself is endpoint-agnostic: it stores rows and metadata, not a
special endpoint class. Endpoint type is determined by which endpoint columns
you include in the data frame.

| Endpoint type | Typical data construction | Example column |
|---|---|---|
| Continuous | `rnorm()` draws in a numeric column | `response` |
| Binary | `rbinom(..., size = 1, prob = p)` as 0/1 | `response_bin` |
| Time-to-event | `rexp()` / Weibull draws for event times | `tte` |

This matters because the analysis function in each `Condition` must match the
endpoint scale: t-test or linear model for continuous endpoints, proportion or
logistic methods for binary endpoints, and survival methods for time-to-event
outcomes.

```{r endpoint-types}
set.seed(4040)

pop_cont <- Population$new(
  name = "cont",
  data = data.frame(
    id = 1:5,
    response = rnorm(5, mean = 0, sd = 1),
    readout_time = 1
  )
)

pop_bin <- Population$new(
  name = "bin",
  data = data.frame(
    id = 1:5,
    response_bin = rbinom(5, size = 1, prob = 0.35),
    readout_time = 1
  )
)

pop_tte <- Population$new(
  name = "tte",
  data = data.frame(
    id = 1:5,
    tte = rexp(5, rate = 0.2),
    readout_time = 1
  )
)

names(pop_cont$data)
names(pop_bin$data)
names(pop_tte$data)
```

## Role in Trial

`Population` is one arm-level building block inside `Trial`. At each timer
update, `Trial$run()` calls `set_enrolled()` and `set_dropped()` for each arm,
then builds a locked snapshot from the currently enrolled subjects. That
snapshot carries forward the original population columns and adds
`enroll_time`, `drop_time`, `measurement_time`, and the current trial `time`.

This example defines one simple analysis so that the trial stores a
`locked_data` snapshot at calendar time 5.

```{r role-in-trial}
set.seed(303)

trial_pop <- Population$new(name = "active", data = subject_data)

timer <- Timer$new(name = "population_timer")
timer$add_timepoint(time = 1, arm = "active", enroll = 2L, drop = 0L)
timer$add_timepoint(time = 3, arm = "active", enroll = 1L, drop = 0L)
timer$add_timepoint(time = 5, arm = "active", enroll = 0L, drop = 1L)

snapshot_condition <- Condition$new(
  where = calendar_trigger(5),
  analysis = function(df, current_time) {
    data.frame(
      current_time = current_time,
      n_subjects = length(unique(df$id))
    )
  },
  name = "snapshot"
)

trial <- Trial$new(
  name = "population_trial",
  seed = 303,
  timer = timer,
  population = list(trial_pop),
  conditions = list(snapshot_condition)
)

trial$run()

trial$locked_data$time_5[
  , c(
    "id",
    "arm",
    "readout_time",
    "response",
    "enroll_time",
    "drop_time",
    "measurement_time",
    "time"
  )
]

trial$results$time_5$snapshot
```

## Tips and gotchas

A few rules matter when you build populations by hand. `id` should identify the
subject, not the row, so repeated readouts reuse the same `id`. Omitting `arm`
is fine because `Population$new()` fills it from `name`. Finally, the data frame
must have at least four columns after `arm` is added: the three structural
columns and at least one endpoint column.

The checks below summarise those rules on the objects already created above.

```{r tips-gotchas}
data.frame(
  check = c(
    "Unique subjects in multi_pop",
    "Rows per subject in multi_pop",
    "Arm auto-filled in pop",
    "Minimum column count met"
  ),
  value = c(
    multi_pop$n,
    paste(unique(as.integer(table(multi_pop$data$id))), collapse = ", "),
    all(pop$data$arm == pop$name),
    ncol(pop$data) >= 4
  )
)
```

## Next steps

To see how `Population` fits into the broader workflow, continue with:

- [Enrollment and Dropout](enrollment.html): stochastic and deterministic enrollment, Timer reference
- [Trial reference](../reference/Trial.html): orchestration across population,
  timer, and conditions
