---
title: "Timer: Managing the Trial Clock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Timer: Managing the Trial Clock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)

timer_as_df <- function(timer) {
  out <- do.call(rbind, lapply(timer$timelist, as.data.frame))
  rownames(out) <- NULL
  out[order(out$time, out$arm), ]
}
```

## What is a Timer?

A `Timer` is the trial clock. It stores a list of timepoint records, where each
record says when something happens in one arm and how many subjects are
affected. Each record has four fields: `time`, `arm`, `enroller`, and
`dropper`.

`Trial$run()` does not step through subjects one by one. Instead, it collects
all distinct event times from the timer, sorts them, and updates the trial at
those times. That makes `Timer` the object that defines the calendar of the
simulation.

```{r timer_mental_model}
trial_clock <- Timer$new(name = "trial_clock")
trial_clock$add_timepoint(time = 1, arm = "pbo", dropper = 0L, enroller = 3L)
trial_clock$add_timepoint(time = 1, arm = "trt", dropper = 0L, enroller = 3L)
trial_clock$add_timepoint(time = 4, arm = "pbo", dropper = 1L, enroller = 0L)

trial_clock$timelist
sort(unique(vapply(trial_clock$timelist, function(x) x$time, numeric(1))))
```

## Building a Timer manually

Manual construction is useful when you want complete control over the event
schedule. You create a `Timer`, add one event at a time with
`add_timepoint()`, and then inspect it with the query methods.

The example below creates a two-arm timer with three timepoints per arm.

```{r manual_timer_build}
manual_timer <- Timer$new(name = "manual_timer")

manual_timer$add_timepoint(time = 1, arm = "pbo", dropper = 0L, enroller = 3L)
manual_timer$add_timepoint(time = 2, arm = "pbo", dropper = 1L, enroller = 2L)
manual_timer$add_timepoint(time = 4, arm = "pbo", dropper = 1L, enroller = 0L)

manual_timer$add_timepoint(time = 1, arm = "trt", dropper = 0L, enroller = 2L)
manual_timer$add_timepoint(time = 3, arm = "trt", dropper = 0L, enroller = 3L)
manual_timer$add_timepoint(time = 4, arm = "trt", dropper = 1L, enroller = 0L)

timer_as_df(manual_timer)
```

Once the timer exists, the accessors give quick structural summaries.

```{r manual_timer_queries}
manual_timer$get_end_timepoint()
manual_timer$get_n_arms()
sort(manual_timer$get_unique_times())
manual_timer$get_timepoint("trt", 3)
```

## Stochastic enrollment: `gen_plan()`

`gen_plan()` creates a random event schedule. The `enrollment` and `dropout`
arguments are functions that return inter-arrival times. For enrollment,
`function(n) rexp(n, rate = 1)` means: draw `n` waiting times from an
exponential distribution, then take cumulative sums to obtain calendar times.

This is the natural choice when enrollment timing is itself uncertain. Each
call draws a fresh schedule, which is why `replicate_trial()` can give each
replicate a different trial clock.

In the current `rxsim` implementation, `gen_plan()` returns a unified event
schedule with columns `time`, `arm`, `enroller`, and `dropper`. Rows where
`enroller == 1L` are enrollment events; rows where `dropper == 1L` are dropout
events.

```{r gen_plan_example}
set.seed(2024)

plan_stoch <- gen_plan(
  sample_size = 8,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 1),
  dropout = function(n) rexp(n, rate = 0.15)
)

head(plan_stoch, 8)
```

Passing the result to `add_timepoints()` turns the data frame into a `Timer`.

```{r gen_plan_to_timer}
stochastic_timer <- Timer$new(name = "stochastic_timer")
add_timepoints(stochastic_timer, plan_stoch)

head(stochastic_timer$get_unique_times())
stochastic_timer$get_end_timepoint()
```

A second call produces a different schedule because the inter-arrival times are
redrawn.

```{r gen_plan_repeat}
set.seed(2025)

plan_stoch_a <- gen_plan(
  sample_size = 8,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 1),
  dropout = function(n) rexp(n, rate = 0.15)
)

plan_stoch_b <- gen_plan(
  sample_size = 8,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 1),
  dropout = function(n) rexp(n, rate = 0.15)
)

identical(plan_stoch_a, plan_stoch_b)
```

## Deterministic enrollment: `gen_timepoints()`

`gen_timepoints()` creates a fixed schedule from piecewise-constant rates. The
`enrollment` and `dropout` arguments are lists with `end_time` and `rate`
elements. Conceptually, each rate applies until the next end time is reached.

Because the inputs are deterministic, repeated calls with the same values give
exactly the same timer. This is useful when you want a reproducible operating
plan and do not want timing variability to differ across replicates.

```{r gen_timepoints_example}
enrollment_schedule <- list(
  end_time = c(2, 4, 6),
  rate = c(2, 2, 2)
)

dropout_schedule <- list(
  end_time = c(2, 4, 6),
  rate = c(0, 2, 2)
)

plan_det <- gen_timepoints(
  sample_size = 12,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = enrollment_schedule,
  dropout = dropout_schedule
)

plan_det
```

The resulting schedule can be added directly to a timer.

```{r gen_timepoints_to_timer}
deterministic_timer <- Timer$new(name = "deterministic_timer")
add_timepoints(deterministic_timer, plan_det)

deterministic_timer$get_unique_times()
deterministic_timer$get_end_timepoint()
```

Calling `gen_timepoints()` again with the same inputs returns the same result.

```{r gen_timepoints_repeat}
plan_det_2 <- gen_timepoints(
  sample_size = 12,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = enrollment_schedule,
  dropout = dropout_schedule
)

identical(plan_det, plan_det_2)
```

## Comparison table

The two generators populate the same `Timer` interface, but they encode very
different assumptions.

| Feature | `gen_plan()` | `gen_timepoints()` |
|:--|:--|:--|
| Time scale | Usually continuous | Usually integer |
| Variability | Random each call | Deterministic |
| Event rows | One event per row | Counts can exceed 1 |
| Best use | Replicate-specific timing uncertainty | Fixed operational plan |
| Typical role | Study duration varies across runs | Study duration held fixed |

## Allocation ratios

The `allocation` vector gives relative weights across arms. rxsim converts the
weights to proportions, computes per-arm targets with `round()`, and then adds
or removes any leftover subjects so that the final total still equals
`sample_size`.

A three-arm example makes the rounding step easy to see. With allocation
`c(2, 1, 1)` and `sample_size = 11`, simple rounding gives `c(6, 3, 3)`, which
sums to 12 rather than 11. rxsim corrects that discrepancy when it generates
the schedule.

```{r allocation_example}
set.seed(88)

round(c(2, 1, 1) / sum(c(2, 1, 1)) * 11)

allocation_plan <- gen_plan(
  sample_size = 11,
  arms = c("pbo", "dose1", "dose2"),
  allocation = c(2, 1, 1),
  enrollment = function(n) rexp(n, rate = 1),
  dropout = function(n) rexp(n, rate = 0.15)
)

table(allocation_plan$arm[allocation_plan$enroller == 1L])
```

The exact arm that gains or loses a subject can depend on the random draw used
for the correction step. Setting a seed makes the example reproducible.

## Querying a Timer

The same query methods work whether the timer was built manually or generated by
one of the helper functions.

```{r timer_queries}
list(
  manual_end = manual_timer$get_end_timepoint(),
  manual_arms = manual_timer$get_n_arms(),
  manual_unique_times = sort(manual_timer$get_unique_times()),
  stochastic_end = stochastic_timer$get_end_timepoint(),
  deterministic_lookup = deterministic_timer$get_timepoint("trt", 4)
)
```

Use these helpers when you need to validate a design before running a trial or
when you want to inspect the generated calendar after `gen_plan()` or
`gen_timepoints()`.

## Role in Trial

`Timer` matters because `Trial$run()` uses it as the execution calendar. At
each unique trial time, populations are updated for enrollment and dropout,
then the current snapshot is passed to the analysis conditions.

The example below uses the manual timer from above and asks for a simple summary
at every unique time. The condition is allowed to fire repeatedly by setting
`max_triggers = Inf`.

```{r timer_in_trial}
pop_pbo <- Population$new(
  name = "pbo",
  data = data.frame(
    id = seq_len(5),
    arm = "pbo",
    readout_time = 0,
    response = rnorm(5)
  )
)

pop_trt <- Population$new(
  name = "trt",
  data = data.frame(
    id = seq_len(5),
    arm = "trt",
    readout_time = 0,
    response = rnorm(5, mean = 0.5)
  )
)

clock_condition <- Condition$new(
  where = calendar_trigger(sort(manual_timer$get_unique_times())),
  analysis = function(df, current_time) {
    data.frame(
      time = current_time,
      n_enrolled = sum(!is.na(df$enroll_time)),
      n_dropped = sum(!is.na(df$drop_time))
    )
  },
  name = "clock",
  max_triggers = Inf
)

trial <- Trial$new(
  name = "timer_demo",
  seed = 123,
  timer = manual_timer,
  population = list(pop_pbo, pop_trt),
  conditions = list(clock_condition)
)

trial$run()

names(trial$locked_data)
do.call(rbind, lapply(trial$results, function(x) x$clock))
```

This is the key relationship to remember: `Timer` defines when the trial moves;
`Population` defines who is available to move; `Condition` defines what gets
checked when the clock advances.

## Next steps

- [Population deep dive](class-population.html)
- [Condition deep dive](class-condition.html)
- [Enrollment and dropout modeling](enrollment-dropout.html)
