---
title: "Condition: Triggers and Analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Condition: Triggers and Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

This vignette is a deep-dive reference for the `Condition` R6 class and the
trigger API. Every example builds a small snapshot data frame by hand so you can
inspect `check_conditions()` directly before wiring conditions into a full
`Trial`.

```{r snapshots}
make_snapshot <- function(current_time, enroll_time,
                          drop_time = rep(NA_real_, 6)) {
  data.frame(
    id = 1:6,
    arm = c(
      "control", "control", "control",
      "treatment", "treatment", "treatment"
    ),
    response = c(7.2, 6.8, 8.1, 9.4, 10.1, 9.7),
    biomarker = c(0.30, 0.50, 0.40, 0.80, 0.90, 0.70),
    enroll_time = enroll_time,
    drop_time = drop_time,
    time = current_time,
    stringsAsFactors = FALSE
  )
}

snapshot_half <- make_snapshot(
  current_time = 8,
  enroll_time = c(1, 2, 4, NA, NA, NA)
)

snapshot_week12 <- make_snapshot(
  current_time = 12,
  enroll_time = c(1, 2, 4, 6, 8, 10)
)

snapshot_week16 <- make_snapshot(
  current_time = 16,
  enroll_time = c(1, 2, 4, 6, 8, 10),
  drop_time = c(NA, NA, 15, NA, NA, 14)
)

count_rows <- function(df, current_time) {
  data.frame(
    n = nrow(df),
    fired_at = current_time
  )
}
```

## What is a Condition?

A `Condition` is three things packaged into one stateful R6 object:

1. a **filter** built from trigger primitives
2. an optional **analysis** function
3. internal **state** that remembers whether it has already fired

`check_conditions(locked_data, current_time)` applies three gates in order:

1. the filtered snapshot must be non-empty
2. the cooldown must have elapsed
3. `trigger_count` must still be below `max_triggers`

If all three gates pass, the condition runs its analysis and returns a named
list. The name is taken from `name`; if `name` is `NULL`, the key falls back to
`1L`.

```{r what-is-a-condition}
arm_summary <- Condition$new(
  where = value_trigger("arm", "==", "treatment"),
  analysis = function(df, current_time) {
    data.frame(
      arm = unique(df$arm),
      n = nrow(df),
      mean_response = mean(df$response),
      fired_at = current_time
    )
  },
  name = "treatment_summary",
  cooldown = 0,
  max_triggers = 2L
)

arm_summary$check_conditions(snapshot_week12, current_time = 12)
```

## The locked_data snapshot

The snapshot passed to `check_conditions()` is the same data frame your analysis
function receives as `df`. It contains the subject-level columns from
`Population$data`, plus trial bookkeeping columns added by rxsim.

| Column group | Meaning |
|--------------|---------|
| Endpoint columns | Any subject-level columns already in `Population$data`, such as `id`, `response`, `biomarker`, covariates, or readout variables |
| `arm` | Arm label for each row |
| `enroll_time` | Calendar time of enrollment; `NA` means not yet enrolled |
| `drop_time` | Calendar time of dropout; `NA` means not dropped |
| `time` | Current calendar time of the snapshot |

In the manual examples below, rows stay in the snapshot and use `NA` in
`enroll_time` until a subject enrolls. Dropped subjects can remain visible, with
`drop_time` marking when they left.

```{r locked-data-snapshot}
snapshot_week16
```

In practice, think of `locked_data` as the current frozen trial view. Your
trigger reads columns from that snapshot, and your analysis works on the subset
that survived the trigger filter.

## Trigger primitives

rxsim provides four trigger constructors:

- `value_trigger(col, op, rhs)`
- `count_trigger(col, op, rhs)`
- `enroll_trigger(fraction, sample_size)`
- `calendar_trigger(cal_time)`

Valid operators are `>=`, `<=`, `>`, `<`, `==`, `!=`, and `%in%`.

### `value_trigger()`

`value_trigger()` compares a column value to a right-hand side. This is the most
natural choice for row-wise filters such as arm, subgroup, or calendar time.

```{r value-trigger}
value_cond <- Condition$new(
  where = value_trigger("arm", "==", "treatment"),
  analysis = count_rows,
  name = "treatment_rows"
)

value_cond$check_conditions(snapshot_week12, current_time = 12)
```

### `count_trigger()`

`count_trigger()` compares the count of non-`NA` values in a column. This is a
cohort-level gate: when it passes, the whole snapshot is eligible unless you
combine it with another row-wise trigger.

```{r count-trigger}
count_cond <- Condition$new(
  where = count_trigger("enroll_time", ">=", 4),
  analysis = function(df, current_time) {
    data.frame(
      enrolled = sum(!is.na(df$enroll_time)),
      n_rows_passed = nrow(df),
      fired_at = current_time
    )
  },
  name = "four_enrolled"
)

count_cond$check_conditions(snapshot_half, current_time = 8)
count_cond$check_conditions(snapshot_week12, current_time = 12)
```

### `enroll_trigger()`

`enroll_trigger()` is a shortcut for the common question "has a fraction of the
planned sample size enrolled yet?"

```{r enroll-trigger}
enroll_cond <- Condition$new(
  where = enroll_trigger(fraction = 0.5, sample_size = 6),
  analysis = function(df, current_time) {
    data.frame(
      enrolled = sum(!is.na(df$enroll_time)),
      fired_at = current_time
    )
  },
  name = "half_enrolled"
)

enroll_cond$check_conditions(snapshot_half, current_time = 8)
```

### `calendar_trigger()`

`calendar_trigger()` is a shortcut for fixed calendar looks.

```{r calendar-trigger}
calendar_cond <- Condition$new(
  where = calendar_trigger(12),
  analysis = count_rows,
  name = "week12_lock"
)

calendar_cond$check_conditions(snapshot_half, current_time = 8)
calendar_cond$check_conditions(snapshot_week12, current_time = 12)
```

## Combining triggers

Triggers compose with `&` and `|`.

- `trigger1 & trigger2`: both conditions must hold
- `trigger1 | trigger2`: either condition may hold

A common pattern is to combine a cohort threshold with a calendar rule.

```{r combining-and}
combo_and <- Condition$new(
  where = enroll_trigger(0.5, 6) & value_trigger("time", ">=", 12),
  analysis = function(df, current_time) {
    data.frame(
      enrolled = sum(!is.na(df$enroll_time)),
      fired_at = current_time
    )
  },
  name = "half_and_past_week12"
)

combo_and$check_conditions(snapshot_half, current_time = 8)
combo_and$check_conditions(snapshot_week12, current_time = 12)
```

`|` is useful for scheduled looks that can happen at more than one time.

```{r combining-or}
combo_or <- Condition$new(
  where = calendar_trigger(8) | calendar_trigger(12),
  analysis = count_rows,
  name = "scheduled_look",
  max_triggers = Inf
)

combo_or$check_conditions(snapshot_half, current_time = 8)
combo_or$check_conditions(snapshot_week12, current_time = 12)
```

## The analysis function

A condition analysis is called as `function(df, current_time, ...)`.

- `df` is already pre-filtered by the trigger
- `current_time` is the calendar time that fired the condition
- `...` is where values from `analysis_args` are injected

You can return any R object, but a one-row `data.frame` is the easiest shape to
bind across many trials and timepoints.

```{r analysis-function}
mean_cond <- Condition$new(
  where = value_trigger("arm", "==", "treatment"),
  analysis = function(df, current_time) {
    data.frame(
      analysis = "treatment_mean",
      n = nrow(df),
      mean_response = mean(df$response),
      current_time = current_time
    )
  },
  name = "treatment_mean"
)

mean_cond$check_conditions(snapshot_week12, current_time = 12)
```

You do not need to subset again for the same condition inside the analysis.
That work has already been done for you. If you need scenario-specific constants
or thresholds, pass them with `analysis_args`; see
[Using analysis_args](#using-analysis_args).

## Trigger state

Each `Condition` tracks its own state:

- `max_triggers`: hard cap on the number of fires
- `cooldown`: minimum time between successful fires
- `trigger_count`: number of successful fires so far
- `last_trigger_time`: time of the most recent fire

The helper below runs the same condition across a series of timepoints and shows
how the state changes.

```{r trigger-state-helper}
run_times <- function(cond, times) {
  out <- lapply(times, function(tt) {
    snap <- snapshot_week12
    snap$time <- tt
    fired <- cond$check_conditions(snap, current_time = tt)

    data.frame(
      time = tt,
      fired = length(fired) > 0,
      trigger_count = cond$trigger_count,
      last_trigger_time = cond$last_trigger_time
    )
  })

  do.call(rbind, out)
}
```

### `max_triggers = 1`: a final analysis

```{r trigger-state-once}
final_once <- Condition$new(
  where = value_trigger("time", ">=", 12),
  analysis = count_rows,
  name = "final_once",
  max_triggers = 1L
)

run_times(final_once, c(12, 16, 20))
```

### `max_triggers = Inf`: repeated logging

```{r trigger-state-inf}
logger <- Condition$new(
  where = value_trigger("time", ">=", 12),
  analysis = count_rows,
  name = "logger",
  cooldown = 2,
  max_triggers = Inf
)

run_times(logger, c(12, 13, 14, 16))
```

### `max_triggers = 2`: interim plus final

```{r trigger-state-two}
two_looks <- Condition$new(
  where = value_trigger("time", ">=", 12),
  analysis = count_rows,
  name = "two_looks",
  cooldown = 4,
  max_triggers = 2L
)

run_times(two_looks, c(12, 14, 16, 20))
```

Use `max_triggers = 1L` for a single final look, `Inf` for diagnostics or
logging at every eligible timepoint, and a small integer such as `2L` for an
interim-plus-final design.

## Convenience constructors

Use `trigger_by_calendar()` and `trigger_by_fraction()` when you want a single
common trigger without writing `Condition$new()` by hand.

`trigger_by_calendar()` is ideal for fixed looks.

```{r trigger-by-calendar}
by_calendar <- trigger_by_calendar(
  cal_time = 12,
  analysis = function(df, current_time) {
    data.frame(
      n = nrow(df),
      fired_at = current_time
    )
  },
  name = "week12"
)

by_calendar$check_conditions(snapshot_week12, current_time = 12)
```

`trigger_by_fraction()` is ideal for enrollment milestones.

```{r trigger-by-fraction}
by_fraction <- trigger_by_fraction(
  fraction = 0.5,
  sample_size = 6,
  analysis = function(df, current_time) {
    data.frame(
      enrolled = sum(!is.na(df$enroll_time)),
      fired_at = current_time
    )
  },
  name = "half"
)

by_fraction$check_conditions(snapshot_half, current_time = 8)
```

Reach for `Condition$new()` when you need custom `cooldown`, custom
`max_triggers`, `analysis_args`, or combined trigger logic.

## Attaching to a Trial

A `Trial` evaluates every condition in `conditions = list(...)` at every
scheduled timepoint. Multiple conditions can fire at different timepoints, and
each condition contributes one named result entry.

```{r attaching-to-trial}
set.seed(123)

pop_control <- Population$new(
  name = "control",
  data = vector_to_dataframe(c(0.1, 0.2, -0.1, 0.3))
)

pop_treatment <- Population$new(
  name = "treatment",
  data = vector_to_dataframe(c(0.5, 0.6, 0.7, 0.8))
)

tm <- Timer$new("mini_timer")
tm$add_timepoint(time = 1, arm = "control", enroller = 2L, dropper = 0L)
tm$add_timepoint(time = 1, arm = "treatment", enroller = 2L, dropper = 0L)
tm$add_timepoint(time = 2, arm = "control", enroller = 2L, dropper = 0L)
tm$add_timepoint(time = 2, arm = "treatment", enroller = 2L, dropper = 0L)

cond_first <- trigger_by_calendar(
  cal_time = 1,
  analysis = function(df, current_time) {
    data.frame(time = current_time, n = nrow(df))
  },
  name = "first_lock"
)

cond_final <- trigger_by_calendar(
  cal_time = 2,
  analysis = function(df, current_time) {
    data.frame(time = current_time, mean_data = mean(df$data))
  },
  name = "final_mean"
)

trial <- Trial$new(
  name = "condition_demo",
  seed = 123,
  timer = tm,
  population = list(pop_control, pop_treatment),
  conditions = list(cond_first, cond_final)
)

trial$run()

names(trial$results)
lapply(trial$results, names)
```

The result keys come from each condition's `name`. That makes it easy to tell a
week-12 interim from a final analysis in `trial$results` or
`collect_results()`.

## Using analysis_args

`analysis_args` is the clean way to inject scenario metadata or design
parameters into an analysis without depending on global variables.

```{r using-analysis-args}
scenario <- list(
  scenario_id = "base_case",
  target_n = 6L
)

scenario_cond <- Condition$new(
  where = calendar_trigger(12),
  analysis = function(df, current_time, scenario) {
    data.frame(
      scenario_id = scenario$scenario_id,
      target_n = scenario$target_n,
      observed_n = sum(!is.na(df$enroll_time)),
      mean_response = mean(df$response, na.rm = TRUE),
      fired_at = current_time
    )
  },
  analysis_args = list(scenario = scenario),
  name = "scenario_summary"
)

scenario_cond$check_conditions(snapshot_week12, current_time = 12)
```

This pattern scales well when you build conditions programmatically. The
analysis stays explicit about its inputs, and the condition carries the extra
arguments it needs.

## Next steps

To place `Condition` in the wider rxsim workflow, continue with:

- `class-trial`
- `api-styles`
- `interims`
