---
title: "Trial: Orchestrating the Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trial: Orchestrating the Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rxsim)
```

This vignette is a deep-dive reference for the `Trial` R6 class. `Trial`
is the orchestrator in rxsim: it coordinates `Population` objects, a
`Timer`, and one or more `Condition` objects to produce a complete simulation
run. One call to `trial$run()` is one simulated replicate.

## What is a Trial?

A `Trial` is the object that moves the simulation clock forward and freezes
state at each look. It owns five pieces of state that matter most during a
run:

- `timer`: the enrollment and dropout schedule
- `population`: the arm-level subject pools
- `conditions`: the trigger and analysis rules
- `locked_data`: the frozen subject-level snapshots
- `results`: the analysis outputs derived from those snapshots

The direct API builds these objects explicitly with `Trial$new()`. The
generator API wraps the same idea with helpers such as `replicate_trial()`.
Understanding `Trial` makes both styles easier to reason about.

## Building a Trial directly

We will build one end-to-end example and reuse it throughout the vignette.
The design is intentionally simple:

- two arms: `control` and `treatment`
- a subject pool of 40 candidates per arm
- a planned enrollment of 12 subjects per arm over four calendar looks
- one condition named `summary` that fires at times 3 and 8

Because the pool is larger than the planned enrollment, each run randomly
selects which subjects are enrolled. That makes `Trial` useful even in the
direct API: the same template can be cloned and rerun many times.

### Step 1: create the arm populations

```{r direct-populations}
set.seed(2025)

effect <- 0.6
pool_size <- 40L
planned_n <- 24L

control_pop <- Population$new(
  name = "control",
  data = data.frame(
    id = seq_len(pool_size),
    readout_time = 0,
    response = rnorm(pool_size, mean = 0, sd = 1)
  )
)

treatment_pop <- Population$new(
  name = "treatment",
  data = data.frame(
    id = seq_len(pool_size),
    readout_time = 0,
    response = rnorm(pool_size, mean = effect, sd = 1)
  )
)
```

### Step 2: create the timer

```{r direct-timer}
trial_timer <- Timer$new(name = "class_trial_timer")

schedule <- data.frame(
  time = c(1, 1, 3, 3, 6, 6, 8, 8),
  arm = c(
    "control", "treatment",
    "control", "treatment",
    "control", "treatment",
    "control", "treatment"
  ),
  enroller = c(4L, 4L, 3L, 3L, 3L, 3L, 2L, 2L),
  dropper = c(0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L)
)

add_timepoints(trial_timer, schedule)
schedule
```

### Step 3: create the condition

The analysis receives the locked snapshot for the current look. Here we keep
only subjects whose measurement is available by `current_time` and who have
not dropped before that look.

```{r direct-condition}
summary_condition <- Condition$new(
  where = value_trigger("time", "%in%", c(3, 8)),
  analysis = function(df, current_time) {
    observed <- subset(
      df,
      measurement_time <= current_time &
        (is.na(drop_time) | drop_time > current_time)
    )

    tt <- t.test(response ~ arm, data = observed)

    data.frame(
      look = current_time,
      n_control = sum(observed$arm == "control"),
      n_treatment = sum(observed$arm == "treatment"),
      effect = mean(observed$response[observed$arm == "treatment"]) -
        mean(observed$response[observed$arm == "control"]),
      p_value = unname(tt$p.value)
    )
  },
  name = "summary",
  max_triggers = Inf
)
```

### Step 4: create and run the trial

```{r direct-trial}
trial <- Trial$new(
  name = "class_trial",
  seed = 2025,
  timer = trial_timer,
  population = list(control_pop, treatment_pop),
  conditions = list(summary_condition)
)

trial$run()
```

### Step 5: inspect the outputs

```{r direct-inspect}
names(trial$locked_data)
names(trial$results)
collect_results(trial)
```

The `summary` analysis fired at calendar times 3 and 8, so both
`trial$locked_data` and `trial$results` contain `time_3` and `time_8` keys in
this example.

To reuse the same design later, wrap the construction in a helper.

```{r trial-template-helper}
make_trial_template <- function(effect = 0.6, seed = 2025) {
  if (!is.null(seed)) {
    set.seed(seed)
  }

  control_pop <- Population$new(
    name = "control",
    data = data.frame(
      id = seq_len(pool_size),
      readout_time = 0,
      response = rnorm(pool_size, mean = 0, sd = 1)
    )
  )

  treatment_pop <- Population$new(
    name = "treatment",
    data = data.frame(
      id = seq_len(pool_size),
      readout_time = 0,
      response = rnorm(pool_size, mean = effect, sd = 1)
    )
  )

  trial_timer <- Timer$new(name = "class_trial_timer")
  add_timepoints(trial_timer, schedule)

  summary_condition <- Condition$new(
    where = value_trigger("time", "%in%", c(3, 8)),
    analysis = function(df, current_time) {
      observed <- subset(
        df,
        measurement_time <= current_time &
          (is.na(drop_time) | drop_time > current_time)
      )

      tt <- t.test(response ~ arm, data = observed)

      data.frame(
        look = current_time,
        n_control = sum(observed$arm == "control"),
        n_treatment = sum(observed$arm == "treatment"),
        effect = mean(observed$response[observed$arm == "treatment"]) -
          mean(observed$response[observed$arm == "control"]),
        p_value = unname(tt$p.value)
      )
    },
    name = "summary",
    max_triggers = Inf
  )

  Trial$new(
    name = "class_trial_template",
    seed = seed,
    timer = trial_timer,
    population = list(control_pop, treatment_pop),
    conditions = list(summary_condition)
  )
}
```

## The `run()` loop explained

`Trial$run()` is a look-by-look loop over the unique times stored in the
`Timer`. For each timepoint `t`, think in four steps.

1. **Update each arm**: for every population, `Trial` reads the timer counts
   for that arm at time `t` and calls `set_enrolled()` and `set_dropped()`.
2. **Build a snapshot**: it combines the arm-level data, adds
   `enroll_time`, `drop_time`, `measurement_time`, and `time`, then binds the
   arms into one data frame.
3. **Lock the snapshot**: the combined data frame is stored under a time key
   such as `"time_3"`.
4. **Evaluate conditions**: each `Condition` receives the snapshot and the
   current time. Any non-empty analysis output is merged into `results`.

You can see the timepoints from the timer directly.

```{r run-loop-times}
trial$timer$get_unique_times()
```

And you can inspect one locked snapshot to see the extra bookkeeping columns.

```{r run-loop-snapshot}
head(
  trial$locked_data$time_3[
    , c(
      "id",
      "arm",
      "response",
      "enroll_time",
      "drop_time",
      "measurement_time",
      "time"
    )
  ]
)
```

That snapshot is exactly what the `summary` condition receives at time 3.
The corresponding derived analysis lives under the matching key in
`trial$results`.

```{r run-loop-result}
trial$results$time_3$summary
```

## `locked_data` vs `results`

`locked_data` and `results` answer different questions.

- Use `locked_data` when you need the full subject-level state at a look.
  This is the debugging and audit layer.
- Use `results` when you need the derived quantities from a condition, such
  as a p-value, a posterior probability, or a decision flag.

In this example the keys line up because the only stored analyses happen at
calendar times 3 and 8.

```{r locked-vs-results}
lapply(trial$locked_data, nrow)
collect_results(trial)
```

A good rule of thumb is: reach for `locked_data` when you want to understand
who was in the cohort, and reach for `results` when you want to summarize what
the analysis concluded.

## Seeds and reproducibility

`Trial$new(seed = ...)` sets the random seed during trial construction. In the
direct API, that seed governs both the random subject pool we generate and the
random enrollment and dropout choices made during `run()`.

Two trials created from the same template code with the same seed produce the
same result. Changing the seed changes the realized subject pool and the order
in which subjects are enrolled or dropped.

```{r reproducibility}
trial_a <- make_trial_template(effect = 0.6, seed = 2025)
trial_b <- make_trial_template(effect = 0.6, seed = 2025)
trial_c <- make_trial_template(effect = 0.6, seed = 2026)

trial_a$run()
trial_b$run()
trial_c$run()

identical(collect_results(trial_a), collect_results(trial_b))
identical(collect_results(trial_a), collect_results(trial_c))
```

For replication workflows, the important detail is not just the seed. It is
also the object graph. `clone_trial()` deep-copies the `Timer`,
`Population`, and `Condition` objects, so running one clone does not mutate the
state of another clone.

## Cloning for replication

`clone_trial(trial, n)` is the direct API replication tool. It starts from one
trial template and returns `n` independent copies.

A quick check makes the deep-copy behavior visible.

```{r clone-independence}
clone_pair <- clone_trial(make_trial_template(effect = 0.6, seed = 2025), n = 2)

clone_pair[[1]]$population[[1]]$set_enrolled(1, time = 0)

c(
  clone_1 = sum(!is.na(clone_pair[[1]]$population[[1]]$enrolled)),
  clone_2 = sum(!is.na(clone_pair[[2]]$population[[1]]$enrolled))
)
```

Now run a real batch of replicates.

```{r clone-run-50, results = FALSE}
set.seed(6001)

clones_50 <- clone_trial(make_trial_template(effect = 0.6, seed = 2025), n = 50)
run_trials(clones_50)
```

```{r clone-results-50}
clone_results_50 <- subset(collect_results(clones_50), timepoint == 8)
head(clone_results_50)
```

A larger batch gives an empirical power estimate for this fixed design.

```{r clone-run-200, results = FALSE}
set.seed(6002)

clones_200 <- clone_trial(
  make_trial_template(effect = 0.6, seed = 2025),
  n = 200
)
run_trials(clones_200)
```

```{r clone-power-200}
clone_results_200 <- subset(collect_results(clones_200), timepoint == 8)
power_200 <- mean(clone_results_200$p_value < 0.05)
power_200
```

This is the core operating-characteristic workflow with the direct API:

1. build one template trial
2. clone it many times
3. run the clones
4. flatten the results with `collect_results()`
5. summarize the statistic of interest

## `collect_results()`

`collect_results(x)` accepts a single `Trial` object or a list of `Trial`
objects and returns a tidy data frame with `replicate`, `timepoint`, and
`analysis` columns prepended to each row of analysis output.

```{r collect-results}
collect_results(trial)
```

For replicate work, `collect_results()` is usually the last step before you
compute power, type I error, expected duration, or any other operating
characteristic.

## The generator API shortcut

The generator API packages the same ideas into a higher-level entry point.
Instead of building `Population`, `Timer`, and `Condition` objects yourself,
you provide generator functions and let `replicate_trial()` assemble the
trials for you.

```{r generator-api, results = FALSE}
generator_trials <- replicate_trial(
  trial_name = "class_trial_shortcut",
  sample_size = planned_n,
  arms = c("control", "treatment"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 0.7),
  dropout = function(n) rexp(n, rate = 0.15),
  analysis_generators = list(
    summary = list(
      trigger = enroll_trigger(1.0, planned_n),
      analysis = function(df, current_time) {
        observed <- subset(
          df,
          measurement_time <= current_time &
            (is.na(drop_time) | drop_time > current_time)
        )

        tt <- t.test(response ~ arm, data = observed)

        data.frame(
          look = current_time,
          n_control = sum(observed$arm == "control"),
          n_treatment = sum(observed$arm == "treatment"),
          effect = mean(observed$response[observed$arm == "treatment"]) -
            mean(observed$response[observed$arm == "control"]),
          p_value = unname(tt$p.value)
        )
      }
    )
  ),
  population_generators = list(
    control = function(n) {
      data.frame(
        id = seq_len(n),
        readout_time = 0,
        response = rnorm(n, mean = 0, sd = 1)
      )
    },
    treatment = function(n) {
      data.frame(
        id = seq_len(n),
        readout_time = 0,
        response = rnorm(n, mean = 0.6, sd = 1)
      )
    }
  ),
  n = 3
)

run_trials(generator_trials)
```

```{r generator-results}
collect_results(generator_trials)
```

Use `Trial$new()` when you want full control over the object graph. Use
`replicate_trial()` when you want the same trial concept with less setup. The
[Two API Styles](api-styles.html) vignette compares the two approaches side by
side.

## Power curve with 300 clones

To end, we can sweep the treatment effect and estimate power for each value
with 300 clones of the same direct-API template.

```{r power-curve-compute, cache = TRUE, results = FALSE}
estimate_power <- function(effect, n = 300) {
  set.seed(8000 + round(effect * 100))

  reps <- clone_trial(make_trial_template(effect = effect, seed = 2025), n = n)
  run_trials(reps)

  res <- subset(collect_results(reps), timepoint == 8)
  mean(res$p_value < 0.05)
}

effect_grid <- c(0.0, 0.4, 0.8, 1.2)

power_curve <- data.frame(
  effect = effect_grid,
  power = vapply(effect_grid, estimate_power, numeric(1), n = 300)
)
```

```{r power-curve-table}
power_curve
```

```{r power-curve-plot}
plot(
  power_curve$effect,
  power_curve$power,
  type = "b",
  pch = 19,
  ylim = c(0, 1),
  xlab = "True treatment effect",
  ylab = "Pr(p_value < 0.05)",
  main = "Power curve from 300 direct-API clones"
)
abline(h = 0.8, lty = 2, col = "grey50")
abline(h = 0.05, lty = 3, col = "grey50")
```

## Next steps

From here, the most useful follow-on references are:

- [Two API Styles](api-styles.html): direct API versus generator API
- [Conditions and Triggers](conditions.html): designing multi-look analyses
- [Examples](example-1.html): complete trial designs built on the same core
  `Trial` workflow
