---
title: "Core Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Register a knitr engine that converts ```{mermaid} blocks into
# <div class="mermaid"> elements for the Mermaid JS library to render.
knitr::knit_engines$set(mermaid = function(options) {
  code <- paste(options$code, collapse = "\n")
  knitr::engine_output(
    options, code,
    out = paste0('<div class="mermaid">\n', code, '\n</div>\n')
  )
})
```

```{r setup, include = FALSE}
library(rxsim)
```

## Overview

rxsim organises a clinical trial simulation around four collaborating
objects. A `Population` owns the subject-level data and tracks each
subject's enrollment and dropout times. A `Timer` drives the trial clock:
it stores discrete timepoints per arm and defines when the simulation
clock advances. A `Condition` pairs a filter expression with an optional
analysis function and manages its own trigger state - it fires when the
snapshot data meets a criterion. A `Trial` orchestrates the simulation
by iterating over timepoints, updating populations, snapshotting the
enrolled cohort, and collecting results.

```{mermaid}
graph LR
  P1(Control) --> TR(Trial)
  P2(Treatment) --> TR
  TI(Timer) --> TR
  CO(Condition) --> TR
  TR --> LD(Locked Data)
  TR --> RS(Results)
```

The `run()` loop that `Trial` executes at each timepoint follows this
sequence:

```{mermaid}
graph TD
  A([Next timepoint]) --> B[Enroll and drop subjects]
  B --> C[Snapshot enrolled subjects]
  C --> D{Condition triggered?}
  D -- Yes --> E[Run analysis]
  D -- No --> F[Advance clock]
  E --> F
  F --> G{More timepoints?}
  G -- Yes --> A
  G -- No --> H([Done])
```

In most workflows you will never construct these objects by hand.
Instead you use the high-level entry point `replicate_trial()` +
`run_trials()`, which build and execute `n` independent `Trial` objects
from your generator functions. Understanding the three classes directly
is useful when you want to:

- inspect the locked snapshot mid-simulation for debugging
- write custom multi-timepoint designs that `gen_plan()` cannot express
- use `Trial$new()` directly for a one-off single-run simulation
  (as in [Example 8](example-8.html))

The sections below give a short summary of each building block. For the
full reference, see the dedicated vignettes linked in the table.

## Building blocks at a glance

| Class | Role | Deep dive |
|-------|------|-----------|
| `Population` | Holds subject-level endpoint data and enrollment/dropout state for one arm | [Population](class-population.html) |
| `Timer` | Stores the trial clock: when subjects enroll or drop in each arm | [Timer](class-timer.html) |
| `Condition` | Pairs a trigger expression with an analysis function; manages trigger state | [Condition](class-condition.html) |
| `Trial` | Orchestrates the simulation loop; stores snapshots and results | [Trial](class-trial.html) |

## How the pieces fit together

A typical two-arm simulation follows this pattern:

```r
# 1. Populations: one per arm
pop_pbo <- Population$new("pbo", pbo_data)
pop_trt <- Population$new("trt", trt_data)

# 2. Timer: register enrollment/dropout events
tmr <- Timer$new("my_timer")
add_timepoints(tmr, gen_plan(n, arms, alloc, enroll_fn, dropout_fn))

# 3. Condition: fire at full enrollment, run a t-test
cond <- Condition$new(
  where    = enroll_trigger(1.0, n),
  analysis = function(df, t) { ... },
  name     = "final"
)

# 4. Trial: assemble and run
trial <- Trial$new("trial", timer=tmr,
                   population=list(pop_pbo, pop_trt),
                   conditions=list(cond))
trial$run()
```

For a step-by-step walkthrough see [Getting Started](getting-started.html).
For the generator shortcut (`replicate_trial`) see
[Two API Styles](api-styles.html).

## Next steps

- [Getting Started](getting-started.html) - complete end-to-end walkthrough
- [Population](class-population.html) - subject data format and enrollment
  tracking
- [Timer](class-timer.html) - building the trial clock
- [Condition](class-condition.html) - triggers and analysis functions
- [Trial](class-trial.html) - the simulation loop, results structure, and
  replication
- [Two API Styles](api-styles.html) - direct vs. generator API

