---
title: "Core Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(rxsim)
```

## Overview

rxsim organises a clinical trial simulation around four collaborating
objects. A `Population` owns the subject-level data and tracks each
subject's enrollment and dropout times. A `Timer` drives the trial clock:
it stores discrete timepoints per arm and defines when the simulation
clock advances. A `Condition` pairs a filter expression with an optional
analysis function and manages its own trigger state - it fires when the
snapshot data meets a criterion. A `Trial` orchestrates the simulation
by iterating over timepoints, updating populations, snapshotting the
enrolled cohort, and collecting results.

```{=html}
<pre class="mermaid">
graph LR
  P1(Control) --> TR(Trial)
  P2(Treatment) --> TR
  TI(Timer) --> TR
  CO(Condition) --> TR
  TR --> LD(Locked Data)
  TR --> RS(Results)
</pre>
```

In most workflows you will never construct these objects by hand.
Instead you use the high-level entry point `replicate_trial()` +
`run_trials()`, which build and execute `n` independent `Trial` objects
from your generator functions. Understanding the four classes directly
is useful when you want to:

- inspect the locked snapshot mid-simulation for debugging
- write custom multi-timepoint designs that `stochastic_schedule()` cannot express
- use `Trial$new()` directly for a one-off single-run simulation
  (as in [Example 5](https://boehringer-ingelheim.github.io/rxsim-gallery/examples/example-5.html))

The sections below give a short summary of each building block. For the
full reference, see the dedicated vignettes linked in the table.

## Building blocks at a glance

| Class | Role | Deep dive |
|-------|------|-----------|
| `Population` | Holds subject-level endpoint data and enrollment/dropout state for one arm | [Population](class-population.html) |
| `Timer` | Stores the trial clock: when subjects enroll or drop in each arm | [Enrollment and Dropout](enrollment.html) |
| `Condition` | Pairs a trigger expression with an analysis function; manages trigger state | [Conditions and Triggers](conditions.html) |
| `Trial` | Orchestrates the simulation loop; stores snapshots and results | [Trial reference](../reference/Trial.html) |

## How the pieces fit together

A typical two-arm simulation follows this pattern:

```{r assembly}
set.seed(7)
n <- 20
arms <- c("pbo", "trt")

# 1. Timer: draw a stochastic enrollment plan and register it
tmr <- Timer$new("my_timer")
plan <- stochastic_schedule(
  sample_size = n, arms = arms, allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 1)
)
tmr$add_schedule(plan)

# 2. Populations: one per arm, sized from the plan
n_pbo <- sum(plan$enroll[plan$arm == "pbo"])
n_trt <- sum(plan$enroll[plan$arm == "trt"])
pop_pbo <- Population$new("pbo", data.frame(id = seq_len(n_pbo), y = rnorm(n_pbo, 0.0), readout_time = 1))
pop_trt <- Population$new("trt", data.frame(id = seq_len(n_trt), y = rnorm(n_trt, 0.5), readout_time = 1))

# 3. Condition: fire at full enrollment, run a t-test
cond <- Condition$new(
  where    = enroll_trigger(1.0, n),
  analysis = function(df, current_time) {
    data.frame(p_value = t.test(y ~ arm, data = df)$p.value)
  },
  name     = "final"
)

# 4. Trial: assemble and run
trial <- Trial$new(
  name = "trial", timer = tmr,
  population = list(pop_pbo, pop_trt),
  conditions = list(cond)
)
trial$run()

trial$results
```

For a step-by-step walkthrough see [Enrollment and Dropout](enrollment.html).
For the generator shortcut (`replicate_trial`) see
[Two API Styles](api-styles.html).

## Next steps

- [Enrollment and Dropout](enrollment.html) - stochastic and deterministic enrollment
- [Population](class-population.html) - subject data format and enrollment
  tracking
- [Conditions and Triggers](conditions.html) - triggers and analysis functions
- [Trial reference](../reference/Trial.html) - class documentation for run loop
  and outputs
- [Two API Styles](api-styles.html) - direct vs. generator API
