---
title: "Core Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Core Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Register a knitr engine that converts ```{mermaid} blocks into
# <div class="mermaid"> elements for the Mermaid JS library to render.
knitr::knit_engines$set(mermaid = function(options) {
  code <- paste(options$code, collapse = "\n")
  knitr::engine_output(
    options, code,
    out = paste0('<div class="mermaid">\n', code, '\n</div>\n')
  )
})
```

```{r setup, include = FALSE}
library(rxsim)
```

```{=html}
<script src="https://cdn.jsdelivr.net/npm/mermaid/dist/mermaid.min.js"></script>
<script>mermaid.initialize({ startOnLoad: true, theme: "default" });</script>
```

## Overview

rxsim organises a clinical trial simulation around three collaborating
objects. A `Population` owns the subject-level data and tracks each
subject's enrollment and dropout times. A `Timer` drives the trial clock.
It stores discrete timepoints per arm and evaluates
condition-triggered analyses when the data meet a criterion. A `Trial`
orchestrates the simulation by iterating over timepoints, updating
populations, snapshotting the enrolled cohort, and collecting results.

```{mermaid}
graph LR
  P1(Control) --> TR(Trial)
  P2(Treatment) --> TR
  TI(Timer) --> TR
  TR --> LD(Locked Data)
  TR --> RS(Results)
```

The `run()` loop that `Trial` executes at each timepoint follows this
sequence:

```{mermaid}
graph TD
  A([Next timepoint]) --> B[Enroll and drop subjects]
  B --> C[Snapshot enrolled subjects]
  C --> D{Condition triggered?}
  D -- Yes --> E[Run analysis]
  D -- No --> F[Advance clock]
  E --> F
  F --> G{More timepoints?}
  G -- Yes --> A
  G -- No --> H([Done])
```

In most workflows you will never construct these objects by hand.
Instead you use the high-level entry point `replicate_trial()` +
`run_trials()`, which build and execute `n` independent `Trial` objects
from your generator functions. Understanding the three classes directly
is useful when you want to:

- inspect the locked snapshot mid-simulation for debugging
- write custom multi-timepoint designs that `gen_plan()` cannot express
- use `Trial$new()` directly for a one-off single-run simulation
  (as in [Example 8](example-8.html))

The rest of this vignette unpacks each building block in turn.

## Population

A `Population` object represents a single arm's worth of subjects. It
stores the endpoint data alongside two tracking vectors (`enrolled`,
`dropped`) that record the calendar time each subject was enrolled or
dropped — or `NA` if neither event has occurred yet.

### Data structure

The `data` field must be a data frame with at least four columns:

| Column | Type | Meaning |
|--------|------|---------|
| `id` | integer | Unique subject identifier within the arm |
| `arm` | character | Arm label (auto-filled from `name` if absent) |
| `readout_time` | numeric | Lag from enrollment to endpoint observation |
| *(endpoint)* | numeric | At least one endpoint column (e.g., `data`, `y`, `tte`) |

The `readout_time` column is not a calendar time — it is the lag
between a subject's enrollment date and the date their endpoint is
observed. Use `0` for immediately available data (e.g., a baseline
measure) and a positive value for delayed readouts (e.g.,
`readout_time = 12` means the endpoint is read 12 weeks after
enrollment).

### vector_to_dataframe()

For the common case of a single continuous endpoint,
`vector_to_dataframe()` wraps a numeric vector into the required format
with columns `id`, `data`, and `readout_time = 0`:

```{r vector_to_df}
ep <- rnorm(6, mean = 1, sd = 0.5)
df <- vector_to_dataframe(ep)
df
```

You can also build the data frame yourself — useful when you have
covariates, time-varying endpoints, or different readout lags per
subject.

### Repeated measurements and n_readouts

When a subject contributes multiple measurements (a longitudinal or
pharmacokinetic design), each measurement appears as its own row with a
distinct `readout_time`. `Population` detects this automatically and
stores the number of rows per subject in `n_readouts`:

```{r n_readouts}
# Two timepoints per subject: baseline (0) and week 12 (12)
long_df <- data.frame(
  id           = rep(1:4, each = 2),
  readout_time = rep(c(0, 12), times = 4),
  response     = rnorm(8)
)

pop_long <- Population$new(name = "treatment", data = long_df)
pop_long$n          # 4 unique subjects
pop_long$n_readouts # 2 rows per subject
```

### Enrollment and dropout vectors

`enrolled` and `dropped` are numeric vectors of length `n`, initialised
to `NA` for every subject. `NA` means "not yet acted on": the subject
exists in the pool but has not been enrolled or dropped. The `Trial`
assigns calendar times to these vectors as the simulation clock
advances.

`set_enrolled(n, time)` marks `n` randomly chosen currently unenrolled
subjects as enrolled at `time`. `set_dropped(n, time)` similarly marks
`n` randomly chosen enrolled-and-not-yet-dropped subjects as dropped at
`time`.

```{r enrolled_dropped}
set.seed(1)
pop <- Population$new(name = "control", data = vector_to_dataframe(rnorm(8)))
pop$enrolled  # all NA to start

pop$set_enrolled(5, time = 2)
pop$enrolled  # 5 subjects enrolled at t=2, 3 still NA

pop$set_dropped(2, time = 7)
pop$dropped   # 2 of the enrolled subjects dropped at t=7
```

The `Trial`'s `run()` method calls these setters automatically based on
the `Timer`'s schedule — you rarely need to call them directly.

## Timer

A `Timer` drives the trial clock. It holds two data structures: a
`timelist` that specifies how many subjects to enroll or drop in each
arm at each time unit, and a `conditions` list that defines
analysis-triggering criteria. At each unique time in the `timelist`,
`Trial$run()` asks the `Timer` to evaluate its conditions against the
current snapshot.

### Timepoints

Each entry in the `timelist` has four fields:

| Field | Type | Meaning |
|-------|------|---------|
| `time` | numeric | Calendar time of the event |
| `arm` | character | Which arm the event applies to |
| `enroller` | integer | Number of subjects to enroll at this time |
| `dropper` | integer | Number of subjects to drop at this time |

Timepoints are per-arm because the two arms in a trial may enroll
subjects on different schedules, and enrollment events in one arm do not
affect the other.

```{r timer_basic}
t <- Timer$new(name = "my_timer")

t$add_timepoint(time = 1,  arm = "control",   enroller = 5L, dropper = 0L)
t$add_timepoint(time = 1,  arm = "treatment", enroller = 5L, dropper = 0L)
t$add_timepoint(time = 2,  arm = "control",   enroller = 3L, dropper = 1L)
t$add_timepoint(time = 2,  arm = "treatment", enroller = 4L, dropper = 0L)
t$add_timepoint(time = 10, arm = "control",   enroller = 0L, dropper = 0L)
t$add_timepoint(time = 10, arm = "treatment", enroller = 0L, dropper = 0L)

t$get_unique_times()   # c(1, 2, 10)
t$get_end_timepoint()  # 10
t$get_n_arms()         # 2
```

Rather than adding timepoints one by one, `add_timepoints()` accepts a
data frame with the four columns above — exactly what `gen_plan()` and
`gen_timepoints()` return. See
[Enrollment & Dropout Modeling](enrollment-dropout.html) for details.

### Conditions

A condition pairs a filter expression with an optional analysis
function. When the `Trial` calls `check_conditions()` at each
timepoint, each condition's expression is evaluated against the current
snapshot using `dplyr::filter()`. If the filtered result is non-empty
(the condition is satisfied), the associated analysis function is called
and its return value is stored.

Conditions are added with `add_condition()`. The `...` arguments are
`dplyr::filter`-style boolean expressions, evaluated lazily against the
snapshot columns:

```{r timer_condition}
# A toy snapshot data frame
snapshot <- data.frame(
  id          = 1:8,
  arm         = rep(c("control", "treatment"), 4),
  enroll_time = c(1, 1, 2, 2, NA, NA, NA, NA),
  data        = rnorm(8)
)

t2 <- Timer$new(name = "demo")

# Fire when at least 4 subjects are enrolled
t2$add_condition(
  sum(!is.na(enroll_time)) >= 4,
  analysis = function(df, current_time) {
    data.frame(n_enrolled = sum(!is.na(df$enroll_time)),
               fired_at   = current_time)
  },
  name = "interim"
)

res <- t2$check_conditions(locked_data = snapshot, current_time = 5)
res[["interim"]]
```

If no analysis function is provided, `check_conditions()` returns the
filtered subset instead of calling a function — convenient for
inspection or debugging.

## Trial

A `Trial` wires one or more `Population`s and a `Timer` together and
runs the simulation.

### Constructor

```r
Trial$new(
  name       = "my_trial",
  seed       = 42,           # optional; set for reproducibility
  timer      = my_timer,
  population = list(pop_a, pop_b)
)
```

The `population` argument is a list of `Population` objects, one per
arm. The arm labels in each `Population`'s `name` field must match the
`arm` identifiers in the `Timer`'s timelist.

### What run() does

Calling `trial$run()` executes the following loop:

1. Iterate over each unique time `t` in the `Timer`'s timelist, in
   ascending order.
2. For each arm, apply `set_enrolled()` and `set_dropped()` according
   to that arm's timepoint entry.
3. Build a combined snapshot by row-binding all arms' enrolled subjects.
   Four columns are appended by `Trial`:

| Column | Meaning |
|--------|---------|
| `enroll_time` | Calendar time the subject was enrolled (`NA` if not yet enrolled) |
| `drop_time` | Calendar time the subject dropped out (`NA` if still active) |
| `measurement_time` | `enroll_time + readout_time` |
| `time` | The current clock time `t` |

4. Call `timer$check_conditions(snapshot, t)` — evaluate all conditions
   and collect analysis results.
5. If any condition fired, store the snapshot in `locked_data[["time_t"]]`
   and the analysis outputs in `results[["time_t"]]`.

### locked_data and results

`locked_data` is a named list of snapshots, one per unique timepoint at
which at least one analysis fired. Each snapshot is the full
subject-level data frame at that moment — useful for auditing which
subjects were enrolled, computing post-hoc statistics, or debugging an
analysis function.

`results` has the same time-indexed structure, but each element is
itself a named list — one entry per condition that fired at that
timepoint. The value is whatever your analysis function returned.

### A minimal end-to-end example

```{r trial_example}
set.seed(7)

# Two small populations
pop_a <- Population$new("A", data = vector_to_dataframe(rnorm(10)))
pop_b <- Population$new("B", data = vector_to_dataframe(rnorm(10, mean = 0.5)))

# Timer: enroll in two waves, final readout at time 15
tm <- Timer$new("tm")
tm$add_timepoint(time = 1,  arm = "A", enroller = 5L, dropper = 0L)
tm$add_timepoint(time = 1,  arm = "B", enroller = 5L, dropper = 0L)
tm$add_timepoint(time = 3,  arm = "A", enroller = 5L, dropper = 0L)
tm$add_timepoint(time = 3,  arm = "B", enroller = 5L, dropper = 0L)
tm$add_timepoint(time = 15, arm = "A", enroller = 0L, dropper = 0L)
tm$add_timepoint(time = 15, arm = "B", enroller = 0L, dropper = 0L)

# Trigger: fire at calendar time 15
trigger_by_calendar(15, tm, analysis = function(df, current_time) {
  enrolled <- subset(df, !is.na(enroll_time))
  data.frame(
    n       = nrow(enrolled),
    mean_A  = mean(enrolled$data[enrolled$arm == "A"]),
    mean_B  = mean(enrolled$data[enrolled$arm == "B"])
  )
})

trial <- Trial$new(name = "demo", seed = 7, timer = tm,
                   population = list(pop_a, pop_b))
trial$run()

# Inspect results
prettify_results(trial$results)
```

```{r locked_data}
# How many subjects were enrolled at each timepoint?
sapply(trial$locked_data, function(snap) sum(!is.na(snap$enroll_time)))
```

## Analysis triggers in depth

Triggers are the mechanism by which rxsim knows when to analyse the
data and what to compute. Understanding how they are stored and
evaluated is key to writing correct and flexible simulation code.

### Why rlang::exprs()?

A trigger condition needs to be a *stored* expression, not an
immediately evaluated one. When you write
`sum(!is.na(enroll_time)) >= 20`, R would evaluate it at the point of
definition — before any data exists. `rlang::exprs()` wraps the
expression in a quoted form so it is only evaluated later, inside
`check_conditions()`, against the actual snapshot:

```{r exprs_demo, eval = FALSE}
# This is a stored expression, not an evaluated boolean:
rlang::exprs(sum(!is.na(enroll_time)) >= 20)
```

### The !! (bang-bang) operator

When you want to inject a value from the current R environment into a
stored expression, use `!!`. Without it, the variable name is treated
as a column name in the snapshot data frame — almost certainly not what
you want:

```{r bang_bang, eval = FALSE}
target_n <- 40

# CORRECT: !! injects the value 40 at definition time
trigger <- rlang::exprs(sum(!is.na(enroll_time)) >= !!target_n)

# WRONG: "target_n" would be looked for as a column name in the snapshot
trigger_bad <- rlang::exprs(sum(!is.na(enroll_time)) >= target_n)
```

This distinction matters when you loop over scenarios with different
sample sizes: each iteration should bake in its own `target_n` value
via `!!`.

### Columns available in a trigger expression

The trigger expression is evaluated against the snapshot data frame,
which contains all columns from the `Population`'s `data` plus the four
columns appended by `Trial`:

- `enroll_time` — calendar enrollment time (`NA` if not enrolled)
- `drop_time` — calendar dropout time (`NA` if not dropped)
- `measurement_time` — `enroll_time + readout_time`
- `time` — current clock time
- `arm` — arm label

Any user-defined endpoint column (e.g., `data`, `response`, `tte`) is
also available.

### trigger_by_calendar() and trigger_by_fraction()

These two helpers wrap the most common trigger patterns:

`trigger_by_calendar(cal_time, timer, analysis)` fires at a specific
calendar time — useful for a pre-planned final analysis or a scheduled
interim review:

```{r trig_calendar, eval = FALSE}
trigger_by_calendar(24, timer = tm, analysis = function(df, current_time) {
  data.frame(n_enrolled = sum(!is.na(df$enroll_time)))
})
```

`trigger_by_fraction(fraction, timer, sample_size, analysis)` fires
once a given fraction of the total target sample has enrolled — natural
for information-fraction-based interim rules:

```{r trig_fraction, eval = FALSE}
# Interim at 50% enrollment
trigger_by_fraction(0.5, timer = tm, sample_size = 100,
                    analysis = function(df, current_time) {
                      enrolled <- subset(df, !is.na(enroll_time))
                      data.frame(n = nrow(enrolled), time = current_time)
                    })
```

### Custom conditions

You can condition on any expression involving the snapshot columns. For
time-to-event endpoints this commonly means waiting for a target number
of events rather than a target enrollment count:

```{r custom_condition, eval = FALSE}
# Fire when 30 events have been observed (event = 1, censored = 0)
timer$add_condition(
  sum(event == 1 & !is.na(enroll_time)) >= !!n_events,
  analysis = my_tte_analysis,
  name = "event_driven_interim"
)
```

Multiple conditions can be combined with commas (they are ANDed
together, just as in `dplyr::filter()`):

```{r multi_condition, eval = FALSE}
# Fire only for the treatment arm when 15 subjects are enrolled there
timer$add_condition(
  arm == "treatment",
  sum(!is.na(enroll_time[arm == "treatment"])) >= 15,
  analysis = my_analysis,
  name = "trt_interim"
)
```

### The analysis function signature

Every analysis function receives two arguments:

- `df`: the **full snapshot** — all arms, all currently enrolled
  subjects. The condition filter determined *whether* to fire; the
  analysis function decides *what to compute* from the full data.
- `current_time`: the numeric clock time at which the condition fired.

```{r analysis_fn, eval = FALSE}
my_analysis <- function(df, current_time) {
  enrolled <- subset(df, !is.na(enroll_time))
  data.frame(
    fired_at  = current_time,
    n_ctrl    = sum(enrolled$arm == "control"),
    n_trt     = sum(enrolled$arm == "treatment"),
    mean_diff = mean(enrolled$data[enrolled$arm == "treatment"]) -
                mean(enrolled$data[enrolled$arm == "control"])
  )
}
```

### Return values

The analysis function's return value is stored verbatim in `results`.
`collect_results()` handles three cases:

- **`data.frame`** (recommended) — the standard pattern; one row per
  trigger event is the convention.
- **named `list`** — coerced to a single-row data frame.
- **`NULL`** — silently skipped by `collect_results()`.

Returning `NULL` is useful for side-effects-only analyses (e.g.,
writing a log entry) or for marking that a trigger condition was not
yet satisfied.

## Putting it together: replicate_trial() and run_trials()

Building a single `Trial` by hand is useful for exploration, but the
purpose of rxsim is operating-characteristic simulation across many
replicates. `replicate_trial()` + `run_trials()` handle this ergonomically.

### How replicate_trial() works

For each of the `n` replicates, `replicate_trial()`:

1. Calls `gen_plan()` with your `enrollment` and `dropout` functions to
   generate a fresh, stochastic enrollment/dropout schedule.
2. Builds a new `Timer`, loads the schedule via `add_timepoints()`, and
   registers all analysis triggers.
3. Calls each population generator function to draw fresh subject-level
   endpoint data.
4. Constructs a `Trial$new()` wiring the new `Timer` and `Population`
   objects together.

Each replicate therefore has independent endpoint data *and* independent
enrollment timing — both sources of variability that operating
characteristic simulations are designed to characterise.

### A complete six-step example

```{r full_example}
set.seed(42)

# Step 1 — design parameters
sample_size <- 30
arms        <- c("control", "treatment")
allocation  <- c(1, 1)
true_delta  <- 0.5

scenario <- data.frame(sample_size = sample_size, true_delta = true_delta)

# Step 2 — population generators (called fresh per replicate)
population_generators <- list(
  control   = function(n) vector_to_dataframe(rnorm(n)),
  treatment = function(n) vector_to_dataframe(rnorm(n, mean = true_delta))
)

# Step 3 — enrollment and dropout (inter-arrival functions)
enrollment <- function(n) rexp(n, rate = 2)
dropout    <- function(n) rexp(n, rate = 0.02)

# Step 4 — analysis trigger: fire when full enrollment is reached
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      data.frame(
        scenario,
        fired_at  = current_time,
        n_ctrl    = sum(enrolled$arm == "control"),
        n_trt     = sum(enrolled$arm == "treatment"),
        mean_ctrl = mean(enrolled$data[enrolled$arm == "control"]),
        mean_trt  = mean(enrolled$data[enrolled$arm == "treatment"])
      )
    }
  )
)

# Step 5 — create and run replicates
trials <- replicate_trial(
  trial_name            = "concepts_demo",
  sample_size           = sample_size,
  arms                  = arms,
  allocation            = allocation,
  enrollment            = enrollment,
  dropout               = dropout,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 10
)

run_trials(trials)

# Step 6 — collect and inspect
results <- collect_results(trials)
results
```

### collect_results() and the analysis= filter

When a trial has multiple named analyses (e.g., an interim and a
final), `collect_results()` returns rows for all of them. Use the
`analysis` argument to restrict to a specific name:

```{r collect_filter, eval = FALSE}
# Only the final analysis rows
collect_results(trials, analysis = "final")
```

Each row in the output corresponds to one replicate firing one named
analysis. The `replicate`, `timepoint`, and `analysis` columns identify
the provenance of every result row, making it straightforward to
compare interim and final results within the same replicate or
aggregate across replicates.

## Next steps

- [Getting Started](getting-started.html) — end-to-end walkthrough of
  the six-step pattern shown above, with commentary on each design
  choice.
- [Enrollment & Dropout Modeling](enrollment-dropout.html) — choose
  between stochastic (`gen_plan`) and piecewise-constant
  (`gen_timepoints`) schedules.
- [Example 1](example-1.html) through [Example 8](example-8.html) —
  progressively complex designs: correlated endpoints, time-to-event,
  multi-arm dose-finding, subgroup analyses, and Bayesian Go/No-Go
  rules.
