---
title: "Conditions and Triggers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditions and Triggers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(rxsim)
library(dplyr)
```

## What a Condition does

A `Condition` pairs two things:

1. a trigger (`where`) that decides **when** the condition fires
2. an analysis function (`analysis`) that decides **what** to compute when it
   fires

You can think of one `Condition` as one planned look. If your design has two
looks (for example, interim + final), define two `Condition` objects. More looks
means more `Condition` instances.

## Trigger primitives

rxsim provides trigger primitives that return `trigger` objects.

```{r trigger-primitives}
sample_size <- 120

trig_enroll_50  <- enroll_trigger(0.5, sample_size)     # 50% enrolled
trig_enroll_100 <- enroll_trigger(1.0, sample_size)     # full enrollment
trig_calendar   <- calendar_trigger(8)                         # single look
trig_count      <- count_trigger("enroll_time", ">=", 40)       # count-based
trig_value      <- value_trigger("analysis_flag", "==", TRUE)   # value-based

class(trig_enroll_50)
class(trig_calendar)
```

Use cases:

- `enroll_trigger()`: percentage-based interims tied to accrued sample size
- `calendar_trigger()`: fixed-time looks (e.g., month 4, 8, 12)
- `count_trigger()` / `value_trigger()`: custom state-driven logic

## Composing triggers with `&` and `|`

Triggers are composable:

- `a & b` means both conditions must hold
- `a | b` means either condition can fire the look

```{r trigger-composition}
combined_and <- enroll_trigger(0.5, sample_size) & calendar_trigger(6)
combined_or  <- calendar_trigger(12) | enroll_trigger(1.0, sample_size)

class(combined_and)
class(combined_or)
```

This is useful when an interim should happen only after both enough enrollment
and sufficient follow-up time.

## Helper constructors for common conditions

Two helper constructors cover common workflows and return `Condition` objects
directly.

```{r helper-constructors}
cond_calendar <- condition_calendar_time(
  cal_time = 8,
  analysis = function(df, current_time) {
    enrolled <- subset(df, !is.na(enroll_time))
    data.frame(time = current_time, n = nrow(enrolled))
  },
  name = "cal_8"
)

cond_fraction <- condition_enrollment_fraction(
  fraction = 0.5,
  sample_size = sample_size,
  analysis = function(df, current_time) {
    enrolled <- subset(df, !is.na(enroll_time))
    data.frame(time = current_time, frac = nrow(enrolled) / sample_size)
  },
  name = "frac_50"
)

class(cond_calendar)
class(cond_fraction)
```

## Interim + final example (two Conditions)

This pattern is the core of multi-look designs: define one interim condition and
one final condition, then pass both into `Trial$new()`.

```{r define-design}
set.seed(111)

arms <- c("pbo", "trt")
allocation <- c(1, 1)
sample_size <- 80

plan <- deterministic_schedule(
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = list(end_time = c(4, 8), rate = c(8, 2)),
  dropout = list(end_time = c(4, 8), rate = c(0, 1))
)

tmr <- Timer$new(name = "conditions_timer")
add_timepoints(tmr, plan)

n_by_arm <- c(
  pbo = sum(plan$enroll[plan$arm == "pbo"]),
  trt = sum(plan$enroll[plan$arm == "trt"])
)

pbo_data <- data.frame(
  id = seq_len(n_by_arm[["pbo"]]),
  y = rnorm(n_by_arm[["pbo"]], mean = 0),
  readout_time = 1
)

trt_data <- data.frame(
  id = seq_len(n_by_arm[["trt"]]),
  y = rnorm(n_by_arm[["trt"]], mean = 0.4),
  readout_time = 1
)

populations <- list(
  Population$new("pbo", pbo_data),
  Population$new("trt", trt_data)
)
```

```{r define-conditions}
interim <- condition_calendar_time(
  cal_time = 4,
  analysis = function(df, current_time) {
    dat <- df |> filter(!is.na(enroll_time))
    data.frame(
      look = "interim",
      time = current_time,
      n_total = nrow(dat),
      p_value = t.test(y ~ arm, data = dat)$p.value
    )
  },
  name = "interim"
)

final <- condition_enrollment_fraction(
  fraction = 1.0,
  sample_size = sample_size,
  analysis = function(df, current_time) {
    dat <- df |> filter(!is.na(enroll_time))
    data.frame(
      look = "final",
      time = current_time,
      n_total = nrow(dat),
      p_value = t.test(y ~ arm, data = dat)$p.value
    )
  },
  name = "final"
)
```

```{r run-trial, results=FALSE}
trial <- Trial$new(
  name = "conditions_demo",
  seed = 111,
  timer = tmr,
  population = populations,
  conditions = list(interim, final)
)

trial$run()
```

```{r inspect-results}
trial$results
collect_results(trial)
```

Each look writes to a different entry under `trial$results` and appears as a
separate row in `collect_results()` via the `analysis` column.

## Cooldown and max triggers

For recurrent conditions, you can cap firing frequency with:

- `cooldown`: minimum time gap between triggers
- `max_triggers`: maximum total number of triggers

These settings are useful for frequent data-driven checks where you want
guardrails against over-triggering.

## Pattern summary

1. Define one `Condition` per planned look.
2. Keep trigger logic explicit and readable.
3. Use composable triggers (`&`, `|`) for richer rules.
4. Aggregate with `collect_results()` for downstream operating characteristics.

## Next steps

- [Enrollment and Dropout](enrollment.html) - building timers for fixed or
  stochastic schedules
- [Trial reference](../reference/Trial.html) - how Trial orchestrates timer,
  populations, and conditions
- [Example 5](example-5.html) - seamless Ph2a/2b with interim and final looks
