---
title: "Endpoint Types: Continuous, Binary, and Time-to-Event"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Endpoint Types: Continuous, Binary, and Time-to-Event}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{rxsim, survival}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(rxsim)
set.seed(42)
```

In `rxsim`, a `population_generator` is a function `function(n)` that returns
one row per subject and includes `id`, `readout_time`, and one or more endpoint
columns. `arm` is optional because `rxsim` fills it in automatically.

`readout_time` is the delay between enrollment and endpoint availability. When
`readout_time = 0`, the endpoint is available as soon as the subject enrolls.
For delayed endpoints, it is helpful to think in terms of
`enroll_time + readout_time`. In the examples below, the analysis functions use
the `measurement_time` column saved in locked snapshots to decide which endpoint
values are available at a given look.

## Continuous endpoints

Continuous endpoints are often the simplest starting point. Simulate one value
per subject with `rnorm()`, encode the treatment effect as a mean shift
`delta`, and set `readout_time = 0` when the endpoint is observed immediately.

```{r continuous-example}
sample_size <- 120
n_per_arm <- sample_size / 2
arms <- c("pbo", "trt")
allocation <- c(1, 1)
delta <- 0.5

continuous_scenario <- tidyr::expand_grid(
  endpoint = "continuous",
  sample_size = sample_size,
  n_per_arm = n_per_arm,
  delta = delta
)

continuous_population <- list(
  pbo = function(n) {
    data.frame(
      id = seq_len(n),
      y = rnorm(n, mean = 0, sd = 1),
      readout_time = 0
    )
  },
  trt = function(n) {
    data.frame(
      id = seq_len(n),
      y = rnorm(n, mean = delta, sd = 1),
      readout_time = 0
    )
  }
)

continuous_analysis <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      available <- subset(
        df,
        !is.na(enroll_time) & measurement_time <= current_time
      )
      tt <- t.test(y ~ arm, data = available)
      means <- tapply(available$y, available$arm, mean)

      data.frame(
        continuous_scenario,
        analysis_time = current_time,
        n_available = nrow(available),
        mean_pbo = unname(means["pbo"]),
        mean_trt = unname(means["trt"]),
        mean_diff = unname(means["trt"] - means["pbo"]),
        p_value = unname(tt$p.value),
        stringsAsFactors = FALSE
      )
    }
  )
)

continuous_trials <- replicate_trial(
  trial_name = "continuous_endpoint_vignette",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = continuous_analysis,
  population_generators = continuous_population,
  n = 3
)
```

The analysis uses a two-sample t-test and reports both the p-value and the
observed treatment minus placebo mean difference.

```{r continuous-run, results = FALSE}
run_trials(continuous_trials)
```

```{r continuous-results}
collect_results(continuous_trials)
```

Because `readout_time = 0`, every enrolled subject is immediately analyzable,
so the availability filter and the enrollment filter select the same subjects at
final analysis.

## Binary endpoints

Binary endpoints use `rbinom()` to simulate a 0 or 1 response. Here the placebo
response rate is 0.20, the treatment response rate is 0.40, and the endpoint is
again available immediately.

```{r binary-example}
sample_size <- 160
n_per_arm <- sample_size / 2
arms <- c("pbo", "trt")
allocation <- c(1, 1)
p_pbo <- 0.20
p_trt <- 0.40

binary_scenario <- tidyr::expand_grid(
  endpoint = "binary",
  sample_size = sample_size,
  n_per_arm = n_per_arm,
  p_pbo = p_pbo,
  p_trt = p_trt
)

binary_population <- list(
  pbo = function(n) {
    data.frame(
      id = seq_len(n),
      response = rbinom(n, size = 1, prob = p_pbo),
      readout_time = 0
    )
  },
  trt = function(n) {
    data.frame(
      id = seq_len(n),
      response = rbinom(n, size = 1, prob = p_trt),
      readout_time = 0
    )
  }
)

binary_analysis <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      available <- subset(
        df,
        !is.na(enroll_time) & measurement_time <= current_time
      )
      successes <- with(available, tapply(response, arm, sum))
      totals <- with(available, tapply(response, arm, length))
      test <- prop.test(x = successes, n = totals)
      rates <- successes / totals

      data.frame(
        binary_scenario,
        analysis_time = current_time,
        n_available = nrow(available),
        rate_pbo = unname(rates["pbo"]),
        rate_trt = unname(rates["trt"]),
        rate_diff = unname(rates["trt"] - rates["pbo"]),
        p_value = unname(test$p.value),
        stringsAsFactors = FALSE
      )
    }
  )
)

binary_trials <- replicate_trial(
  trial_name = "binary_endpoint_vignette",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = function(n) rexp(n, rate = 10),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = binary_analysis,
  population_generators = binary_population,
  n = 3
)
```

You could swap `prop.test()` for Fisher's exact test if expected counts are
small. The structure is the same: compute the arm-level rates, then return the
rate difference and p-value in a one-row data frame.

```{r binary-run, results = FALSE}
run_trials(binary_trials)
```

```{r binary-results}
collect_results(binary_trials)
```

## Time-to-event endpoints

For time-to-event endpoints, generate each subject's true event time and set
`readout_time` equal to that value. At a calendar look, subjects whose
`measurement_time` is still in the future are censored: they contribute follow
up time through `current_time - enroll_time`, with `event_observed = 0`.

The example below uses exponential event times with medians of 12 months in the
placebo arm and 18 months in the treatment arm. The analysis fires at calendar
time 24.

```{r tte-example}
library(survival)

sample_size <- 120
n_per_arm <- sample_size / 2
arms <- c("pbo", "trt")
allocation <- c(1, 1)
median_pbo <- 12
median_trt <- 18
rate_pbo <- log(2) / median_pbo
rate_trt <- log(2) / median_trt

survival_scenario <- tidyr::expand_grid(
  endpoint = "time_to_event",
  sample_size = sample_size,
  n_per_arm = n_per_arm,
  median_pbo = median_pbo,
  median_trt = median_trt
)

make_tte_population <- function(rate) {
  function(n) {
    event_time <- rexp(n, rate = rate)
    data.frame(
      id = seq_len(n),
      event_time = event_time,
      event_observed = 1L,
      readout_time = event_time
    )
  }
}

tte_population <- list(
  pbo = make_tte_population(rate_pbo),
  trt = make_tte_population(rate_trt)
)

tte_analysis <- list(
  month_24 = list(
    trigger = calendar_trigger(24),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      follow_up <- pmax(0, current_time - enrolled$enroll_time)
      enrolled$event_observed <- as.integer(
        enrolled$measurement_time <= current_time
      )
      enrolled$tte <- ifelse(
        enrolled$event_observed == 1L,
        enrolled$event_time,
        follow_up
      )

      surv_obj <- Surv(enrolled$tte, enrolled$event_observed)
      logrank <- survdiff(surv_obj ~ arm, data = enrolled)
      cox_fit <- coxph(surv_obj ~ arm, data = enrolled)
      cox_summary <- summary(cox_fit)
      p_value <- 1 - pchisq(logrank$chisq, df = 1)

      data.frame(
        survival_scenario,
        analysis_time = current_time,
        n_enrolled = nrow(enrolled),
        events = sum(enrolled$event_observed),
        median_follow_up = stats::median(enrolled$tte),
        hazard_ratio = unname(cox_summary$coef[1, "exp(coef)"]),
        p_value = unname(p_value),
        stringsAsFactors = FALSE
      )
    }
  )
)

tte_trials <- replicate_trial(
  trial_name = "tte_endpoint_vignette",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = tte_analysis,
  population_generators = tte_population,
  n = 3
)
```

```{r tte-run, results = FALSE}
run_trials(tte_trials)
```

```{r tte-results}
collect_results(tte_trials)
```

In this setup, `event_time` is the subject's true time to event, while
`event_observed` is rebuilt at each look. Subjects with
`measurement_time <= current_time` contribute events; subjects still at risk at
that calendar time are censored and contribute only the follow up observed so
far.

If you prefer a Weibull model, replace `rexp()` with `rweibull()` and carry the
same idea through the analysis function.

## Multiple endpoints

A single population generator can return several endpoint columns. That lets you
analyze, for example, a continuous symptom score and a binary response flag from
the same simulated subjects.

```{r multiple-endpoints-example}
sample_size <- 100
n_per_arm <- sample_size / 2
arms <- c("pbo", "trt")
allocation <- c(1, 1)
score_delta <- 0.4
p_pbo <- 0.25
p_trt <- 0.45

multi_scenario <- tidyr::expand_grid(
  endpoint = "multiple",
  sample_size = sample_size,
  n_per_arm = n_per_arm,
  score_delta = score_delta,
  p_pbo = p_pbo,
  p_trt = p_trt
)

multi_population <- list(
  pbo = function(n) {
    data.frame(
      id = seq_len(n),
      score = rnorm(n, mean = 0, sd = 1),
      response = rbinom(n, size = 1, prob = p_pbo),
      readout_time = 0
    )
  },
  trt = function(n) {
    data.frame(
      id = seq_len(n),
      score = rnorm(n, mean = score_delta, sd = 1),
      response = rbinom(n, size = 1, prob = p_trt),
      readout_time = 0
    )
  }
)

multi_analysis <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      available <- subset(
        df,
        !is.na(enroll_time) & measurement_time <= current_time
      )
      score_test <- t.test(score ~ arm, data = available)
      responses <- with(available, tapply(response, arm, sum))
      totals <- with(available, tapply(response, arm, length))
      response_test <- prop.test(x = responses, n = totals)
      rates <- responses / totals
      means <- tapply(available$score, available$arm, mean)

      data.frame(
        multi_scenario,
        analysis_time = current_time,
        score_diff = unname(means["trt"] - means["pbo"]),
        score_p_value = unname(score_test$p.value),
        response_diff = unname(rates["trt"] - rates["pbo"]),
        response_p_value = unname(response_test$p.value),
        stringsAsFactors = FALSE
      )
    }
  )
)

multi_trials <- replicate_trial(
  trial_name = "multiple_endpoint_vignette",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = multi_analysis,
  population_generators = multi_population,
  n = 3
)
```

```{r multiple-endpoints-run, results = FALSE}
run_trials(multi_trials)
```

```{r multiple-endpoints-results}
collect_results(multi_trials)
```

This example analyzes each endpoint independently and returns both results in
one record. If you need multiplicity control, simple corrections such as Holm
or Bonferroni fit naturally here. See [Example 3](example-3.html) for a fuller
two-endpoint workflow.

## `readout_time` mechanics

Suppose a subject enrolls at time 3 and has `readout_time = 2`. Their endpoint
becomes available at time 5. The snapshot contains both `enroll_time` and the
derived `measurement_time = enroll_time + readout_time`, so you can filter to
available data with `measurement_time <= current_time`.

```{r readout-mechanics-example}
readout_population <- list(
  single = function(n) {
    data.frame(
      id = seq_len(n),
      delayed_endpoint = c(10, 11, 12, 13),
      readout_time = c(2, 2, 2, 2)
    )
  }
)

readout_analysis <- list(
  month_5 = list(
    trigger = calendar_trigger(5),
    analysis = function(df, current_time) {
      available <- subset(df, measurement_time <= current_time)
      subject_three_time <- df$measurement_time[df$id == 3][1]

      data.frame(
        analysis_time = current_time,
        n_enrolled = nrow(df),
        n_available = nrow(available),
        available_ids = paste(available$id, collapse = ", "),
        subject_3_available_at = subject_three_time,
        stringsAsFactors = FALSE
      )
    }
  )
)

readout_trials <- replicate_trial(
  trial_name = "readout_mechanics_vignette",
  sample_size = 4,
  arms = "single",
  allocation = 1,
  enrollment = function(n) rep(1, n),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = readout_analysis,
  population_generators = readout_population,
  n = 1
)
```

```{r readout-mechanics-run, results = FALSE}
run_trials(readout_trials)
```

```{r readout-mechanics-results}
collect_results(readout_trials)
```

At `current_time = 5`, all four subjects are enrolled, but only the three with
`measurement_time <= 5` are available. Subject 3 enrolled at time 3 and has
`readout_time = 2`, so their endpoint appears exactly at time 5. This is the
same pattern you would use for delayed endpoints such as progression-free
survival measured 6 months after enrollment.

## Illustrative power comparison

The table below compares rough power across endpoint types using the same
per-arm sample size and the same numeric `delta = 0.5`, applied on each model's
natural scale: a mean shift for the continuous endpoint, a log-odds shift for
the binary endpoint, and a log-hazard reduction for the time-to-event endpoint.

```{r power-comparison-setup}
delta_common <- 0.5
n_per_arm <- 60
sample_size <- 2 * n_per_arm
alpha <- 0.05
n_sim <- 50

power_continuous_population <- list(
  pbo = function(n) {
    data.frame(
      id = seq_len(n),
      y = rnorm(n, mean = 0, sd = 1),
      readout_time = 0
    )
  },
  trt = function(n) {
    data.frame(
      id = seq_len(n),
      y = rnorm(n, mean = delta_common, sd = 1),
      readout_time = 0
    )
  }
)

power_binary_population <- list(
  pbo = function(n) {
    data.frame(
      id = seq_len(n),
      response = rbinom(n, size = 1, prob = 0.20),
      readout_time = 0
    )
  },
  trt = function(n) {
    data.frame(
      id = seq_len(n),
      response = rbinom(
        n,
        size = 1,
        prob = plogis(qlogis(0.20) + delta_common)
      ),
      readout_time = 0
    )
  }
)

power_tte_population <- list(
  pbo = function(n) {
    event_time <- rexp(n, rate = log(2) / 12)
    data.frame(
      id = seq_len(n),
      event_time = event_time,
      event_observed = 1L,
      readout_time = event_time
    )
  },
  trt = function(n) {
    event_time <- rexp(n, rate = (log(2) / 12) * exp(-delta_common))
    data.frame(
      id = seq_len(n),
      event_time = event_time,
      event_observed = 1L,
      readout_time = event_time
    )
  }
)

power_continuous_analysis <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      available <- subset(df, measurement_time <= current_time)
      data.frame(p_value = t.test(y ~ arm, data = available)$p.value)
    }
  )
)

power_binary_analysis <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      available <- subset(df, measurement_time <= current_time)
      successes <- with(available, tapply(response, arm, sum))
      totals <- with(available, tapply(response, arm, length))
      data.frame(p_value = prop.test(successes, totals)$p.value)
    }
  )
)

power_tte_analysis <- list(
  month_24 = list(
    trigger = calendar_trigger(24),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      follow_up <- pmax(0, current_time - enrolled$enroll_time)
      enrolled$event_observed <- as.integer(
        enrolled$measurement_time <= current_time
      )
      enrolled$tte <- ifelse(
        enrolled$event_observed == 1L,
        enrolled$event_time,
        follow_up
      )
      surv_obj <- survival::Surv(enrolled$tte, enrolled$event_observed)
      logrank <- survival::survdiff(surv_obj ~ arm, data = enrolled)
      data.frame(p_value = 1 - pchisq(logrank$chisq, df = 1))
    }
  )
)

power_continuous_trials <- replicate_trial(
  trial_name = "power_continuous",
  sample_size = sample_size,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = power_continuous_analysis,
  population_generators = power_continuous_population,
  n = n_sim
)

power_binary_trials <- replicate_trial(
  trial_name = "power_binary",
  sample_size = sample_size,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = power_binary_analysis,
  population_generators = power_binary_population,
  n = n_sim
)

power_tte_trials <- replicate_trial(
  trial_name = "power_tte",
  sample_size = sample_size,
  arms = c("pbo", "trt"),
  allocation = c(1, 1),
  enrollment = function(n) rexp(n, rate = 8),
  dropout = function(n) rexp(n, rate = 0.001),
  analysis_generators = power_tte_analysis,
  population_generators = power_tte_population,
  n = n_sim
)
```

```{r power-comparison-run, results = FALSE}
run_trials(power_continuous_trials)
run_trials(power_binary_trials)
run_trials(power_tte_trials)
```

```{r power-comparison-results}
power_results <- rbind(
  data.frame(
    endpoint = "continuous",
    delta = delta_common,
    n_per_arm = n_per_arm,
    estimated_power = mean(
      collect_results(power_continuous_trials)$p_value < alpha
    )
  ),
  data.frame(
    endpoint = "binary",
    delta = delta_common,
    n_per_arm = n_per_arm,
    estimated_power = mean(
      collect_results(power_binary_trials)$p_value < alpha
    )
  ),
  data.frame(
    endpoint = "time_to_event",
    delta = delta_common,
    n_per_arm = n_per_arm,
    estimated_power = mean(
      collect_results(power_tte_trials)$p_value < alpha
    )
  )
)

power_results
```

Treat this as an illustrative comparison, not a final design calculation. Power
depends not only on `n`, but also on how the same numeric `delta` maps onto the
endpoint model and analysis.

## Next steps

- Class-condition: read the `Condition` documentation to build richer trigger
  logic and reusable analyses.
- Interims: add multiple `enroll_trigger()` or `calendar_trigger()` looks to
  simulate interim decisions alongside the final analysis.
- Examples: compare this vignette with [Example 3](example-3.html) and
  [Example 4](example-4.html) for more elaborate endpoint setups.
