---
title: "Enrollment & Dropout Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enrollment & Dropout Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

## Why enrollment modeling matters

Enrollment pace determines when each subject enters the trial and therefore
when the trial clock can reach the timepoints that trigger analyses. In most
real programs, the rate at which sites activate and patients are screened is
uncertain — enrollment is itself a random process. When that randomness is
ignored (for example, by assuming all subjects enroll on a fixed schedule),
the simulated distribution of study duration is artificially narrow, and
operating characteristics such as power or expected decision error rates can
be materially overestimated. Capturing stochastic enrollment propagates
uncertainty in study duration through to every downstream quantity the
simulation is designed to evaluate.

## The two approaches

rxsim offers two complementary strategies, each suited to a different
modeling philosophy:

- **`gen_plan()`** — stochastic. You supply a random function for
  inter-enrollment times. Every call draws a new realization, so each
  replicate gets its own unique enrollment timeline. Use this when
  trial-duration variability is substantively important and you want
  operating characteristics to reflect it.

- **`gen_timepoints()`** — piecewise-constant. You specify how many subjects
  enroll per unit time in each time period. The schedule is fixed and
  deterministic. Every replicate uses exactly the same enrollment pattern.
  Use this when you have a well-characterized enrollment plan and want to
  isolate variability in endpoints and analyses from variability in timing.

Both functions return a data frame that is passed directly to
`add_timepoints()` to populate a `Timer`.

## Stochastic enrollment with gen_plan()

### Concept

In the stochastic approach, enrollment is modeled as a sequence of waiting
times between successive subject arrivals. You provide a function that, given
`n`, returns a vector of `n` independent inter-arrival durations. rxsim then
takes the cumulative sum of those durations to produce the calendar time at
which each subject enrolls. This is the standard structure of a Poisson
process — a simple and widely used model for patient arrival — although any
non-negative distribution can be substituted.

For example, `function(n) rexp(n, rate = 1)` draws waiting times from an
exponential distribution with rate 1, which means the average gap between
consecutive enrollments is 1 time unit. Over the course of the trial these
gaps accumulate to produce enrollment calendar times that, in expectation,
grow linearly but vary around that line across replicates.

### Example

Below we generate a plan for 12 subjects across two arms in balanced
allocation. The dropout function uses a much lower rate (0.05), reflecting a
trial where most subjects are expected to complete.

```{r gen_plan_example}
set.seed(101)

plan_a <- gen_plan(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = function(n) rexp(n, rate = 1),
  dropout     = function(n) rexp(n, rate = 0.05)
)

plan_a
```

### Output columns

Each row in the returned data frame represents a single event — either an
enrollment or a dropout:

| Column | Meaning |
|--------|---------|
| `arm` | Arm label (`"pbo"` or `"trt"`) |
| `enroll_time` | Calendar time of enrollment (cumulative sum of inter-arrival gaps) |
| `drop_time` | Calendar time of dropout; `NA` if the subject completes |
| `enroller` | `1` if this row is an enrollment event, `0` otherwise |
| `dropper` | `1` if this row is a dropout event, `0` otherwise |

Because the times are cumulative sums of exponential random variables, they
are continuous-valued and always positive.

### Every call produces a different schedule

Because the inter-arrival times are drawn freshly each time, repeating the
call yields a different timeline. This is the core mechanism by which
`replicate_trial()` produces independent replicates — it calls `gen_plan()`
once per replicate.

```{r gen_plan_twice}
set.seed(202)
plan_b <- gen_plan(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = function(n) rexp(n, rate = 1),
  dropout     = function(n) rexp(n, rate = 0.05)
)

# Enrollment times differ across the two calls
head(plan_a, 4)
head(plan_b, 4)
```

### Passing the plan to a Timer

Once you have the data frame, pass it to `add_timepoints()` to register all
events on a `Timer` object. The timer can then be attached to a `Trial`.

```{r gen_plan_timer}
set.seed(404)
plan <- gen_plan(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = function(n) rexp(n, rate = 1),
  dropout     = function(n) rexp(n, rate = 0.05)
)

tmr <- Timer$new(name = "stochastic_timer")
add_timepoints(tmr, plan)

tmr$get_end_timepoint()   # last event time
tmr$get_n_arms()          # number of unique arms
```

## Piecewise-constant enrollment with gen_timepoints()

### Concept

In many programs, enrollment is planned in phases: sites start slowly, ramp
up as more centers open, and may plateau near the end. The piecewise-constant
approach lets you encode this structure directly. You supply a list with two
elements:

- `end_time`: a numeric vector of period endpoints (time boundaries).
- `rate`: a numeric vector of enrollment rates (subjects per unit time) for
  each period.

rxsim interprets these as a step function. In the period from time 0 up to
`end_time[1]`, `rate[1]` subjects are enrolled per unit time; from
`end_time[1]` to `end_time[2]` the rate switches to `rate[2]`; and so on.
The schedule is deterministic — every call with the same inputs returns the
same data frame. This makes it straightforward to verify the exact enrollment
profile being fed into the simulation.

### Example

Here enrollment ramps up across three four-time-unit periods: 3, 6, and 9
subjects per unit time. Dropout rate is low and constant.

```{r gen_timepoints_example}
enrollment_schedule <- list(
  end_time = c(4, 8, 12),
  rate     = c(3, 6, 9)
)

dropout_schedule <- list(
  end_time = c(4, 8, 12),
  rate     = c(0, 1, 1)
)

tp <- gen_timepoints(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = enrollment_schedule,
  dropout     = dropout_schedule
)

tp
```

### Output columns

The output format uses the same column names as `gen_plan()`, but the
semantics differ:

| Column | Meaning |
|--------|---------|
| `arm` | Arm label |
| `enroll_time` | Integer calendar time at which one or more subjects enroll |
| `drop_time` | Integer calendar time of dropout event |
| `enroller` | Count of subjects enrolling at this time (can be > 1) |
| `dropper` | Count of subjects dropping out at this time |

Because multiple subjects may enroll in the same time unit, `enroller` can
be greater than 1. All times are positive integers.

### Every call returns the same schedule

Calling `gen_timepoints()` twice with the same arguments produces identical
output — confirming the deterministic nature of this approach.

```{r gen_timepoints_deterministic}
tp2 <- gen_timepoints(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = enrollment_schedule,
  dropout     = dropout_schedule
)

identical(tp, tp2)
```

### Passing the schedule to a Timer

As with `gen_plan()`, pass the result directly to `add_timepoints()`.

```{r gen_timepoints_timer}
tmr2 <- Timer$new(name = "deterministic_timer")
add_timepoints(tmr2, tp)

tmr2$get_end_timepoint()
tmr2$get_unique_times()
```

## Side-by-side comparison

Running both functions with the same target sample size and the same two-arm
design makes the structural differences clear.

```{r comparison}
set.seed(555)

# Stochastic
plan_stoch <- gen_plan(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = function(n) rexp(n, rate = 1),
  dropout     = function(n) rexp(n, rate = 0.05)
)

# Deterministic
plan_det <- gen_timepoints(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = list(end_time = c(4, 8, 12), rate = c(3, 6, 9)),
  dropout     = list(end_time = c(4, 8, 12), rate = c(0, 1, 1))
)

# gen_plan: continuous (fractional) enrollment times
head(plan_stoch[plan_stoch$enroller == 1, ], 5)

# gen_timepoints: integer times, counts >= 1
head(plan_det[plan_det$enroller > 0, ], 5)
```

The key trade-offs:

| | `gen_plan()` | `gen_timepoints()` |
|---|---|---|
| Enrollment times | Continuous (real-valued) | Integer |
| Replicate-to-replicate | Different each call | Identical every call |
| Study-duration distribution | Fully propagated | Fixed by design |
| Best for | Uncertainty in timing | Sensitivity to endpoints only |

For large-scale simulations where study duration is the main uncertainty,
`gen_plan()` is the natural choice. When you need a reproducible baseline
enrollment profile — for example, to match a pre-specified operational plan
or to run sensitivity analyses against a fixed schedule — `gen_timepoints()`
is preferable.

## Dropout modeling

Dropout uses exactly the same two-approach structure as enrollment,
controlled by the `dropout` argument.

### Stochastic dropout with gen_plan()

Pass a function to `dropout`. A low rate means most subjects complete the
study:

```{r dropout_plan}
# rate = 0.05 => average time-to-dropout is 20 time units
dropout_fn <- function(n) rexp(n, rate = 0.05)
```

Each subject is assigned an independent dropout time drawn from this
distribution. Dropout events appear in the returned data frame as rows where
`dropper == 1` and `enroller == 0`. In the locked data snapshot produced
during a trial run, subjects who have not yet dropped out have
`drop_time = NA`.

### Piecewise-constant dropout with gen_timepoints()

Pass a list to `dropout` using the same `end_time` / `rate` structure:

```{r dropout_timepoints}
# No dropouts in the first period; 1 per unit time thereafter
dropout_stepped <- list(
  end_time = c(4, 8, 12),
  rate     = c(0, 1,  1)
)
```

The dropout rate applies independently of enrollment: it is not conditioned
on how many subjects are currently enrolled. Both processes are generated
separately and then combined into a single schedule.

## Allocation ratios

The `allocation` argument is a numeric vector of relative weights — one entry
per arm. rxsim converts these to proportions internally:

```{r allocation_example}
# Three-arm trial: 2:1:1 allocation
plan_3arm <- gen_plan(
  sample_size = 12,
  arms        = c("pbo", "d1", "d2"),
  allocation  = c(2, 1, 1),
  enrollment  = function(n) rexp(n, rate = 1),
  dropout     = function(n) rexp(n, rate = 0.05)
)

# Per-arm enrollment counts
table(plan_3arm$arm[plan_3arm$enroller == 1])
```

For three arms with `allocation = c(2, 1, 1)`, this yields proportions 0.50,
0.25, 0.25, giving a 2:1:1 split. The target per-arm sample sizes are
computed as `round(ratio * sample_size)`, and any rounding discrepancy is
corrected automatically by adding or removing subjects from arms sampled
proportionally to their weights — so the total always equals `sample_size`
exactly.

The same `allocation` argument works identically for `gen_timepoints()`.

## Using with replicate_trial()

When you call `replicate_trial()` with functions for `enrollment` and
`dropout`, it calls `gen_plan()` internally once per replicate and constructs
an independent `Timer` for each trial. You do not need to call `gen_plan()`
directly in this workflow.

```{r replicate_trial_example}
set.seed(42)

population_generators <- list(
  pbo = function(n) data.frame(
    id = seq_len(n), response = rnorm(n, 0.0), readout_time = 0
  ),
  trt = function(n) data.frame(
    id = seq_len(n), response = rnorm(n, 0.5), readout_time = 0
  )
)

analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(sum(!is.na(enroll_time)) >= 12L),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      data.frame(
        n         = nrow(enrolled),
        mean_pbo  = mean(enrolled$response[enrolled$arm == "pbo"]),
        mean_trt  = mean(enrolled$response[enrolled$arm == "trt"])
      )
    }
  )
)

trials <- replicate_trial(
  trial_name            = "demo",
  sample_size           = 12,
  arms                  = c("pbo", "trt"),
  allocation            = c(1, 1),
  enrollment            = function(n) rexp(n, rate = 1),
  dropout               = function(n) rexp(n, rate = 0.05),
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 3
)

results <- run_trials(trials)

# Collect results across replicates
collect_results(trials)
```

Each element of `trials` holds an independent `Timer` built from a freshly
drawn enrollment plan, so study duration differs across replicates.

## Building a Timer manually with gen_timepoints()

When you need a fixed enrollment schedule and want more control than
`replicate_trial()` exposes, you can assemble the `Timer` directly. The
pattern is:

1. Call `gen_timepoints()` to get the schedule data frame.
2. Create a `Timer$new()`.
3. Register timepoints with `add_timepoints()`.
4. Add analysis triggers (e.g., with `trigger_by_calendar()`).
5. Build populations and construct a `Trial$new()`.
6. Use `clone_trial()` if you want multiple independent replicates sharing
   the same fixed schedule.

```{r manual_timer}
# Fixed enrollment schedule
enroll_sched  <- list(end_time = c(4, 8, 12), rate = c(3, 6, 9))
dropout_sched <- list(end_time = c(4, 8, 12), rate = c(0, 1, 1))

fixed_plan <- gen_timepoints(
  sample_size = 12,
  arms        = c("pbo", "trt"),
  allocation  = c(1, 1),
  enrollment  = enroll_sched,
  dropout     = dropout_sched
)

tmr3 <- Timer$new(name = "fixed_timer")
add_timepoints(tmr3, fixed_plan)

# Add an analysis at the final calendar time
final_t <- tmr3$get_end_timepoint()
trigger_by_calendar(
  cal_time = final_t,
  timer    = tmr3,
  analysis = function(df, current_time) {
    enrolled <- subset(df, !is.na(enroll_time))
    data.frame(n_enrolled = nrow(enrolled), time = current_time)
  }
)

# Build populations (sized to match the fixed plan)
pop_pbo <- gen_population(
  name        = "pbo",
  generator   = function(n) data.frame(
    id = seq_len(n), response = rnorm(n, 0), readout_time = 0
  ),
  sample_size = 6
)
pop_trt <- gen_population(
  name        = "trt",
  generator   = function(n) data.frame(
    id = seq_len(n), response = rnorm(n, 0.5), readout_time = 0
  ),
  sample_size = 6
)

trial_fixed <- Trial$new(
  name       = "fixed_schedule_trial",
  timer      = tmr3,
  seed       = 123,
  population = list(pop_pbo, pop_trt)
)

trial_fixed$run()
prettify_results(trial_fixed$results)
```

Because every replicate cloned from `trial_fixed` uses the same `fixed_plan`,
enrollment timing is constant — only subject-level endpoint variability
contributes to simulation noise.

## Next steps

- **[Getting Started](getting-started.html)** — end-to-end walkthrough of a
  complete simulation using stochastic enrollment via `replicate_trial()`.
- **[Core Concepts](concepts.html)** — detailed explanations of `Timer`,
  `Population`, `Trial`, and how they interact during a simulation run.
- **[Example 3](example-3.html)** — a worked example using `gen_timepoints()`
  with two correlated continuous endpoints and a piecewise-linear enrollment
  ramp.
