---
title: "Example 4: Two arm | Fixed design | Correlated continuous & time-to-event endpoints | t-test + survival analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)      # for mvrnorm (simulate correlated latent variables)
library(survival)  # for Surv, survdiff, coxph
```

## Scenario

We consider a **two-arm fixed design** with a **continuous endpoint (Y)** and a **time-to-event (TTE)** endpoint per subject.  
The endpoints are **correlated** via a shared Gaussian latent structure. Enrollment is piece-wise linear; analysis happens **once at the final time**.

```{r scenario}
# Total target N (across both arms)
sample_size <- 1200

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

# Piece-wise linear enrollment and dropout rates (per unit time)
# (Tune these to your development plan.)
enrollment <- list(
  end_time = c(6, 12, 18, 24),
  rate     = c(4,  8, 12, 16)
)
dropout <- list(
  end_time = c(6, 12, 18, 24),
  rate     = c(1,  2,  4,  6)
)

# Correlation between continuous endpoint driver (Z1) and frailty for TTE (Z2)
rho <- 0.50

# Continuous endpoint parameters
mu_pbo <- 0.0
mu_trt <- 0.3      # treatment mean shift
sd_y   <- 1.0

# TTE model parameters (Exponential PH with subject frailty exp(eta))
lambda0 <- 0.08    # baseline hazard (per time unit)
HR_trt  <- 0.70    # treatment hazard ratio (<1 favors treatment)

# Frailty scale (eta = sigma_frailty * Z2)
sigma_frailty <- 0.6

# Covariance of (Z1, Z2)
SigmaZ <- matrix(c(1, rho, rho, 1), nrow = 2)
```

## Time points

Generate the discrete timepoints and add them to a `Timer`.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

t <- Timer$new(name = "trial_timer_cont_tte_two_endpoints")
add_timepoints(t, timepoints)

final_time <- t$get_end_timepoint()
final_time
```

## Arms (Populations)

Per arm we simulate subject-level latent bivariate normals `(Z1, Z2)` with correlation `rho`.  
- The **continuous** endpoint is `Y = mu_arm + sd_y * Z1`.
- The **TTE** endpoint uses a proportional-hazards exponential model with subject-specific frailty:  
  `hazard_i = lambda_arm * exp(eta_i)`, where `eta_i = sigma_frailty * Z2` and `lambda_trt = lambda0 * HR_trt`.
  Event time `T_i ~ Exponential(rate = hazard_i)`.

```{r population}
set.seed(2026)

n_per_arm <- sample_size / length(arms)

# helper to simulate per arm
gen_arm <- function(n, mu_y, lambda_arm) {
  Z <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = SigmaZ)
  Z1 <- Z[, 1]
  Z2 <- Z[, 2]
  # Continuous endpoint
  y  <- mu_y + sd_y * Z1
  # Frailty -> event time
  eta <- sigma_frailty * Z2
  rate_i <- lambda_arm * exp(eta)
  tte_true <- rexp(n, rate = rate_i)  # true time-to-event from enrollment

  data.frame(
    subject_id = seq_len(n),
    y  = y,
    tte_true = tte_true
  )
}

# Control
lambda_pbo <- lambda0
control_df <- gen_arm(n_per_arm, mu_y = mu_pbo, lambda_arm = lambda_pbo)

# Treatment
lambda_trt <- lambda0 * HR_trt
trt_df     <- gen_arm(n_per_arm, mu_y = mu_trt, lambda_arm = lambda_trt)

# Build Population objects
control <- Population$new(name = "pbo", data = control_df)
treatment <- Population$new(name = "trt", data = trt_df)
```

## Trigger & Analysis

This is a **fixed design**: analyze **once at the final timepoint**.
We run a **two-sample t-test** for the continuous endpoint and a **log-rank test** plus **Cox PH** for the TTE endpoint.

Observed follow-up per subject at current time `t*` is:

\[
T_{obs} = \min\{	ext{tte\_true},\; 	ext{drop\_time} - 	ext{enroll\_time},\; t^* - 	ext{enroll\_time}\},\quad
\delta = \mathbf{1}\{	ext{tte\_true} = T_{obs}\}.
\]

```{r trigger}
t$add_condition(
  time == final_time,
  analysis = function(df, time) {
    df_e <- df |> dplyr::filter(!is.na(enroll_time))
    if (nrow(df_e) == 0) return(list(time = time, note = "no enrolled subjects"))

    # Build observed TTE with censoring
    # If dropped is NA, set censor_drop to +Inf; admin censor is (time - enroll_time)
    censor_drop  <- ifelse(is.na(df_e$drop_time), Inf, pmax(0, df_e$drop_time - df_e$enroll_time))
    censor_admin <- pmax(0, time - df_e$enroll_time)

    tte_true <- df_e$tte_true
    t_obs    <- pmin(tte_true, censor_drop, censor_admin)
    status   <- as.integer(tte_true <= pmin(censor_drop, censor_admin))

    # Continuous endpoint t-test
    p_t <- tryCatch(
      t.test(y ~ arm, data = df_e)$p.value,
      error = function(e) NA_real_
    )

    # Survival analysis: log-rank and Cox
    surv_p <- NA_real_
    hr <- hr_lo <- hr_hi <- NA_real_
    events <- sum(status)

    # Only run if we have variation and at least one event
    if (length(unique(df_e$arm)) == 2 && events > 0 && all(t_obs >= 0)) {
      S <- Surv(time = t_obs, event = status)

      # Log-rank
      surv_p <- tryCatch({
        sd <- survdiff(S ~ arm, data = df_e)
        1 - pchisq(sd$chisq, df = length(sd$n) - 1)
      }, error = function(e) NA_real_)

      # Cox PH
      cox <- tryCatch(coxph(S ~ arm, data = df_e), error = function(e) NULL)
      if (!is.null(cox)) {
        s <- summary(cox)
        # Assume armtrt vs armpbo; extract exp(coef) and CI
        hr    <- unname(s$coef[1, "exp(coef)"])
        hr_lo <- unname(s$conf.int[1, "lower .95"])
        hr_hi <- unname(s$conf.int[1, "upper .95"])
      }
    }

    list(
      n_total    = nrow(df_e),
      n_pbo      = sum(df_e$arm == "pbo"),
      n_trt      = sum(df_e$arm == "trt"),
      # continuous
      p_ttest_y  = p_t,
      # survival
      logrank_p  = surv_p,
      hr_cox     = hr,
      hr_ci_lo   = hr_lo,
      hr_ci_hi   = hr_hi,
      events     = events
    )
  },
  name = "final"
)
```

## Trial

Create and run the trial.

```{r trial}
trial <- Trial$new(
  name       = "two_endpoints_continuous_tte_fixed",
  seed       = 123,
  timer      = t,
  population = list(control, treatment)
)
```

## Simulate

```{r sim}
trial$run()
```

## Results

Analysis results at the final look:

```{r results}
trial$results
```

Flatten to a tidy frame if desired:

```{r results_pretty}
prettify_results(trial$results)
```

## Locked data (snapshot)

Inspect the snapshot used for analysis:

```{r locked}
lapply(trial$locked_data, head)
```

## (Optional) Sanity checks

### Realized correlation between *Y* and (log TTE) among enrolled

This will typically be weaker than the design `rho` due to censoring and sampling variation.

```{r correlation_check}
locked_final <- trial$locked_data[[paste0("time_", final_time)]]

locked_final |>
  filter(!is.na(enroll_time)) |>
  mutate(log_tte_true = log(tte_true)) |>
  group_by(arm) |>
  summarise(
    n = n(),
    cor_y_logtte = cor(y, log_tte_true)
  )
```

### KM curves (illustrative)

```{r km_plot, message=FALSE, warning=FALSE}
locked_final <- trial$locked_data[[paste0("time_", final_time)]]

# Build observed times/status at final time for KM
if (!is.null(locked_final)) {
  df_e <- locked_final |> filter(!is.na(enroll_time))
  if (nrow(df_e) > 0) {
    censor_drop  <- ifelse(is.na(df_e$drop_time), Inf, pmax(0, df_e$drop_time - df_e$enroll_time))
    censor_admin <- pmax(0, final_time - df_e$enroll_time)
    tte_true <- df_e$tte_true
    t_obs    <- pmin(tte_true, censor_drop, censor_admin)
    status   <- as.integer(tte_true <= pmin(censor_drop, censor_admin))

    S <- Surv(time = t_obs, event = status)
    fit <- survfit(S ~ arm, data = df_e)
    plot(fit, col = c("black", "steelblue"), lwd = 2, xlab = "Follow-up time", ylab = "Survival")
    legend("bottomleft", legend = levels(factor(df_e$arm)), col = c("black", "steelblue"), lwd = 2, bty = "n")
  }
}
```
