---
title: "Example 2: Two arm | Interim | Continuous | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2: Two arm | Interim | Continuous | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms <- c("pbo", "trt")
allocation <- c(1,1)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate = c(5, 10, 15, 20)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate = c(1, 2, 4, 8)
)
```

# Time points
Get a list of time points where important events occur. Add time points to a timer.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size,
  arms,
  allocation,
  enrollment,
  dropout
)

t <- Timer$new(name = "trial_timer")
add_timepoints(t, timepoints)
```

## Arms
Generate populations and endpoint data.

```{r population}
control <- Population$new(
  name = "pbo",
  data = data.frame(
    id = 1:50,
    value = rnorm(50),
    readout_time = 1
  )
)

treatment <- Population$new(
  name = "trt",
  data = data.frame(
    id = 1:50,
    value = rnorm(50, 0.2),
    readout_time = 1
  )
)
```

## Triggers & Analysis
We want to do a t-test when 100 subjects have been enrolled. 

```{r trigger}
t$add_condition(
  time %in% c(6, 12),
  analysis = function(df, time){
    df_enrolled <- df |> dplyr::filter(!is.na(enroll_time))
    t.test(value ~ arm, data = df_enrolled)$p.value
  },
  name = "t-test_final"
)
```

## Trial
Make the trial object.

```{r trial}
trial <- Trial$new(
  name = "test_trial",
  timer = t,
  population = list(control, treatment)
)
```

## Simluate
To simulate 1 replicate call:

```{r sim}
trial$run()
```

Get results.
```{r results}
trial$results
```

Get locked data.
```{r locked}
lapply(trial$locked_data, head)
```
