---
title: "Example Multi Run: Two arm | Interim | Continuous | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Multi Run: Two arm | Interim | Continuous | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms        <- c("pbo", "trt")
allocation  <- c(1, 1)
enrollment  <- list(end_time = c(4, 8, 12, 16), rate = c(5, 10, 15, 20))
dropout     <- list(end_time = c(4, 8, 12, 16), rate = c(1, 2, 4, 8))


```

# Time points
Get a list of time points where important events occur. Add time points to a timer.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size,
  arms,
  allocation,
  enrollment,
  dropout
)

t <- Timer$new(name = "trial_timer")
add_timepoints(t, timepoints)
```

## Arms
Generate populations and endpoint data using generators (functions).

```{r population}
#update the generators to have sample size 50
gen_control <- function() {
  data.frame(
    subject_id = 1:50,
    value = stats::rnorm(50)
  )
}

gen_trt <- function() {
  data.frame(
    subject_id = 1:50,
    value = stats::rnorm(50, mean = 0.2)
  )
}

pops_gen <- list(
  list("pbo", gen_control),
  list("trt", gen_trt,parms=..)
)

```

## Triggers & Analysis
We want to do a t-test when 100 subjects have been enrolled. 

```{r trigger}
t$add_condition(
  time %in% c(6, 12),
  analysis = function(df, time){
    df_enrolled <- df |> dplyr::filter(!is.na(enroll_time))
    t.test(value ~ arm, data = df_enrolled)$p.value
  },
  name = "t-test_final"
)
```

## Trial
Make the trial objects for when data is using generators.

```{r trial}
trials <- create_multiple_trials_gen_pop(
  trial_name    = "trial",
  data_gen_list = pops_gen,
  timer         = t,
  n             = 100
)
```

## Simluate
To run a list of trial objects us the run_multiple_trial function

```{r sim}
 run_multiple_trial(trials)
```

Get results of first trial.
```{r results}
trials[[1]]$results
```

Get locked data of first trial.
```{r locked}
lapply(trials[[1]]$locked_data, head)
```
