---
title: "Example 1: Two arm | Fixed design | Continuous"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms <- c("pbo", "trt")
allocation <- c(1,1)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate = c(5, 10, 15, 20)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate = c(1, 2, 4, 8)
)
```

# Time points
Get a list of time points where important events occur. Add time points to a timer.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size,
  arms,
  allocation,
  enrollment,
  dropout
)

t <- Timer$new(name = "trial_timer")
add_timepoints(t, timepoints)
```

## Arms
Generate populations and endpoint data.

```{r population}
control <- Population$new(
  name = "pbo",
  data = data.frame(
    subject_id = 1:50,
    data = rnorm(50)
  )
)

treatment <- Population$new(
  name = "trt",
  data = data.frame(
    subject_id = 1:50,
    data = rnorm(50, 0.15)
  )
)
```

## Triggers & Analysis
We want to do a t-test when 100 subjects have been enrolled. 

```{r trigger}
t$add_condition(
  nrow(!is.na(enroll_time)) >= 100,
  analysis = function(df, time){
    df |> 
  }
)
