---
title: "Example 4: Two arm | Fixed design | Correlated continuous & time-to-event endpoints | t-test + survival analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 4: Two arm | Fixed design | Correlated continuous & time-to-event endpoints | t-test + survival analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)      # for mvrnorm (simulate correlated latent variables)
library(survival)  # for Surv, survdiff, coxph
```

Oncology and cardiovascular trials often co-primary a continuous biomarker
alongside a time-to-event (TTE) endpoint such as progression-free survival or
overall survival. Correlation between a biomarker and TTE arises naturally
through shared biological mechanisms. This example demonstrates simulating such
a joint endpoint structure and applying both a t-test (for the continuous
component) and a log-rank test plus Cox proportional hazards model (for TTE) at
a single final analysis. See [Example 3](example-3.html) for the analogous
two-correlated-continuous setup.

## Scenario

We consider a **two-arm fixed design** with a **continuous endpoint (Y)** and a
**time-to-event (TTE)** endpoint per subject. The endpoints are **correlated**
via a shared Gaussian latent structure. Enrollment is piece-wise linear;
analysis happens **once at the final time**.

```{r scenario}
# Total target N (across both arms)
sample_size <- 1200

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)

# Correlation between continuous endpoint driver (Z1) and frailty for TTE (Z2)
rho <- 0.50

# Continuous endpoint parameters
mu_pbo <- 0.0
delta  <- 0.3      # true treatment - placebo mean difference
sd_y   <- 1.0

# TTE model parameters (Exponential PH with subject frailty exp(eta))
lambda0 <- 0.08    # baseline hazard (per time unit)
HR_trt  <- 0.70    # treatment hazard ratio (<1 favors treatment)

# Frailty scale (eta = sigma_frailty * Z2)
sigma_frailty <- 0.6

# Covariance of (Z1, Z2)
SigmaZ <- matrix(c(1, rho, rho, 1), nrow = 2)

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation  = list(allocation),
  delta       = delta,
  rho         = rho,
  hr_trt_true = HR_trt
)
```

`HR_trt = 0.70` encodes a 30% hazard reduction in the treatment arm — a
clinically meaningful effect in many oncology settings. `lambda0 = 0.08` is the
baseline event rate per time unit; combined with n=1200, this ensures a
substantial number of TTE events even with only 3 replicates. `sigma_frailty =
0.6` controls the strength of the subject-level frailty term: larger values
introduce more between-subject heterogeneity in TTE, making the frailty the
shared biological driver linking the continuous and TTE endpoints through the
bivariate latent (Z1, Z2) with `rho = 0.50`.

## Time points

Generate the discrete timepoints and add them to a `Timer`.

```{r timepoints}
plan <- gen_plan(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment_fn,
  dropout     = dropout_fn
)
```

`gen_plan()` is called with `enrollment_fn` as a stochastic function, so each
invocation by `replicate_trial()` generates a fresh per-replicate enrollment
schedule. This differs from the piecewise-constant list approach in [Example
3](example-3.html): here there is no deterministic timer pre-built — enrollment
timing varies across replicates, which is the more realistic simulation regime
for large trials.

## Arms (Populations)

Per arm we simulate subject-level latent bivariate normals `(Z1, Z2)` with
correlation `rho`. - The **continuous** endpoint is `Y = mu_arm + sd_y * Z1`. -
The **TTE** endpoint uses a proportional-hazards exponential model with
subject-specific frailty: `hazard_i = lambda_arm * exp(eta_i)`, where `eta_i =
sigma_frailty * Z2` and `lambda_trt = lambda0 * HR_trt`. Event time `T_i ~
Exponential(rate = hazard_i)`.

```{r population}
# Create generator function for arm-specific data
mk_population_generator <- function(mu_y, lambda_arm) {
  function(n) {
    Z <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = SigmaZ)
    Z1 <- Z[, 1]
    Z2 <- Z[, 2]
    y  <- mu_y + sd_y * Z1
    eta <- sigma_frailty * Z2
    rate_i <- lambda_arm * exp(eta)
    tte_true <- rexp(n, rate = rate_i)

    data.frame(
      id = seq_len(n),
      y  = y,
      tte_true = tte_true,
      readout_time = 1
    )
  }
}

lambda_pbo <- lambda0
lambda_trt <- lambda0 * HR_trt

population_generators <- list(
  pbo = mk_population_generator(mu_y = mu_pbo, lambda_arm = lambda_pbo),
  trt = mk_population_generator(mu_y = mu_pbo + delta, lambda_arm = lambda_trt)
)
```

Each subject gets a latent pair (Z1, Z2) drawn from a bivariate standard normal
with correlation `rho`. Z1 drives the continuous endpoint: `Y = mu_arm + sd_y *
Z1`. Z2 drives TTE through a multiplicative frailty: `eta = sigma_frailty * Z2`,
so `hazard_i = lambda_arm * exp(eta_i)`, and `T_i ~ Exponential(hazard_i)`. The
shared latent structure means subjects with higher Z2 (higher frailty, shorter
TTE) also tend to have higher Z1 (higher Y), modelling a biological scenario
where the biomarker and TTE share a common underlying driver.

## Trigger & Analysis

This is a **fixed design**: analyze **once at the final timepoint**. We run a
**two-sample t-test** for the continuous endpoint and a **log-rank test** plus
**Cox PH** for the TTE endpoint.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      df_e <- df[!is.na(df$enroll_time), , drop = FALSE]
      if (nrow(df_e) == 0) {
        return(data.frame(scenario, note = "no enrolled subjects", stringsAsFactors = FALSE))
      }

      # Build observed TTE with censoring
      censor_drop  <- ifelse(is.na(df_e$drop_time), Inf, pmax(0, df_e$drop_time - df_e$enroll_time))
      censor_admin <- pmax(0, max(df_e$enroll_time) + 100 - df_e$enroll_time)

      tte_true <- df_e$tte_true
      t_obs    <- pmin(tte_true, censor_drop, censor_admin)
      status   <- as.integer(tte_true <= pmin(censor_drop, censor_admin))

      # Continuous endpoint t-test
      p_t <- tryCatch(
        t.test(y ~ arm, data = df_e)$p.value,
        error = function(e) NA_real_
      )

      # Survival analysis: log-rank and Cox
      surv_p <- NA_real_
      hr <- hr_lo <- hr_hi <- NA_real_
      events <- sum(status)

      if (length(unique(df_e$arm)) == 2 && events > 0 && all(t_obs >= 0)) {
        S <- Surv(time = t_obs, event = status)

        surv_p <- tryCatch({
          sd <- survdiff(S ~ arm, data = df_e)
          1 - pchisq(sd$chisq, df = length(sd$n) - 1)
        }, error = function(e) NA_real_)

        cox <- tryCatch(coxph(S ~ arm, data = df_e), error = function(e) NULL)
        if (!is.null(cox)) {
          s <- summary(cox)
          hr    <- unname(s$coef[1, "exp(coef)"])
          hr_lo <- unname(s$conf.int[1, "lower .95"])
          hr_hi <- unname(s$conf.int[1, "upper .95"])
        }
      }

      data.frame(
        scenario,
        n_total    = nrow(df_e),
        n_pbo      = sum(df_e$arm == "pbo"),
        n_trt      = sum(df_e$arm == "trt"),
        p_ttest_y  = p_t,
        logrank_p  = surv_p,
        hr_cox     = hr,
        hr_ci_lo   = hr_lo,
        hr_ci_hi   = hr_hi,
        events     = events,
        stringsAsFactors = FALSE
      )
    }
  )
)
```

`censor_drop` censors a subject at the time elapsed between enrollment and
dropout (or infinity if no dropout). `censor_admin` provides administrative
censoring: each subject is followed for 100 time units after the last
enrollment, a common end-of-study rule. `t_obs` and `status` capture what would
actually be observed. The `Surv()` object is then passed to `survdiff()` for the
log-rank p-value and `coxph()` for the Cox HR and 95% CI. `tryCatch` guards
against edge cases (e.g., one arm with no events) so the simulation continues
rather than stopping on an error.

## Trial

Create and run multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "two_endpoints_continuous_tte_fixed",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r sim}
run_trials(trials)
```

## Results

One row per replicate, row-bound across all replicates:

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`collect_results()` prepends `replicate`, `timepoint`, and `analysis` columns to
the result data. The `events` column shows how many TTE events were observed —
with n=1200 and `lambda0 = 0.08`, this should be large. `logrank_p` tests the
null of equal survival curves; `hr_cox` is the estimated hazard ratio from the
Cox model (values < 1 favour treatment), and `hr_ci_lo`/`hr_ci_hi` are its 95%
CI bounds. Because the continuous endpoint Y and TTE share the same latent
drivers, replicates with a small `p_ttest_y` tend to also have a small
`logrank_p`, though the correspondence is imperfect due to the different
statistical tests and the censoring mechanism.
