---
title: "Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Bayesian borrowing utilities
library(RBesT)

# Helpers
library(dplyr)

set.seed(777)
```

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single continuous endpoint** and a **single final analysis**.

- Placebo arm uses **historical borrowing** through an informative prior derived from historical placebo data and robustified with a non-informative component.
- Treatment arm uses a **non-informative prior**.
- Go/No-Go rule at final analysis:

\[
\text{Go if } P(\Delta > \delta \mid \text{data}) \ge \gamma,
\]

where \(\Delta = \mu_T - \mu_P\), with \(\delta = 0.1\) and \(\gamma = 0.8\).

```{r scenario}
# Trial design
sample_size <- 80
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Data-generating truth for this simulation example
mu_placebo_true <- 0.00
mu_treat_true   <- 0.30
sigma_known     <- 1.0

# Enrollment/dropout profile (fixed design timeline)
enrollment <- list(
  end_time = c(4, 8),
  rate     = c(6, 10)
)
dropout <- list(
  end_time = c(4, 8),
  rate     = c(0, 0)
)

# Bayesian decision thresholds
delta <- 0.1
gamma <- 0.8

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  delta = delta,
  gamma = gamma
)
```

## Historical placebo dataset and priors

We include a small historical placebo dataset as study-level means and sample sizes.

```{r priors}
hist_placebo <- data.frame(
  study = c("H1", "H2", "H3"),
  n     = c(24, 18, 20),
  mean  = c(-0.05, 0.02, 0.00)
)

hist_placebo

# Pooled historical summary used to derive an informative placebo prior
n_hist_total <- sum(hist_placebo$n)
m_hist_pooled <- weighted.mean(hist_placebo$mean, hist_placebo$n)

# Non-informative prior (very small pseudo-sample size in mn parametrization)
# Used as treatment prior and as base for historical update.
prior_noninf <- mixnorm(
  c(1, 0, 1e-6),
  sigma = sigma_known,
  param = "mn"
)

# Informative placebo prior from historical placebo summary
prior_placebo_inf <- postmix(
  prior_noninf,
  n = n_hist_total,
  m = m_hist_pooled
)

# Robustified placebo prior to protect against prior-data conflict
prior_placebo <- robustify(
  prior_placebo_inf,
  weight = 0.2,
  mean   = 0,
  n      = 1,
  sigma  = sigma_known
)

# Treatment prior remains non-informative
prior_treatment <- prior_noninf

summary(prior_placebo)
```

## Populations

```{r population}
population_generators <- list(
  placebo = function(n) {
    data.frame(
      id = 1:n,
      y = rnorm(n, mean = mu_placebo_true, sd = sigma_known),
      readout_time = 1
    )
  },
  treatment = function(n) {
    data.frame(
      id = 1:n,
      y = rnorm(n, mean = mu_treat_true, sd = sigma_known),
      readout_time = 1
    )
  }
)
```

## Trial Parameters

```{r trial_params}
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
```

## Final analysis trigger (Bayesian Go/No-Go)

Only one trigger is added at final analysis.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      dat <- df |>
        dplyr::filter(!is.na(enroll_time)) |>
        dplyr::mutate(arm = factor(arm, levels = c("placebo", "treatment")))

      y_p <- dat$y[dat$arm == "placebo"]
      y_t <- dat$y[dat$arm == "treatment"]

      n_p <- length(y_p)
      n_t <- length(y_t)

      m_p <- mean(y_p)
      m_t <- mean(y_t)

      # Posterior for each arm mean
      post_placebo <- RBesT::postmix(prior_placebo, n = n_p, m = m_p)
      post_treat   <- RBesT::postmix(prior_treatment, n = n_t, m = m_t)

      # Posterior probability for treatment effect exceeding delta
      prob_delta <- RBesT::pmixdiff(post_treat, post_placebo, delta, lower.tail = FALSE)

      # Posterior summaries for Delta = mu_T - mu_P
      delta_ci <- RBesT::qmixdiff(post_treat, post_placebo, c(0.025, 0.5, 0.975))

      data.frame(
        scenario,
        n_placebo = n_p,
        n_treatment = n_t,
        mean_placebo = m_p,
        mean_treatment = m_t,
        post_prob_delta = prob_delta,
        delta_q025 = delta_ci[1],
        delta_q500 = delta_ci[2],
        delta_q975 = delta_ci[3],
        decision_go = as.integer(prob_delta >= gamma),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Trial

```{r trial}
trials <- replicate_trial(
  trial_name = "example_7_bayes_two_arm_fixed",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r simulate}
run_trials(trials)
```

## Results

```{r results}
replicate_results <- do.call(rbind, lapply(seq_along(trials), function(i) {
  data.frame(
    replicate = i,
    trials[[i]]$results[[1]]$final,
    stringsAsFactors = FALSE
  )
}))

replicate_results
```

`decision_go = 1` indicates **Go**, and `decision_go = 0` indicates **No-Go** under the criterion
\(P(\Delta > 0.1 \mid \text{data}) \ge 0.8\).
