---
title: "Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Bayesian borrowing utilities
library(RBesT)

# Helpers
library(dplyr)

set.seed(777)
```

Early development trials often face small sample sizes where borrowing
information from historical placebo data can increase efficiency and reduce
required sample size. This example demonstrates a Bayesian Go/No-Go decision
rule for a Phase IIa trial, where the placebo posterior is informed by
historical data via a robust mixture prior, and the treatment posterior uses a
non-informative prior. A "Go" decision is issued when the posterior probability
of a meaningful treatment effect (Δ > δ) exceeds a predefined threshold γ.

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single
continuous endpoint** and a **single final analysis**.

- Placebo arm uses **historical borrowing** through an informative prior derived
  from historical placebo data and robustified with a non-informative component.
- Treatment arm uses a **non-informative prior**.
- Go/No-Go rule at final analysis:

\[
\text{Go if } P(\Delta > \delta \mid \text{data}) \ge \gamma,
\]

where \(\Delta = \mu_T - \mu_P\), with \(\delta = 0.1\) and \(\gamma = 0.8\).

`delta = 0.1` defines the minimum treatment-placebo difference of clinical
relevance — the effect must exceed this threshold for the drug to be worth
advancing. `gamma = 0.8` requires 80% posterior probability of exceeding that
threshold before a Go decision is issued, balancing false-positive risk against
the cost of a missed opportunity. The true simulated effect `mu_treat = 0.30`
lies above `delta`, so a well-powered trial should frequently result in a Go.

```{r scenario}
# Trial design
sample_size <- 80
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Data-generating truth for this simulation example
mu_placebo_true <- 0.00
delta_true      <- 0.30  # true treatment - placebo mean difference
sigma_known     <- 1.0

# Enrollment/dropout profile (fixed design timeline)
enrollment <- list(
  end_time = c(4, 8),
  rate     = c(6, 10)
)
dropout <- list(
  end_time = c(4, 8),
  rate     = c(0, 0)
)

# Bayesian decision thresholds
delta <- 0.1
gamma <- 0.8

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation  = list(allocation),
  delta_true  = delta_true,
  delta       = delta,
  gamma       = gamma
)
```

## Historical placebo dataset and priors

We include a small historical placebo dataset as study-level means and sample
sizes.

Studies H1, H2, and H3 contribute a combined 62 historical placebo observations
and are summarised into an informative prior via `postmix()` — a Bayesian update
of the non-informative starting prior with the pooled historical data.
`robustify()` blends this informative component with a vague component (20%
weight), producing a mixture prior that protects against prior-data conflict: if
the current trial's placebo behaves unexpectedly, the vague component limits how
strongly the historical data pulls the posterior. The treatment arm uses
`prior_noninf` (effectively a flat prior), so the treatment posterior is driven
entirely by the current data.

```{r priors}
hist_placebo <- data.frame(
  study = c("H1", "H2", "H3"),
  n     = c(24, 18, 20),
  mean  = c(-0.05, 0.02, 0.00)
)

hist_placebo

# Pooled historical summary used to derive an informative placebo prior
n_hist_total <- sum(hist_placebo$n)
m_hist_pooled <- weighted.mean(hist_placebo$mean, hist_placebo$n)

# Non-informative prior (very small pseudo-sample size in mn parametrization)
# Used as treatment prior and as base for historical update.
prior_noninf <- mixnorm(
  c(1, 0, 1e-6),
  sigma = sigma_known,
  param = "mn"
)

# Informative placebo prior from historical placebo summary
prior_placebo_inf <- postmix(
  prior_noninf,
  n = n_hist_total,
  m = m_hist_pooled
)

# Robustified placebo prior to protect against prior-data conflict
prior_placebo <- robustify(
  prior_placebo_inf,
  weight = 0.2,
  mean   = 0,
  n      = 1,
  sigma  = sigma_known
)

# Treatment prior remains non-informative
prior_treatment <- prior_noninf

summary(prior_placebo)
```

## Populations

```{r population}
population_generators <- list(
  placebo = function(n) {
    data.frame(
      id = 1:n,
      y = rnorm(n, mean = mu_placebo_true, sd = sigma_known),
      readout_time = 1
    )
  },
  treatment = function(n) {
    data.frame(
      id = 1:n,
      y = rnorm(n, mean = mu_placebo_true + delta_true, sd = sigma_known),
      readout_time = 1
    )
  }
)
```

## Trial Parameters

`enrollment_fn` and `dropout_fn` supply the inter-arrival and time-to-dropout
distributions for individual subjects. These are passed to `replicate_trial()`,
which uses them when building each replicate's enrollment schedule.

```{r trial_params}
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
```

## Final analysis trigger (Bayesian Go/No-Go)

Only one trigger is added at final analysis.

`RBesT::pmixdiff(post_treat, post_placebo, delta, lower.tail = FALSE)` evaluates
P(μ_T − μ_P > δ | data) by numerical integration over the mixture posterior
distributions for treatment and placebo. `qmixdiff()` computes quantiles of the
same posterior difference distribution, yielding a 95% credible interval for Δ =
μ_T − μ_P.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      dat <- df |>
        dplyr::filter(!is.na(enroll_time)) |>
        dplyr::mutate(arm = factor(arm, levels = c("placebo", "treatment")))

      y_p <- dat$y[dat$arm == "placebo"]
      y_t <- dat$y[dat$arm == "treatment"]

      n_p <- length(y_p)
      n_t <- length(y_t)

      m_p <- mean(y_p)
      m_t <- mean(y_t)

      # Posterior for each arm mean
      post_placebo <- RBesT::postmix(prior_placebo, n = n_p, m = m_p)
      post_treat   <- RBesT::postmix(prior_treatment, n = n_t, m = m_t)

      # Posterior probability for treatment effect exceeding delta
      prob_delta <- RBesT::pmixdiff(post_treat, post_placebo, delta, lower.tail = FALSE)

      # Posterior summaries for Delta = mu_T - mu_P
      delta_ci <- RBesT::qmixdiff(post_treat, post_placebo, c(0.025, 0.5, 0.975))

      data.frame(
        scenario,
        n_placebo = n_p,
        n_treatment = n_t,
        mean_placebo = m_p,
        mean_treatment = m_t,
        post_prob_delta = prob_delta,
        delta_q025 = delta_ci[1],
        delta_q500 = delta_ci[2],
        delta_q975 = delta_ci[3],
        decision_go = as.integer(prob_delta >= gamma),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Trial

```{r trial}
trials <- replicate_trial(
  trial_name = "example_7_bayes_two_arm_fixed",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r simulate}
run_trials(trials)
```

## Results

`collect_results()` row-binds analysis outputs across all replicates and
prepends `replicate` (integer index), `timepoint` (calendar time at which the
analysis fired), and `analysis` (the analysis name) to each row.
`post_prob_delta` is the key decision metric: values at or above `gamma = 0.8`
yield `decision_go = 1` (Go) and values below yield `decision_go = 0` (No-Go).
The credible interval columns `delta_q025`, `delta_q500`, and `delta_q975`
characterise the posterior uncertainty about the treatment effect Δ; with n=80
and σ=1, the posterior is still relatively wide. In practice you would run
thousands of replicates and report the Go probability under both the null (Δ =
0) and the alternative (e.g., Δ = 0.30) to assess the design's operating
characteristics. See [Example 8](example-8.html) for a seamless design that
builds on this Bayesian decision rule.

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`decision_go = 1` indicates **Go**, and `decision_go = 0` indicates **No-Go**
under the criterion \(P(\Delta > 0.1 \mid \text{data}) \ge 0.8\).
