---
title: "Example 12: Two arm | Interim design | Binary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 12: Two arm | Interim design | Binary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(rxsim)
```

This example extends [Example 9](https://www.example.com) by introducing two named analyses — `interim` and `final` — that fire at different enrollment milestones. Understanding how analysis names propagate through `collect_results()` is the foundation for multi-analysis designs shown in later examples.

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms <- c("pbo", "trt")
p_pbo <- 0.1
p_trt <- 0.18
allocation <- c(1,1)
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  p_pbo = p_pbo,
  p_trt = p_trt
)
```

`allocation = c(1, 1)` specifies balanced randomisation. `tidyr::expand_grid()` embeds design parameters directly into each result row for traceability across parameter sweeps.

## Populations
Define population generators.

```{r population}
population_generators <- list(
  pbo = function(n) data.frame(
    id = 1:n,
    value = rbinom(n, size = 1, prob = p_pbo),
    readout_time = 1
  ),
  trt = function(n) data.frame(
    id = 1:n,
    value = rbinom(n, size = 1, prob = p_trt),
    readout_time = 1
  )
)
```

Each generator is a function of `n` returning a `data.frame` with one row per subject. `readout_time = 1` means the endpoint is observed 1 time unit after enrollment. The endpoint is binary, with a event probability of `p_pbo = 0.10` in the placebo arm and `p_trt = 0.18` in the treatment arm, corresponding to an absolute risk difference of 0.08 in favor of treatment.

## Triggers & Analysis
Final analysis at full enrollment.

```{r trigger}
analysis_generators <- list(
  interim = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= (!!sample_size) / 2
    ),
    analysis = function(df, timer){
      df_enrolled <- df |> subset(!is.na(enroll_time))
      
      logreg <- glm(value ~ arm, data = df_enrolled, family = "binomial") #logistic regression
      odds_ratio <- exp(coef(logreg)["armtrt"])
      logreg_p <- summary(logreg)$coefficients["armtrt", "Pr(>|z|)"]/2
      logreg_p <- ifelse(odds_ratio > 1, logreg_p, 0.999) # one-sided p-value
      
      data.frame(
        scenario,
        n_total = nrow(df_enrolled),
        mean_pbo = mean(df_enrolled$value[df_enrolled$arm == "pbo"]),
        mean_trt = mean(df_enrolled$value[df_enrolled$arm == "trt"]),
        odds_ratio = unname(odds_ratio),
        p_value = unname(logreg_p),
        stringsAsFactors = FALSE
      )
    }
  ),
  
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer){
      df_enrolled <- df |> subset(!is.na(enroll_time))
      
      logreg <- glm(value ~ arm, data = df_enrolled, family = "binomial") #logistic regression
      odds_ratio <- exp(coef(logreg)["armtrt"])
      logreg_p <- summary(logreg)$coefficients["armtrt", "Pr(>|z|)"]/2
      logreg_p <- ifelse(odds_ratio > 1, logreg_p, 0.999) # one-sided p-value
      
      data.frame(
        scenario,
        n_total = nrow(df_enrolled),
        mean_pbo = mean(df_enrolled$value[df_enrolled$arm == "pbo"]),
        mean_trt = mean(df_enrolled$value[df_enrolled$arm == "trt"]),
        odds_ratio = unname(odds_ratio),
        p_value = unname(logreg_p),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

Two analyses are defined. The `interim` analysis fires at half enrollment and the `final` analysis fires when all subjects are enrolled. These names appear in the `analysis` column of `collect_results()` output, making it straightforward to distinguish interim from final results when stacking rows across timepoints.

## Trial
Make multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "test_trial",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate
To simulate all replicates:

```{r sim}
run_trials(trials)
```

Bind one row per replicate into one data frame.
```{r results}
replicate_results <- do.call(rbind, lapply(seq_along(trials), function(i) {
  data.frame(
    replicate = i,
    trials[[i]]$results[[1]]$interim,
    trials[[i]]$results[[1]]$final,
    stringsAsFactors = FALSE
  )
}))

replicate_results

```

`collect_results()` prepends `replicate`, `timepoint`, and `analysis` columns to your result data. Each row shows either `"interim"` or `"final"` in the `analysis` column, reflecting which named analysis produced it. This naming convention becomes essential in later examples where multiple named analyses fire per replicate.
