---
title: "Example 3: Two arm | Fixed design | Two correlated continuous endpoints | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 3: Two arm | Fixed design | Two correlated continuous endpoints | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)   # for mvrnorm (simulate correlated endpoints)
```

## Scenario

We consider a **two-arm, fixed design** with **two correlated continuous endpoints** per subject.  
Enrollment is piece-wise linear; analysis happens **once at the final time**.

```{r scenario}
# Total target sample size
sample_size <- 100

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

# Piece-wise linear enrollment and dropout rates (per unit time)
# (You can tweak these to match your program's cadence.)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(5, 10, 15, 20)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(1, 2, 4, 8)
)

# Endpoint model parameters
# Endpoint correlation structure (common across arms for simplicity)
rho  <- 0.60            # correlation between endpoint1 and endpoint2
sd1  <- 1.00            # SD of endpoint1
sd2  <- 1.00            # SD of endpoint2

# Mean structure per arm
mu1_pbo <- 0.00         # Control mean for endpoint1
mu2_pbo <- 0.00         # Control mean for endpoint2

# Treatment effects (mean shifts)
delta1  <- 0.30         # Treatment - control difference for endpoint1
delta2  <- 0.20         # Treatment - control difference for endpoint2

mu1_trt <- mu1_pbo + delta1
mu2_trt <- mu2_pbo + delta2

# Construct covariance matrix
Sigma <- matrix(
  c(sd1^2, rho*sd1*sd2,
    rho*sd1*sd2, sd2^2),
  nrow = 2, byrow = TRUE
)
```

## Time points

Generate timeline of discrete timepoints and add them to a `Timer`.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

t <- Timer$new(name = "trial_timer_two_endpoints")
add_timepoints(t, timepoints)

final_time <- t$get_end_timepoint()
final_time
```

## Arms (Populations)

For each arm we simulate **two correlated endpoints** per subject using `MASS::mvrnorm`.  
We keep **50** subjects per arm (consistent with `sample_size = 100` and equal allocation).

```{r population}
set.seed(2026)  # Seed local to data simulation; Trial also allows its own seed

n_per_arm <- sample_size / length(arms)  # 50 per arm

# Control population data (two correlated endpoints)
xy_pbo <- MASS::mvrnorm(
  n     = n_per_arm,
  mu    = c(mu1_pbo, mu2_pbo),
  Sigma = Sigma
)
control_df <- data.frame(
  id = seq_len(n_per_arm),
  y1 = xy_pbo[, 1],
  y2 = xy_pbo[, 2],
  readout_time = 1
)

# Treatment population data (shifted means, same covariance)
xy_trt <- MASS::mvrnorm(
  n     = n_per_arm,
  mu    = c(mu1_trt, mu2_trt),
  Sigma = Sigma
)
treatment_df <- data.frame(
  id = seq_len(n_per_arm),
  y1 = xy_trt[, 1],
  y2 = xy_trt[, 2],
  readout_time = 1
)

# Build Population objects
control <- Population$new(
  name = "pbo",
  data = control_df
)

treatment <- Population$new(
  name = "trt",
  data = treatment_df
)
```

## Trigger & Analysis

This is a **fixed design**: perform analysis **once at the final timepoint**.  
We run two independent two-sample t-tests (one per endpoint).  
(If you want multiplicity control, you can add `p.adjust` below.)

```{r trigger}
t$add_condition(
  time == final_time,
  analysis = function(df, time) {
    df_e <- df |> dplyr::filter(!is.na(enroll_time))

    # Two-sample t-tests by arm for each endpoint
    p1 <- t.test(y1 ~ arm, data = df_e)$p.value
    p2 <- t.test(y2 ~ arm, data = df_e)$p.value

    # Optional: simple multiplicity adjustment (Holm)
    padj <- p.adjust(c(p1, p2), method = "holm")

    # Return a named, row-like object (list or data.frame both fine)
    list(
      p_y1      = unname(p1),
      p_y2      = unname(p2),
      p_holm_y1 = unname(padj[1]),
      p_holm_y2 = unname(padj[2]),
      n_total   = nrow(df_e),
      n_pbo     = sum(df_e$arm == "pbo"),
      n_trt     = sum(df_e$arm == "trt")
    )
  },
  name = "final"
)
```

## Trial

Create and run the trial.  
(We supply a `seed` for reproducibility inside the trial engine—e.g., enrollment/dropout sampling.)

```{r trial}
trial <- Trial$new(
  name       = "two_endpoints_fixed_design",
  seed       = 123,
  timer      = t,
  population = list(control, treatment)
)
```

## Simulate

```{r sim}
trial$run()
```

## Results

The analysis results at the final timepoint:

```{r results}
trial$results
```

If you prefer a flat frame (when returning lists/records), you can prettify:

```{r results_pretty}
prettify_results(trial$results)
```

## Locked data (snapshot)

Inspect the data that was “locked” at analysis time:

```{r locked}
lapply(trial$locked_data, head)
```

## (Optional) Check realized correlation by arm among enrolled

This verifies the simulated correlation structure **among those enrolled** (pooled across all time up to final):

```{r correlation_check, message=FALSE}
locked_final <- trial$locked_data[[paste0("time_", final_time)]]

locked_final |>
  filter(!is.na(enroll_time)) |>
  group_by(arm) |>
  summarise(
    n         = n(),
    cor_y1_y2 = cor(y1, y2)
  )
```

### Notes
- This example uses **independent two-sample t-tests** for each endpoint. You can adapt the `analysis` function to use **multivariate** methods (e.g., Hotelling’s T^2 via `HotellingsT2::HotellingsT2`) or to implement different **multiplicity adjustments** depending on your program’s control strategy.
- The `MASS::mvrnorm` generation ensures **correlated endpoints** per subject with correlation `rho`. Realized sample correlations will vary around `rho` due to sampling and enrollment patterns with dropout.
