---
title: "Example 5: Multi-arm | Fixed design | Single continuous endpoint | Dunnett test + MCP-Mod"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 5: Multi-arm | Fixed design | Single continuous endpoint | Dunnett test + MCP-Mod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core simulation framework (Timer, Population, Trial, gen_timepoints, add_timepoints, ...)
library(rxsim)

# Analyses
library(multcomp)     # Dunnett
library(DoseFinding)  # MCP-Mod

# Helpers
library(dplyr)
set.seed(4566)
```

Dose-finding trials aim to identify the dose-response relationship and establish
an effective dose level. This example simulates a multi-arm fixed design across
a placebo and four active doses, with the endpoint generated from an Emax
dose-response model. At the final analysis, two methods are applied: a Dunnett
test (comparing each active dose against placebo while controlling family-wise
error rate) and MCP-Mod (a model-based contrast approach that selects the
best-fitting dose-response shape from a candidate set).

## Scenario

A **multi-arm, fixed design** (placebo + 4 active doses) with a **single
continuous endpoint**. At the **final analysis**, we perform:

- **Dunnett test**: all active doses vs placebo using a normal-theory linear
  model.
- **MCP-Mod**: model-based multiple contrast test + model fitting on a candidate
  set of dose–response shapes.

The Emax model generates mean responses via `E(d) = e0 + emax × d / (ed50 + d)`,
where `e0 = 0` (placebo baseline), `emax = 1` (maximum achievable effect), and
`ed50 = 20` (dose at half-maximum effect). `arm_names = paste0("d", doses)`
creates human-readable labels (`d0`, `d5`, `d10`, `d20`, `d50`) that propagate
through trial data and analysis results. `delta = 0.1` is the minimum effect
size of clinical relevance used by MCP-Mod, and `alpha = 0.05` controls the
family-wise Type I error rate for both testing procedures.

```{r scenario}
# Dose levels (placebo + 4 actives)
doses <- c(0, 5, 10, 20, 50)

# Total N and allocation (balanced across arms)
sample_size <- 150
allocation  <- rep(1, length(doses))
arm_names   <- paste0("d", doses)  # e.g., d0, d5, ... used as arm labels

# Enrollment & dropout profile (piece-wise constant)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(6, 12, 18, 24)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(0,  1,  2,  3)
)

# Data-generating model: Emax with homoscedastic noise
e0    <- 0.0
emax  <- 1.0
ed50  <- 20.0
sigma <- 1.0

mean_fun <- function(d) e0 + emax * d / (ed50 + d)

# operating characteristics
alpha <- 0.05
delta <- 0.1

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  n_arms = length(doses),
  alpha = alpha,
  delta = delta
)
```

## Time points

`gen_timepoints()` generates a piecewise-constant enrollment schedule with rates
that increase in steps (6, 12, 18, 24 subjects per unit time), reflecting a
realistic ramp-up in site activation. `tr_timer$get_end_timepoint()` returns the
earliest calendar time at which all `sample_size` subjects are expected to be
enrolled — this value is used as the trigger time for the final analysis.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arm_names,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

# Timer
tr_timer <- Timer$new(name = "timer_multiarm")
add_timepoints(tr_timer, timepoints)
final_time <- tr_timer$get_end_timepoint()
final_time
```

## Populations

`mk_pop_gen` is a closure factory: it captures the dose value `d` and returns a
generator function that, when called with `n`, draws `n` responses from N(E(d),
σ²). Each arm's generator stores `dose` as a numeric column because MCP-Mod
requires actual dose values — not just arm labels — to fit dose-response models
and evaluate contrasts at analysis time.

```{r population}
mk_pop_gen <- function(d) {
  function(n) {
    mu <- mean_fun(d)
    y  <- rnorm(n, mean = mu, sd = sigma)
    data.frame(
      id = seq_len(n),
      dose = d,
      y = y,
      readout_time = 1
    )
  }
}

population_generators <- lapply(seq_along(doses), function(i) {
  mk_pop_gen(doses[i])
})
names(population_generators) <- arm_names
```

## Trial Parameters

`enrollment_fn` and `dropout_fn` supply the inter-arrival and time-to-dropout
distributions for individual subjects. These are passed to `replicate_trial()`,
which uses them internally when building each replicate's enrollment schedule.

```{r trial_params}
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
```

## Triggers & Analyses

At the **final time**, run:

1) **Dunnett test** (active vs placebo) using `multcomp::glht` on `lm(y ~ arm)`.

2) **MCP-Mod** using `DoseFinding::MCPMod` with a candidate model set (`Mods`).

The Dunnett test (`multcomp::glht` with `mcp(arm = "Dunnett")`) simultaneously
compares each active dose against placebo while controlling the family-wise
error rate at `alpha`. MCP-Mod is run with five candidate dose-response shapes;
`selModel = "aveAIC"` selects the best model by AIC-weighted averaging, and
`Delta = delta` sets the minimum effect size of clinical relevance for the
contrast step. `mct_min_p` extracts the minimum p-value across all candidate
model contrast tests — a small value indicates that at least one candidate model
detects a dose-response signal.

```{r trigger}
# Candidate model set for MCP-Mod
mods <- Mods(
  linear     = NULL,
  emax       = 20,
  exponential= 50,
  sigEmax    = c(20, 3),
  quadratic  = -0.2,
  doses      = doses
)

analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer){
      df_e <- df |>
        dplyr::filter(!is.na(enroll_time)) |>
        dplyr::mutate(
          arm  = factor(arm, levels = paste0("d", doses)),
          dose = as.numeric(dose)
        )

      # 1) Dunnett (active vs placebo)
      fit <- lm(y ~ arm, data = df_e)
      dun  <- multcomp::glht(fit, linfct = multcomp::mcp(arm = "Dunnett"))
      summ <- summary(dun)

      # 2) MCP-Mod (one-step)
      mm <- DoseFinding::MCPMod(
          dose   = df_e$dose,
          resp   = df_e$y,
          models = mods,
          type   = "normal",
          Delta  = delta,
          alpha  = alpha,
          selModel = "aveAIC"
      )

      mct_min_p <- min(attr(mm$MCTtest$tStat, "pVal"), na.rm = TRUE)

      data.frame(
        scenario,
        n_total      = nrow(df_e),
        dunn_min_p   = summ$test$pvalues |> as.numeric() |> min(),
        mcpmod_min_p = mct_min_p,
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Trial

```{r trial}
trials <- replicate_trial(
  trial_name = "multiarm_dunnett_mcpmod",
  sample_size = sample_size,
  arms = arm_names,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r sim}
run_trials(trials)
```

## Results

`collect_results()` row-binds analysis outputs across all replicates and
prepends three bookkeeping columns: `replicate` (integer index), `timepoint`
(calendar time at which the analysis fired), and `analysis` (the analysis name,
here `"final"`). `dunn_min_p` is the smallest Dunnett-adjusted p-value across
the four active-dose comparisons — a value below `alpha` indicates that at least
one dose is significantly different from placebo after multiplicity correction.
`mcpmod_min_p` is the minimum MCP-Mod contrast test p-value across candidate
models; both metrics test for the presence of a dose-response signal, but
MCP-Mod is generally more powerful when the true dose-response shape is
well-captured by one of the candidate models. See [Core Concepts](concepts.html)
for background on trial parameters.

One row per replicate, row-bound across all replicates:

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```
