---
title: "Example 3: Two arm | Fixed design | Two correlated continuous endpoints | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 3: Two arm | Fixed design | Two correlated continuous endpoints | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)   # for mvrnorm (simulate correlated endpoints)
```

Many trials evaluate co-primary or key secondary endpoints simultaneously.
Ignoring endpoint correlation when planning a study leads to incorrect power
estimates. This example shows how to simulate two correlated continuous
endpoints (e.g., a primary efficacy score and a key secondary biomarker) and
apply multiplicity adjustment via Holm's procedure. See [Example
1](example-1.html) for the simpler single-endpoint baseline.

## Scenario

We consider a **two-arm, fixed design** with **two correlated continuous
endpoints** per subject. Enrollment is piece-wise linear; analysis happens
**once at the final time**.

```{r scenario}
# Total target sample size
sample_size <- 100

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

# Piece-wise linear enrollment and dropout rates (per unit time)
# (You can tweak these to match your program's cadence.)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(5, 10, 15, 20)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(1, 2, 4, 8)
)

# Endpoint model parameters
# Endpoint correlation structure (common across arms for simplicity)
rho  <- 0.60            # correlation between endpoint1 and endpoint2
sd1  <- 1.00            # SD of endpoint1
sd2  <- 1.00            # SD of endpoint2

# Mean structure per arm
mu1_pbo <- 0.00         # Control mean for endpoint1
mu2_pbo <- 0.00         # Control mean for endpoint2

# Treatment effects (mean shifts)
delta1  <- 0.30         # Treatment - control difference for endpoint1
delta2  <- 0.20         # Treatment - control difference for endpoint2

mu1_trt <- mu1_pbo + delta1
mu2_trt <- mu2_pbo + delta2

# Construct covariance matrix
Sigma <- matrix(
  c(sd1^2, rho*sd1*sd2,
    rho*sd1*sd2, sd2^2),
  nrow = 2, byrow = TRUE
)

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  rho = rho,
  delta1 = delta1,
  delta2 = delta2
)
```

`rho = 0.60` represents a moderate positive correlation between the two
endpoints — plausible when both reflect the same underlying biological process.
The covariance matrix `Sigma` is constructed from the marginal SDs and
correlation using the standard identity cov(y1, y2) = rho × sd1 × sd2. Treatment
shifts are delta1 = 0.30 SD for the primary endpoint and delta2 = 0.20 SD for
the secondary, reflecting a scenario where the primary endpoint is more
sensitive to treatment.

## Time points

Generate timeline of discrete timepoints and add them to a `Timer`.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

t <- Timer$new(name = "trial_timer_two_endpoints")
add_timepoints(t, timepoints)

final_time <- t$get_end_timepoint()
final_time
```

`gen_timepoints()` converts piecewise-constant enrollment and dropout rate lists
into a deterministic set of calendar timepoints — in contrast to the stochastic
`rexp` calls used in [Examples 1](example-1.html) and [2](example-2.html).
`t$get_end_timepoint()` returns the latest timepoint at which all expected
enrollment and follow-up will have completed, providing the calendar time for
the single final analysis.

## Arms (Populations)

For each arm we simulate **two correlated endpoints** per subject using
`MASS::mvrnorm`. We keep **50** subjects per arm (consistent with `sample_size =
100` and equal allocation).

```{r population}
# Create closure to capture parameters
mk_population_generator <- function(mu_y1, mu_y2) {
  function(n) {
    xy <- MASS::mvrnorm(
      n     = n,
      mu    = c(mu_y1, mu_y2),
      Sigma = Sigma
    )
    data.frame(
      id = seq_len(n),
      y1 = xy[, 1],
      y2 = xy[, 2],
      readout_time = 1
    )
  }
}

population_generators <- list(
  pbo = mk_population_generator(mu1_pbo, mu2_pbo),
  trt = mk_population_generator(mu1_trt, mu2_trt)
)
```

`MASS::mvrnorm` draws `n` samples from a multivariate normal with the given mean
vector and covariance matrix `Sigma`. The closure pattern
(`mk_population_generator`) captures the arm-specific means at definition time,
so each call to `population_generators$pbo(n)` or `population_generators$trt(n)`
produces data under the correct arm parameters. Endpoint correlation flows
entirely through the shared `Sigma` matrix — changing `rho` in the scenario
automatically updates both arms.

## Trial Parameters

Define enrollment/dropout profiles and parameters.

```{r trial_params}
arms <- c("pbo", "trt")
allocation <- c(1, 1)
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
```

Note that `enrollment_fn` and `dropout_fn` here are stochastic functions passed
to `replicate_trial()`, which uses them in `gen_plan()` to generate
per-replicate enrollment and dropout times. They are separate from the
piecewise-constant `enrollment` and `dropout` lists used by `gen_timepoints()`
above — those lists drive the deterministic timer, while these functions drive
per-subject event times within each simulated replicate.

## Trigger & Analysis

This is a **fixed design**: perform analysis **once**.  
We run two independent two-sample t-tests (one per endpoint).

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      df_e <- df |> subset(!is.na(enroll_time))

      p1 <- t.test(y1 ~ arm, data = df_e)$p.value
      p2 <- t.test(y2 ~ arm, data = df_e)$p.value
      padj <- p.adjust(c(p1, p2), method = "holm")

      data.frame(
        scenario,
        p_y1      = unname(p1),
        p_y2      = unname(p2),
        p_holm_y1 = unname(padj[1]),
        p_holm_y2 = unname(padj[2]),
        n_total   = nrow(df_e),
        n_pbo     = sum(df_e$arm == "pbo"),
        n_trt     = sum(df_e$arm == "trt"),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

The two t-tests are run independently on the enrolled subset, producing
unadjusted p-values `p_y1` and `p_y2`. `p.adjust(..., method = "holm")` applies
Holm's stepwise procedure: `padj[1]` corresponds to the endpoint with the
smaller raw p-value (most significant) and `padj[2]` to the larger. The adjusted
values account for the familywise error rate across the two endpoints, so
`p_holm_y1 >= p_y1` always holds.

## Trial

Create and run the trial. (We supply a `seed` for reproducibility inside the
trial engine—e.g., enrollment/dropout sampling.)

```{r trial}
trials <- replicate_trial(
  trial_name = "two_endpoints_fixed_design",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r sim}
run_trials(trials)
```

## Results

One row per replicate, row-bound across all replicates:

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`collect_results()` prepends `replicate`, `timepoint`, and `analysis` columns to
the four p-value columns. `p_holm_y1` and `p_holm_y2` will always be ≥ their
unadjusted counterparts because Holm correction is more conservative. Because
the two endpoints are positively correlated (rho = 0.6), replicates that reject
for y1 tend to also reject for y2 — joint rejection probability exceeds what you
would expect for independent endpoints under the same effect sizes.
