---
title: "Example 9: Adaptive Timepoint / Trial Plan Switching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 9: Adaptive Timepoint / Trial Plan Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette demonstrates **adaptive trial plan switching** during a single simulation run.

With the updated `Trial$run()` loop, the trial **re-reads** `timer$timelist` at each step, so if an analysis function modifies the timepoints (the plan) at an interim time, the *new plan is actually followed* for all subsequent timepoints.

We will run and compare two scenarios:

1. **Scenario A (Static plan):** A fixed enrollment plan is used throughout.
2. **Scenario B (Adaptive plan):** At an interim analysis time, we *conditionally* modify the enrollment plan (increase, decrease, or stop enrollment for an arm).

We then compare:
- planned enrollment before vs after switching
- actual enrollment by arm
- interim and final summaries
- the trial's `plan_history` (audit trail of plan changes)

# Setup

```{r setup}
library(rxsim)
library(dplyr)

set.seed(123)
```

# Helper functions

```{r helpers}
# Make a simple continuous outcome population with repeated readouts per subject.
# The Population class requires columns: id, arm (optional), readout_time, and at least one endpoint column.
mk_pop <- function(arm, n, mean_value, sd_value = 5, readouts = c(1, 4)) {
  data <- do.call(rbind, lapply(readouts, function(rt) {
    data.frame(
      id = 1:n,
      readout_time = rt,
      value = rnorm(n, mean = mean_value, sd = sd_value)
    )
  }))
  Population$new(arm, data = data)
}

# Summarise "currently enrolled" snapshot at a given time
snap_summary <- function(dat) {
  dat |>
    group_by(arm) |>
    summarise(
      n_subjects = n_distinct(id),
      mean_value = mean(value),
      .groups = "drop"
    ) |>
    arrange(arm)
}

# Convenience: get current plan (Timer timelist -> data.frame)
get_plan_df <- function(timer) {
  dplyr::bind_rows(timer$timelist) |>
    arrange(time, arm)
}
```

# Trial design: populations + baseline plan

We simulate a simple 2-arm trial with repeated readouts.

- Arm A has mean 50
- Arm B has mean 55 (so B tends to look better at interim)

```{r populations}
# Planned max N is driven by the plan; here we just generate "enough" subjects up front.
popA <- mk_pop("Arm A", n = 120, mean_value = 50)
popB <- mk_pop("Arm B", n = 120, mean_value = 55)

list(popA$n, popB$n)   # subject slots available per arm
popA$n_readouts        # repeated measures per subject
```

## Baseline timepoints (fixed plan)

We plan timepoints at times 1..8.
- From time 1 to 8: enroll 4 per time per arm (total 32 per arm)
- No dropouts for simplicity in this vignette (dropper = 0)

```{r baseline-plan}
times <- 1:8

timepoints_base <- data.frame(
  time = rep(times, times = 2),
  arm = rep(c("Arm A", "Arm B"), each = length(times)),
  enroller = 4L,
  dropper = 0L
)

timepoints_base
timepoints_base |> group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop")
```

# Scenario A: Static plan (no switching)

```{r scenario-a}
# Timer with baseline plan
t_static <- Timer$new("timer_static")
add_timepoints(t_static, timepoints_base)

# Interim + Final analyses (read-only)
interim_time <- 4
final_time <- 8

trigger_by_calendar(interim_time, t_static, analysis = function(df, current_time) {
  dat <- df |> filter(!is.na(enroll_time))
  out <- snap_summary(dat)
  out$time <- current_time
  out
})

trigger_by_calendar(final_time, t_static, analysis = function(df, current_time) {
  dat <- df |> filter(!is.na(enroll_time))
  out <- snap_summary(dat)
  out$time <- current_time
  out
})

# Build and run trial
trial_static <- Trial$new(
  name = "trial_static",
  seed = 123,
  timer = t_static,
  population = list(popA$clone(), popB$clone())
)

trial_static$run()

# Show results as stored in trial
names(trial_static$results)
trial_static$results[[paste0("time_", interim_time)]]
trial_static$results[[paste0("time_", final_time)]]
```

## Scenario A: Actual enrollment by arm

Enrollment lives in each Population object, so we inspect the populations used inside the trial.

```{r scenario-a-enroll}
pops_static <- trial_static$population
enrolled_counts_static <- data.frame(
  arm = sapply(pops_static, function(p) p$name),
  enrolled = sapply(pops_static, function(p) sum(!is.na(p$enrolled)))
)

enrolled_counts_static
```

# Scenario B: Adaptive plan switching at interim

In this scenario, we **switch the timepoints** at `time = 4` depending on interim performance.

**Rule:**
- compute interim mean difference: `mean(B) - mean(A)`
- if difference >= 2: **expand Arm B** (increase enroller to 8 after time 4)
- else: **stop Arm B enrollment** after time 4 (set enroller to 0 after time 4)

Arm A continues as planned.

## Create a switched plan generator

```{r switched-plan}
make_switched_plan <- function(current_plan, interim_time, expand = TRUE) {
  # keep past and interim times the same, but modify future times for Arm B
  current_plan |>
    mutate(
      enroller = case_when(
        time > interim_time & arm == "Arm B" & expand ~ 8L,
        time > interim_time & arm == "Arm B" & !expand ~ 0L,
        TRUE ~ as.integer(enroller)
      ),
      dropper = as.integer(dropper)
    )
}

# Example of what expansion would look like:
timepoints_expandB <- make_switched_plan(timepoints_base, interim_time, expand = TRUE)
timepoints_stopB   <- make_switched_plan(timepoints_base, interim_time, expand = FALSE)

timepoints_expandB |> group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop")
timepoints_stopB |> group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop")
```

## Run with switching inside the interim analysis

Key idea: at interim, we **mutate the timer plan** by replacing `t_switch$timelist`.

```{r scenario-b}
t_switch <- Timer$new("timer_switch")
add_timepoints(t_switch, timepoints_base)

trigger_by_calendar(interim_time, t_switch, analysis = function(df, current_time) {

  dat <- df |> filter(!is.na(enroll_time))
  summ <- snap_summary(dat)

  # Interim decision statistic
  mA <- summ$mean_value[summ$arm == "Arm A"]
  mB <- summ$mean_value[summ$arm == "Arm B"]
  diff <- as.numeric(mB - mA)

  expandB <- is.finite(diff) && diff >= 2

  # Build new plan from current timer plan
  old_plan <- get_plan_df(t_switch)
  new_plan <- make_switched_plan(old_plan, interim_time = current_time, expand = expandB)

  # Switch plan: clear and re-add
  t_switch$timelist <- list()
  add_timepoints(t_switch, new_plan)

  list(
    time = current_time,
    mean_A = mA,
    mean_B = mB,
    diff_B_minus_A = diff,
    decision_expandB = expandB,
    planned_enroll_after_switch = new_plan |>
      group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop")
  )
})

trigger_by_calendar(final_time, t_switch, analysis = function(df, current_time) {
  dat <- df |> filter(!is.na(enroll_time))
  out <- snap_summary(dat)
  out$time <- current_time
  out
})

trial_switch <- Trial$new(
  name = "trial_switch",
  seed = 123,
  timer = t_switch,
  population = list(popA$clone(), popB$clone())
)

trial_switch$run()

# Interim output contains the decision and the plan summary after switch
trial_switch$results[[paste0("time_", interim_time)]]
```

## Scenario B: Audit trail of plan changes

`trial_switch$plan_history` records plan changes detected during the run.

```{r scenario-b-planhistory}
length(trial_switch$plan_history)
trial_switch$plan_history
```

## Scenario B: Actual enrollment by arm

```{r scenario-b-enroll}
pops_switch <- trial_switch$population
enrolled_counts_switch <- data.frame(
  arm = sapply(pops_switch, function(p) p$name),
  enrolled = sapply(pops_switch, function(p) sum(!is.na(p$enrolled)))
)

enrolled_counts_switch
```

# Compare Scenario A vs Scenario B

## Planned enrollment (before vs after switching)

```{r compare-plans}
plan_static <- get_plan_df(t_static) |>
  group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop") |>
  mutate(scenario = "A: Static")

plan_switch_final <- get_plan_df(t_switch) |>
  group_by(arm) |> summarise(planned_enroll = sum(enroller), .groups="drop") |>
  mutate(scenario = "B: After switch")

bind_rows(plan_static, plan_switch_final)
```

## Actual enrollment (end of run)

```{r compare-enroll}
enrolled_counts_static$scenario <- "A: Static"
enrolled_counts_switch$scenario <- "B: Switch"

bind_rows(enrolled_counts_static, enrolled_counts_switch) |>
  arrange(arm, scenario)
```

## Final snapshot summary comparison

```{r compare-final-summary}
final_key <- paste0("time_", final_time)

final_static <- trial_static$results[[final_key]][[1]]
final_switch  <- trial_switch$results[[final_key]][[1]]

final_static$scenario <- "A: Static"
final_switch$scenario <- "B: Switch"

bind_rows(final_static, final_switch) |>
  arrange(arm, scenario)
```

# Notes for extending this pattern

- You can switch from **2-arm to multi-arm** by adding new arms to the plan and ensuring you have corresponding `Population` objects in `trial$population`.
- You can implement **go/no-go** by setting future enrollment for an arm to 0 (as shown).
- You can implement **dose expansion** by increasing enroller counts for selected dose arms.

The core mechanism is always the same:
1. Trigger an interim analysis (`trigger_by_calendar()` or `add_condition()`).
2. Use the interim result to build a *new* plan data.frame.
3. Replace `timer$timelist` with the new plan (`t$timelist <- list(); add_timepoints(t, new_plan)`).
4. The updated `Trial$run()` follows the modified plan for future timepoints.
``
