---
title: "Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Helpers
library(dplyr)

set.seed(6606)
```

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single continuous endpoint** and one **final analysis**.

At final analysis, we report:

- **Overall analysis**: treatment effect across all subjects.
- **Exploratory subgroup analysis**: treatment effect within each subgroup + arm-by-subgroup interaction test.

```{r scenario}
# Trial design
sample_size <- 120
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Subgroup settings
subgroup_levels <- c("A", "B")
subgroup_prob   <- c(0.6, 0.4)

# Data-generating truth (example values)
mu_placebo_A <- 0.00
mu_placebo_B <- 0.00
mu_treat_A   <- 0.40
mu_treat_B   <- 0.15
sigma        <- 1.0

# Enrollment/dropout profile
enrollment <- list(
  end_time = c(4, 8),
  rate     = c(8, 7)
)
dropout <- list(
  end_time = c(4, 8),
  rate     = c(0, 0)
)
```

## Time points

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

tr_timer <- Timer$new(name = "timer_example_6")
add_timepoints(tr_timer, timepoints)

final_time <- tr_timer$get_end_timepoint()
final_time
```

## Populations

Create two populations and include a `subgroup` column in each population dataset.

```{r populations}
n_per_arm <- sample_size / 2

mk_arm_data <- function(n, mean_A, mean_B) {
  subgroup <- sample(subgroup_levels, size = n, replace = TRUE, prob = subgroup_prob)
  mu <- ifelse(subgroup == "A", mean_A, mean_B)

  data.frame(
    subject_id = seq_len(n),
    subgroup = subgroup,
    y = rnorm(n, mean = mu, sd = sigma)
  )
}

pop_placebo <- Population$new(
  name = "placebo",
  data = mk_arm_data(n_per_arm, mu_placebo_A, mu_placebo_B)
)

pop_treatment <- Population$new(
  name = "treatment",
  data = mk_arm_data(n_per_arm, mu_treat_A, mu_treat_B)
)
```

## Final analysis trigger

```{r analysis}
tr_timer$add_condition(
  time == final_time,
  analysis = function(df, time) {
    dat <- df |>
      dplyr::filter(!is.na(enroll_time)) |>
      dplyr::mutate(
        arm = factor(arm, levels = c("placebo", "treatment")),
        subgroup = factor(subgroup, levels = subgroup_levels)
      )

    # Overall analysis
    fit_overall <- lm(y ~ arm, data = dat)
    coef_overall <- summary(fit_overall)$coefficients

    overall_est <- unname(coef_overall["armtreatment", "Estimate"])
    overall_se  <- unname(coef_overall["armtreatment", "Std. Error"])
    overall_p   <- unname(coef_overall["armtreatment", "Pr(>|t|)"])

    # Exploratory subgroup analyses (within each subgroup)
    subgroup_results <- lapply(levels(dat$subgroup), function(sg) {
      dsg <- dat[dat$subgroup == sg, , drop = FALSE]
      fit_sg <- lm(y ~ arm, data = dsg)
      coef_sg <- summary(fit_sg)$coefficients

      data.frame(
        subgroup = sg,
        n_total = nrow(dsg),
        n_placebo = sum(dsg$arm == "placebo"),
        n_treatment = sum(dsg$arm == "treatment"),
        estimate = unname(coef_sg["armtreatment", "Estimate"]),
        std_error = unname(coef_sg["armtreatment", "Std. Error"]),
        p_value = unname(coef_sg["armtreatment", "Pr(>|t|)"]),
        stringsAsFactors = FALSE
      )
    })

    subgroup_results <- do.call(rbind, subgroup_results)

    # Exploratory interaction test
    fit_interaction <- lm(y ~ arm * subgroup, data = dat)
    coef_int <- summary(fit_interaction)$coefficients
    int_term <- grep("^arm.*:subgroup", rownames(coef_int), value = TRUE)
    p_interaction <- if (length(int_term) > 0) {
      unname(coef_int[int_term[1], "Pr(>|t|)"])
    } else {
      NA_real_
    }

    list(
      n_total = nrow(dat),
      overall_estimate = overall_est,
      overall_std_error = overall_se,
      overall_p_value = overall_p,
      subgroup_results = subgroup_results,
      interaction_p_value = p_interaction,
      note = "Subgroup and interaction analyses are exploratory and not multiplicity-adjusted."
    )
  },
  name = "final"
)
```

## Trial

```{r trial}
trial <- Trial$new(
  name = "example_6_two_arm_fixed_subgroup",
  seed = 6606,
  timer = tr_timer,
  population = list(pop_placebo, pop_treatment)
)
```

## Simulate

```{r simulate}
trial$run()
```

## Results

```{r results}
trial$results
```

```{r results-pretty}
prettify_results(trial$results)
```
