---
title: "Example 5: Multi-arm | Fixed design | Single continuous endpoint | Dunnett test + MCP-Mod"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 5: Multi-arm | Fixed design | Single continuous endpoint | Dunnett test + MCP-Mod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework (Timer, Population, Trial, gen_timepoints, add_timepoints, ...)
library(rxsim)

# Analyses
library(multcomp)     # Dunnett
library(DoseFinding)  # MCP-Mod

# Helpers
library(dplyr)
set.seed(4566)
```

## Scenario

A **multi-arm, fixed design** (placebo + 4 active doses) with a **single continuous endpoint**.  
At the **final analysis**, we perform:

- **Dunnett test**: all active doses vs placebo using a normal-theory linear model.
- **MCP-Mod**: model-based multiple contrast test + model fitting on a candidate set of dose–response shapes.

```{r scenario}
# Dose levels (placebo + 4 actives)
doses <- c(0, 5, 10, 20, 50)

# Total N and allocation (balanced across arms)
sample_size <- 150
allocation  <- rep(1, length(doses))
arm_names   <- paste0("d", doses)  # e.g., d0, d5, ... used as arm labels

# Enrollment & dropout profile (piece-wise constant)
enrollment <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(6, 12, 18, 24)
)
dropout <- list(
  end_time = c(4, 8, 12, 16),
  rate     = c(0,  1,  2,  3)
)

# Data-generating model: Emax with homoscedastic noise
e0    <- 0.0
emax  <- 1.0
ed50  <- 20.0
sigma <- 1.0

mean_fun <- function(d) e0 + emax * d / (ed50 + d)

# operating characteristics
alpha <- 0.05
delta <- 0.1
```

## Time points

Create a `Timer` with discrete time points to reach the target sample size.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arm_names,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

# Timer
tr_timer <- Timer$new(name = "timer_multiarm")
add_timepoints(tr_timer, timepoints)
final_time <- tr_timer$get_end_timepoint()
final_time
```

## Populations

Simulate subject-level responses for each arm according to the Emax model.

```{r population}
n_per_arm <- sample_size / length(doses)

mk_pop <- function(d){
  mu <- mean_fun(d)
  y  <- rnorm(n_per_arm, mean = mu, sd = sigma)
  data.frame(
    subject_id = seq_len(n_per_arm),
    dose       = d,
    y          = y
  )
}

pops <- lapply(seq_along(doses), function(i){
  nm <- paste0("d", doses[i])
  Population$new(name = nm, data = mk_pop(doses[i]))
})
```

## Triggers & Analyses

At the **final time**, run:

1) **Dunnett test** (active vs placebo) using `multcomp::glht` on `lm(y ~ arm)`.

2) **MCP-Mod** using `DoseFinding::MCPMod` with a candidate model set (`Mods`).

```{r trigger}
# Candidate model set for MCP-Mod (tweak to your compound)
# We pass shape parameters on the original dose scale via 'doses = doses'.
mods <- Mods(
  linear     = NULL,
  emax       = 20,           # ED50 ~ 20
  exponential= 50,           # delta ~ 50
  sigEmax    = c(20, 3),     # ED50 ~ 20, Hill ~ 3
  quadratic  = -0.2,         # standardized curvature; see help(drmodels)
  doses      = doses
)

tr_timer$add_condition(
  time == final_time,
  analysis = function(df, time){
    df_e <- df |>
      dplyr::filter(!is.na(enroll_time)) |>
      dplyr::mutate(
        arm  = factor(arm, levels = paste0("d", doses)),
        dose = as.numeric(dose)
      )

    # 1) Dunnett (active vs placebo)
    fit <- lm(y ~ arm, data = df_e)
    dun  <- multcomp::glht(fit, linfct = multcomp::mcp(arm = "Dunnett"))
    summ <- summary(dun)

    # 2) MCP-Mod (one-step): model-based multiple contrast tests + model fit
    mm <- DoseFinding::MCPMod(
        dose   = df_e$dose,
        resp   = df_e$y,
        models = mods,
        type   = "normal",
        Delta  = delta,
        alpha  = alpha,
        selModel = "aveAIC"
    )

    mct_min_p <- min(attr(mm$MCTtest$tStat, "pVal"), na.rm = TRUE)

    list(
      n_total      = nrow(df_e),
      dunn_min_p   = summ$test$pvalues |> as.numeric() |> min(),
      mcpmod_min_p = mct_min_p
    )
  },
  name = "final"
)
```

## Trial

```{r trial}
trial <- Trial$new(
  name       = "multiarm_dunnett_mcpmod",
  seed       = 456,
  timer      = tr_timer,
  population = pops
)
```

## Simulate

```{r sim}
trial$run()
```

## Results

The full result list (Dunnett table + MCP-Mod summary elements):

```{r results}
trial$results
```

Flatten for quick inspection:

```{r results_pretty}
prettify_results(trial$results)
```
