---
title: "Example 8: Seamless Ph2a/2b | Two-arm -> Multi-arm | Single continuous endpoint | Bayesian Go/No-Go + Bayesian MCP-Mod"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 8: Seamless Ph2a/2b | Two-arm -> Multi-arm | Single continuous endpoint | Bayesian Go/No-Go + Bayesian MCP-Mod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Bayesian utilities
library(RBesT)
library(BayesianMCPMod)
library(DoseFinding)

# Helpers
library(dplyr)

set.seed(8808)
```

## Scenario

A **seamless Ph2a/2b** trial with one continuous endpoint:

- **Ph2a** (early period): 2 arms (placebo + one active dose).
- **Ph2b** (later period): expansion to multi-arm dose finding.
- **Interim**: Bayesian Go/No-Go based on treatment effect for the Ph2a dose.
- **Final**: Bayesian MCP-Mod across all doses.

```{r scenario}
# Dose levels and arm labels
dose_map <- c(d0 = 0, d5 = 5, d10 = 10, d20 = 20, d40 = 40)
arm_names <- names(dose_map)
dose_levels <- as.numeric(dose_map)

# Truth for data generation (Emax shape)
e0    <- 0.0
emax  <- 0.9
ed50  <- 15
sigma <- 1.0
mean_fun <- function(d) e0 + emax * d / (ed50 + d)

# Interim Go/No-Go thresholds
delta_interim <- 0.1
gamma_interim <- 0.8

# Multiplicity level for final Bayesian MCP step
alpha_final <- 0.05

# Planned analysis times
interim_time <- 4
final_time   <- 12
```

## Seamless enrollment schedule (2-arm -> multi-arm)

In this implementation, all arms are pre-defined, but enrollment switches from two-arm to multi-arm by time.

```{r timepoints}
time <- 1:12

# Ph2a: only placebo + d20 active
# Ph2b: expand to d5, d10, d40 while placebo/d20 continue per plan
enroll_plan <- data.frame(
  time = rep(time, times = length(arm_names)),
  arm  = rep(arm_names, each = length(time)),
  enroller = c(
    # d0 placebo (total 80)
    8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 0, 0,
    # d5 (total 20)
    0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0,
    # d10 (total 20)
    0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0,
    # d20 (total 40)
    8, 8, 8, 8, 2, 2, 2, 2, 0, 0, 0, 0,
    # d40 (total 20)
    0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0
  ),
  dropper = 0L
)

enroll_plan$enroller <- as.integer(enroll_plan$enroller)
enroll_plan$dropper  <- as.integer(enroll_plan$dropper)

tr_timer <- Timer$new(name = "timer_example_8")
add_timepoints(tr_timer, enroll_plan)

tr_timer$get_end_timepoint()
```

## Populations

```{r populations}
# Planned max N by arm from enrollment schedule
n_by_arm <- enroll_plan |>
  group_by(arm) |>
  summarise(n = sum(enroller), .groups = "drop")

mk_pop <- function(arm, n) {
  d <- dose_map[[arm]]
  Population$new(
    name = arm,
    data = data.frame(
      subject_id = seq_len(n),
      dose = d,
      y = rnorm(n, mean = mean_fun(d), sd = sigma)
    )
  )
}

pops <- lapply(seq_len(nrow(n_by_arm)), function(i) {
  mk_pop(n_by_arm$arm[i], n_by_arm$n[i])
})
```

## Priors

For interim Go/No-Go, we use placebo borrowing from small historical placebo data and a non-informative prior for active.

For final Bayesian MCP-Mod, we pass arm-wise priors to `BayesianMCPMod`.

```{r priors}
# Historical placebo summary data
hist_placebo <- data.frame(
  study = c("H1", "H2", "H3"),
  n     = c(20, 24, 18),
  mean  = c(-0.06, 0.00, 0.03)
)

n_hist <- sum(hist_placebo$n)
m_hist <- weighted.mean(hist_placebo$mean, hist_placebo$n)

# Base non-informative prior
prior_noninf <- RBesT::mixnorm(c(1, 0, 1e-6), sigma = sigma, param = "mn")

# Placebo borrowing prior for interim and final control arm
prior_placebo_inf <- RBesT::postmix(prior_noninf, n = n_hist, m = m_hist)
prior_placebo <- RBesT::robustify(
  prior_placebo_inf,
  weight = 0.2,
  mean = 0,
  n = 1,
  sigma = sigma
)

# Prior list for Bayesian MCP-Mod (Ctrl + 4 active groups)
prior_list_final <- list(
  Ctrl = prior_placebo,
  DG_1 = prior_noninf,
  DG_2 = prior_noninf,
  DG_3 = prior_noninf,
  DG_4 = prior_noninf
)
```

## Analyses

We add two trigger conditions:

1. `time == interim_time`: Bayesian Go/No-Go on d20 vs placebo.
2. `time == final_time`: Bayesian MCP-Mod over all dose groups.

```{r analyses}
# shared state between interim and final trigger
go_interim <- NA_integer_

# Candidate dose-response models for Bayesian MCP-Mod
mods <- DoseFinding::Mods(
  linear      = NULL,
  emax        = 20,
  exponential = 20,
  sigEmax     = c(20, 3),
  doses       = dose_levels
)

# Interim: Bayesian Go/No-Go (d20 vs placebo)
tr_timer$add_condition(
  time == interim_time,
  analysis = function(df, time) {
    dat <- df |>
      dplyr::filter(!is.na(enroll_time), arm %in% c("d0", "d20"))

    y_p <- dat$y[dat$arm == "d0"]
    y_t <- dat$y[dat$arm == "d20"]

    n_p <- length(y_p)
    n_t <- length(y_t)
    m_p <- mean(y_p)
    m_t <- mean(y_t)

    post_p <- RBesT::postmix(prior_placebo, n = n_p, m = m_p)
    post_t <- RBesT::postmix(prior_noninf,  n = n_t, m = m_t)

    prob_delta <- RBesT::pmixdiff(post_t, post_p, delta_interim, lower.tail = FALSE)
    delta_q <- RBesT::qmixdiff(post_t, post_p, c(0.025, 0.5, 0.975))

    go_interim <<- as.integer(prob_delta >= gamma_interim)

    list(
      n_placebo = n_p,
      n_d20 = n_t,
      post_prob_delta = prob_delta,
      delta_q025 = delta_q[1],
      delta_q500 = delta_q[2],
      delta_q975 = delta_q[3],
      decision_go = go_interim
    )
  },
  name = "interim_go_nogo"
)

# Final: Bayesian MCP-Mod across all dose groups
tr_timer$add_condition(
  time == final_time,
  analysis = function(df, time) {
    dat <- df |>
      dplyr::filter(!is.na(enroll_time))

    # Arm-wise summary at final
    summ <- dat |>
      dplyr::group_by(arm) |>
      dplyr::summarise(
        dose = unique(dose)[1],
        n = dplyr::n(),
        mu_hat = mean(y),
        .groups = "drop"
      ) |>
      dplyr::arrange(dose)

    # Order expected by BayesianMCPMod: Ctrl, DG_1 ...
    # with dose order 0, 5, 10, 20, 40.
    n_vec <- summ$n
    mu_vec <- summ$mu_hat

    # covariance of mean estimates (independent groups here)
    S_hat <- diag((sigma / sqrt(n_vec))^2)

    contr <- BayesianMCPMod::getContr(
      mods = mods,
      dose_levels = dose_levels,
      cov_posterior = S_hat
    )

    crit_prob <- BayesianMCPMod::getCritProb(
      mods = mods,
      dose_levels = dose_levels,
      dose_weights = n_vec |> as.numeric(),
      alpha_crit_val = alpha_final
    )

    post_list <- BayesianMCPMod::getPosterior(
      prior_list = prior_list_final,
      mu_hat = mu_vec,
      S_hat = S_hat,
      calc_ess = TRUE
    )

    bmcp <- BayesianMCPMod::performBayesianMCPMod(
      posterior_list = post_list,
      contr = contr,
      crit_prob_adj = crit_prob,
      simple = TRUE,
      avg_fit = TRUE
    )

    list(
      interim_go = go_interim,
      n_total_final = nrow(dat),
      dose_means = summ,
      crit_prob_adj = crit_prob,
      bayesian_mcpmod = bmcp
    )
  },
  name = "final_bayesian_mcpmod"
)
```

## Trial

```{r trial}
trial <- Trial$new(
  name = "example_8_seamless_ph2a2b",
  seed = 8808,
  timer = tr_timer,
  population = pops
)
```

## Simulate

```{r simulate}
trial$run()
```

## Results

Interim and final analyses are stored under the corresponding time keys.

```{r results}
trial$results
```

```{r inspect-final}
final_key <- paste0("time_", final_time)

# top-level components returned by final analysis
names(trial$results[[final_key]]$final_bayesian_mcpmod)

# Bayesian MCP-Mod object class
class(trial$results[[final_key]]$final_bayesian_mcpmod$bayesian_mcpmod)
```

This example demonstrates the seamless setup and analysis flow in one simulation run:
2-arm Ph2a enrollment, interim Bayesian Go/No-Go check, then final Bayesian MCP-Mod after multi-arm expansion.
