---
title: "Example 2: Two arm | Interim | Continuous | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2: Two arm | Interim | Continuous | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(rxsim)
```

This example extends [Example 1](example-1.html) by introducing two named
analyses — `interim` and `final` — that fire at different enrollment
milestones. Understanding how analysis names propagate through
`collect_results()` is the foundation for multi-analysis designs shown in
later examples.

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms        <- c("pbo", "trt")
allocation  <- c(1, 1)
delta       <- 0.2  # true treatment - placebo mean difference
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn    <- function(n) rexp(n, rate = 0.01)
scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation  = list(allocation),
  delta       = delta
)
```

`allocation = c(1, 1)` specifies balanced randomisation. `tidyr::expand_grid()`
embeds design parameters directly into each result row for traceability across
parameter sweeps.

## Populations
Define population generators.

```{r population}
population_generators <- list(
  pbo = function(n) data.frame(
    id = 1:n,
    value = rnorm(n),
    readout_time = 1
  ),
  trt = function(n) data.frame(
    id = 1:n,
    value = rnorm(n, mean = delta),
    readout_time = 1
  )
)
```

Each generator is a function of `n` returning a `data.frame` with one row per
subject. `readout_time = 1` means the endpoint is observed 1 time unit after
enrollment. The treatment arm has a mean shift of δ = 0.2 SD units over placebo.

## Triggers & Analysis
Final analysis at full enrollment.

```{r trigger}
analysis_generators <- list(
  interim = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= (!!sample_size) / 2
    ),
    analysis = function(df, timer){
      df_enrolled <- df |> subset(!is.na(enroll_time))
      tt <- t.test(value ~ arm, data = df_enrolled)
      data.frame(
        scenario,
        n_total = nrow(df_enrolled),
        mean_pbo = mean(df_enrolled$value[df_enrolled$arm == "pbo"]),
        mean_trt = mean(df_enrolled$value[df_enrolled$arm == "trt"]),
        p_value = unname(tt$p.value),
        stringsAsFactors = FALSE
      )
    }
  ),
  
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer){
      df_enrolled <- df |> subset(!is.na(enroll_time))
      tt <- t.test(value ~ arm, data = df_enrolled)
      data.frame(
        scenario,
        n_total = nrow(df_enrolled),
        mean_pbo = mean(df_enrolled$value[df_enrolled$arm == "pbo"]),
        mean_trt = mean(df_enrolled$value[df_enrolled$arm == "trt"]),
        p_value = unname(tt$p.value),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

Two analyses are defined. The `interim` analysis fires at half enrollment and
the `final` analysis fires when all subjects are enrolled. These names appear
in the `analysis` column of `collect_results()` output, making it
straightforward to distinguish interim from final results when stacking rows
across timepoints.

## Trial
Make multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "test_trial",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate
To simulate all replicates:

```{r sim}
run_trials(trials)
```

Bind one row per replicate into one data frame.
```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`collect_results()` prepends `replicate`, `timepoint`, and `analysis` columns
to your result data. Each row shows either `"interim"` or `"final"` in the
`analysis` column, reflecting which named analysis produced it. This naming
convention becomes essential in later examples where multiple named analyses
fire per replicate.
