---
title: "Example 2: Two arm | Interim + Final | Continuous | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 2: Two arm | Interim + Final | Continuous | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache   = TRUE
)
```

```{r setup}
library(rxsim)
```

Building on [Example 1](example-1.html), this example adds an interim analysis
at 50% enrollment alongside the final analysis at 100%. Two named `Condition`
objects fire per replicate; the `analysis` column in `collect_results()` output
distinguishes their rows.

**Unique focus:** two named analyses (`interim` / `final`), filtering results
by analysis name, comparing nominal power at interim vs final.

## Scenario

```{r scenario}
sample_size   <- 100
arms          <- c("pbo", "trt")
allocation    <- c(1, 1)
delta         <- 0.2
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn    <- function(n) rexp(n, rate = 0.01)

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation  = list(allocation),
  delta       = delta
)
```

## Populations

```{r population}
population_generators <- list(
  pbo = function(n) data.frame(id = 1:n, value = rnorm(n),        readout_time = 1),
  trt = function(n) data.frame(id = 1:n, value = rnorm(n, delta),  readout_time = 1)
)
```

## Triggers and analyses

Two entries in `analysis_generators` create two independent `Condition` objects
per replicate. `interim` fires at 50 enrolled subjects, `final` at 100. Both
run the same t-test; the `analysis` column tags each row accordingly.

```{r trigger}
analysis_generators <- list(
  interim = list(
    trigger  = enroll_trigger(0.5, sample_size),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      tt <- t.test(value ~ arm, data = enrolled)
      data.frame(scenario, n_total  = nrow(enrolled),
                 mean_pbo = mean(enrolled$value[enrolled$arm == "pbo"]),
                 mean_trt = mean(enrolled$value[enrolled$arm == "trt"]),
                 p_value  = unname(tt$p.value))
    }
  ),
  final = list(
    trigger  = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      tt <- t.test(value ~ arm, data = enrolled)
      data.frame(scenario, n_total  = nrow(enrolled),
                 mean_pbo = mean(enrolled$value[enrolled$arm == "pbo"]),
                 mean_trt = mean(enrolled$value[enrolled$arm == "trt"]),
                 p_value  = unname(tt$p.value))
    }
  )
)
```

## Simulate

```{r trial}
set.seed(2)
trials <- replicate_trial(
  trial_name            = "example_2",
  sample_size           = sample_size,
  arms                  = arms,
  allocation            = allocation,
  enrollment            = enrollment_fn,
  dropout               = dropout_fn,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 6
)
```

```{r sim, results=FALSE}
run_trials(trials)
```

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

Each replicate contributes two rows: one for `"interim"` and one for `"final"`.
Filter by analysis name using the `analysis` argument:

```{r filter_analysis}
interim_res <- collect_results(trials, analysis = "interim")
final_res   <- collect_results(trials, analysis = "final")

cat("Mean n at interim:", mean(interim_res$n_total), "\n")
cat("Mean n at final:  ", mean(final_res$n_total),   "\n")
```

## Interim vs final power

500 replicates quantify the power gain from waiting for full enrollment.

```{r power_comparison, cache=TRUE}
set.seed(42)
n_reps_pw <- 300

tr_pw <- replicate_trial(
  trial_name            = "ex2_pw",
  sample_size           = sample_size, arms = arms, allocation = allocation,
  enrollment            = enrollment_fn, dropout = dropout_fn,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = n_reps_pw
)
invisible(run_trials(tr_pw))

pw_interim <- mean(collect_results(tr_pw, analysis = "interim")$p_value < 0.05)
pw_final   <- mean(collect_results(tr_pw, analysis = "final")$p_value   < 0.05)

cat(sprintf("Power at interim (n = 50):  %.3f\n", pw_interim))
cat(sprintf("Power at final   (n = 100): %.3f\n", pw_final))
```

```{r plot_comparison}
barplot(
  c(interim = pw_interim, final = pw_final),
  ylim = c(0, 1), col  = c("steelblue", "tomato"),
  ylab = "Empirical power (alpha = 0.05)",
  main = sprintf("Power: interim vs final, delta = %.1f, n = %d", delta, sample_size)
)
abline(h = 0.80, lty = 2, col = "grey40")
```

## Next steps

- [Interim Analyses](interims.html) - combining triggers, cooldown, alpha
  spending
- [Example 3](example-3.html) - two correlated continuous endpoints
