---
title: "Example Multi-Run: Scenario grid sweep with tidyr::expand_grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Multi-Run: Scenario grid sweep with tidyr::expand_grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache   = TRUE
)
```

```{r setup}
library(rxsim)
library(tidyr)
library(ggplot2)
```

A single rxsim call produces one set of replicates under fixed parameters.
To explore how power or bias changes across a *grid* of scenarios - for
example, varying sample size and effect size simultaneously - you nest
`replicate_trial()` inside a `lapply()` over the scenario rows, then bind
the results. `tidyr::expand_grid()` creates the full factorial combination,
and the `scenario` data frame passed inside each analysis function tags every
result row with the parameters that generated it.

**Unique focus:** parameter sweeps via `tidyr::expand_grid`, running multiple
independent scenario configurations, binding and plotting results across the
grid.

## Scenario grid

We sweep over two sample sizes (60 and 100) and three true effect sizes
(0.0, 0.3, 0.5), giving 6 scenario combinations, each run for 200 replicates.

```{r scenario}
set.seed(123)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn    <- function(n) rexp(n, rate = 0.01)

scenario_grid <- tidyr::expand_grid(
  sample_size = c(60, 100),
  delta       = c(0.0, 0.3, 0.5)
)

scenario_grid
```

## Analysis function

For each scenario we run a two-sample t-test at full enrollment. The
`scenario` data frame (created inside the `lapply`) carries the current
`sample_size` and `delta` into every result row automatically.

```{r analysis_fn}
make_generators <- function(ss, d) {
  sc <- data.frame(sample_size = ss, delta = d)

  population_generators <- list(
    pbo = function(n) {
      data.frame(id = 1:n, value = rnorm(n),    readout_time = 1)
    },
    trt = local({
      delta_i <- d
      function(n) {
        data.frame(id = 1:n, value = rnorm(n, delta_i), readout_time = 1)
      }
    })
  )

  analysis_generators <- list(
    final = list(
      trigger  = enroll_trigger(1.0, ss),
      analysis = function(df, current_time) {
        enrolled <- subset(df, !is.na(enroll_time))
        tt <- t.test(value ~ arm, data = enrolled)
        data.frame(
          sc,
          n_total  = nrow(enrolled),
          p_value  = unname(tt$p.value),
          estimate = unname(diff(tt$estimate))
        )
      }
    )
  )

  list(
    population_generators = population_generators,
    analysis_generators   = analysis_generators
  )
}
```

## Run across grid

```{r run_grid, results=FALSE}
n_reps <- 200

all_trials <- lapply(seq_len(nrow(scenario_grid)), function(i) {
  ss <- scenario_grid$sample_size[i]
  d  <- scenario_grid$delta[i]
  g  <- make_generators(ss, d)
  tr <- replicate_trial(
    trial_name            = paste0("grid_n", ss, "_d", d),
    sample_size           = ss,
    arms                  = arms,
    allocation            = allocation,
    enrollment            = enrollment_fn,
    dropout               = dropout_fn,
    analysis_generators   = g$analysis_generators,
    population_generators = g$population_generators,
    n                     = n_reps
  )
  run_trials(tr)
  tr
})
```

## Collect and summarise

`collect_results()` works on a single trial object. We apply it to each
element of `all_trials` and row-bind the outputs.

```{r collect}
results_df <- do.call(rbind, lapply(all_trials, collect_results))
head(results_df)
```

```{r summarise}
power_summary <- aggregate(
  p_value ~ sample_size + delta,
  data = results_df,
  FUN  = function(p) mean(p < 0.05)
)
names(power_summary)[3] <- "power"
power_summary
```

## Power heatmap

```{r plot_grid}
ggplot(power_summary,
       aes(x = factor(delta), y = factor(sample_size), fill = power)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", power)), size = 5) +
  scale_fill_gradient2(
    low = "white", mid = "steelblue", high = "navy",
    midpoint = 0.5, limits = c(0, 1),
    name = "Power"
  ) +
  labs(
    x     = "True effect delta (SD units)",
    y     = "Sample size",
    title = "Empirical power across scenario grid (alpha = 0.05)"
  ) +
  theme_minimal(base_size = 13)
```

## Next steps

- [API Styles](api-styles.html) - direct vs generator API: when to use each
- [Interim Analyses](interims.html) - extend this grid to compare fixed vs
  adaptive designs
- [Getting Started](getting-started.html) - single-scenario direct API
  walkthrough
