---
title: "Example Switching: Seamless Ph2a/2b with Conditional Plan Switching (Adaptive Timepoints)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Switching: Seamless Ph2a/2b with Conditional Plan Switching (Adaptive Timepoints)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Bayesian utilities
library(RBesT)
library(BayesianMCPMod)
library(DoseFinding)

# Helpers
library(dplyr)

set.seed(8808)
```

# Motivation

This vignette expands **Example 8** by demonstrating the **new adaptive plan/timepoint switching** mechanism.

With the updated `Trial$run()` loop in `rxsim`, the trial **re-reads the Timer plan** (`timer$timelist`) at each step. Therefore, if an interim analysis modifies the timepoints (e.g., expands from 2 arms to multiple arms, or stops enrollment), the trial follows the updated plan in the same run.

We compare:

1. **Static seamless plan** (as in Example 8): plan is fully specified upfront.
2. **Adaptive plan switching**: start with **2-arm Ph2a** only, then at interim:
   - **GO** → switch plan to expand to multi-arm Ph2b and add final Bayesian MCP-Mod condition.
   - **NO-GO** → switch plan to stop future enrollment and skip the final MCP-Mod.

We show both branches by running:
- a **GO scenario** (true effect exists)
- a **NO-GO scenario** (null effect)

---

# Scenario setup

```{r scenario}
# Dose levels and arm labels
dose_map <- c(d0 = 0, d5 = 5, d10 = 10, d20 = 20, d40 = 40)
arm_names <- names(dose_map)
dose_levels <- as.numeric(dose_map)

# Noise
sigma <- 1.0

# Interim Go/No-Go thresholds 
delta_interim <- 0.1
gamma_interim <- 0.8

# Multiplicity level for final Bayesian MCP step
alpha_final <- 0.05

# Planned analysis times
interim_time <- 4
final_time   <- 12
```

---

# Helper: build static vs adaptive plans

We define:

- A **2-arm Ph2a-only plan** (placebo + d20) for times 1..4.
- A **multi-arm Ph2b expansion plan** for times 5..12.
- A **stop plan** that sets all enrollers to 0 after interim.

```{r plan-helpers}
time <- 1:12

make_plan_static <- function() {
  enroll_plan <- data.frame(
    time = rep(time, times = length(arm_names)),
    arm  = rep(arm_names, each = length(time)),
    enroller = c(
      # d0 placebo (total 80)
      8, 8, 8, 8, 8, 8, 8, 8, 8, 8, 0, 0,
      # d5 (total 20)
      0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0,
      # d10 (total 20)
      0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0,
      # d20 (total 40)
      8, 8, 8, 8, 2, 2, 2, 2, 0, 0, 0, 0,
      # d40 (total 20)
      0, 0, 0, 0, 3, 3, 4, 3, 3, 4, 0, 0
    ),
    dropper = 0L
  )
  enroll_plan$enroller <- as.integer(enroll_plan$enroller)
  enroll_plan$dropper  <- as.integer(enroll_plan$dropper)
  enroll_plan
}

make_plan_ph2a_only <- function() {
  # Only placebo + d20 in times 1..12
  enroll_plan <- data.frame(
    time = rep(time, times = length(arm_names)),
    arm  = rep(arm_names, each = length(time)),
    enroller = 0L,
    dropper = 0L
  )

  # d0 and d20 enroll during Ph2a (times 1..4)
  enroll_plan$enroller[enroll_plan$arm == "d0"  & enroll_plan$time %in% 1:4] <- 8L
  enroll_plan$enroller[enroll_plan$arm == "d20" & enroll_plan$time %in% 1:4] <- 8L

  enroll_plan$enroller <- as.integer(enroll_plan$enroller)
  enroll_plan$dropper  <- as.integer(enroll_plan$dropper)
  enroll_plan
}

make_plan_ph2b_expand <- function() {
  plan <- make_plan_ph2a_only()

  # Continue placebo in Ph2b (times 5..10)
  plan$enroller[plan$arm == "d0" & plan$time %in% 5:10] <- 8L

  # Add new dose arms d5,d10,d40 in Ph2b (times 5..10)
  for (a in c("d5", "d10", "d40")) {
    plan$enroller[plan$arm == a & plan$time %in% c(5,6,7,8,9,10)] <- c(3L,3L,4L,3L,3L,4L)
  }

  # Continue d20 a bit in early Ph2b (times 5..8)
  plan$enroller[plan$arm == "d20" & plan$time %in% 5:8] <- 2L

  plan$enroller <- as.integer(plan$enroller)
  plan
}

make_plan_stop_after_interim <- function(plan, interim_time) {
  # After interim_time, stop all enrollment
  plan$enroller[plan$time > interim_time] <- 0L
  plan$enroller <- as.integer(plan$enroller)
  plan
}

summarise_plan <- function(df) {
  df |>
    group_by(arm) |>
    summarise(planned_n = sum(enroller), .groups = "drop") |>
    arrange(match(arm, arm_names))
}
```

---

# Priors

```{r priors}
# Historical placebo summary data
hist_placebo <- data.frame(
  study = c("H1", "H2", "H3"),
  n     = c(20, 24, 18),
  mean  = c(-0.06, 0.00, 0.03)
)

n_hist <- sum(hist_placebo$n)
m_hist <- weighted.mean(hist_placebo$mean, hist_placebo$n)

# Base non-informative prior
prior_noninf <- RBesT::mixnorm(c(1, 0, 1e-6), sigma = sigma, param = "mn")

# Placebo borrowing prior
prior_placebo_inf <- RBesT::postmix(prior_noninf, n = n_hist, m = m_hist)
prior_placebo <- RBesT::robustify(
  prior_placebo_inf,
  weight = 0.2,
  mean = 0,
  n = 1,
  sigma = sigma
)

# Prior list for Bayesian MCP-Mod (Ctrl + 4 active groups)
prior_list_final <- list(
  Ctrl = prior_placebo,
  DG_1 = prior_noninf,
  DG_2 = prior_noninf,
  DG_3 = prior_noninf,
  DG_4 = prior_noninf
)
```

---

# Models for MCP-Mod 

```{r mcpmod-models}
mods <- DoseFinding::Mods(
  linear      = NULL,
  emax        = 20,
  exponential = 20,
  sigEmax     = c(20, 3),
  doses       = dose_levels
)
```

---

# Helper: build Populations under a chosen truth

We create all arm populations upfront (so switching plans only changes enrollment, not which populations exist).

```{r populations-helper}
mk_pops <- function(e0, emax, ed50, sigma) {
  mean_fun <- function(d) e0 + emax * d / (ed50 + d)

  # Use plan maxima for "available subject slots"
  max_n <- summarise_plan(make_plan_static())$planned_n
  names(max_n) <- summarise_plan(make_plan_static())$arm

  lapply(arm_names, function(arm) {
    d <- dose_map[[arm]]
    n <- max_n[[arm]]
    Population$new(
      name = arm,
      data = data.frame(
        id = seq_len(n),
        dose = d,
        y = rnorm(n, mean = mean_fun(d), sd = sigma),
        readout_time = 1
      )
    )
  })
}
```

---

# Core function: run a trial (static or adaptive)

This function runs either:
- `mode = "static"`: plan is fixed upfront (Example 8 behavior)
- `mode = "adaptive"`: start Ph2a-only plan, then switch at interim based on Go/No-Go

It returns a list with:
- trial object
- planned enrollment (original and final plan)
- observed enrollment by arm
- interim decision

```{r run-trial-fn}
get_plan_df <- function(timer) {
  dplyr::bind_rows(timer$timelist) |>
    arrange(time, arm)
}

count_enrolled_by_arm <- function(trial) {
  data.frame(
    arm = vapply(trial$population, function(p) p$name, character(1)),
    enrolled = vapply(trial$population, function(p) sum(!is.na(p$enrolled)), integer(1))
  ) |>
    arrange(match(arm, arm_names))
}

run_trial <- function(mode = c("static","adaptive"),
                      truth = c("go","nogo"),
                      seed = 8808) {

  mode <- match.arg(mode)
  truth <- match.arg(truth)

  # Truth parameters
  if (truth == "go") {
    e0 <- 0.0; emax <- 0.9; ed50 <- 15
  } else {
    e0 <- 0.0; emax <- 0.0; ed50 <- 15
  }

  set.seed(seed)
  pops <- mk_pops(e0, emax, ed50, sigma)

  # Timer + initial plan
  tr_timer <- Timer$new(name = paste0("timer_", mode, "_", truth))

  if (mode == "static") {
    plan0 <- make_plan_static()
  } else {
    plan0 <- make_plan_ph2a_only()
  }

  add_timepoints(tr_timer, plan0)

  # Shared state
  go_interim <- NA_integer_

  # Interim: Go/No-Go on d20 vs placebo
  trigger_by_calendar(interim_time, tr_timer, analysis = function(df, current_time) {

    dat <- df |>
      dplyr::filter(!is.na(enroll_time), arm %in% c("d0", "d20"))

    y_p <- dat$y[dat$arm == "d0"]
    y_t <- dat$y[dat$arm == "d20"]

    n_p <- length(y_p)
    n_t <- length(y_t)
    m_p <- mean(y_p)
    m_t <- mean(y_t)

    post_p <- RBesT::postmix(prior_placebo, n = n_p, m = m_p)
    post_t <- RBesT::postmix(prior_noninf,  n = n_t, m = m_t)

    prob_delta <- RBesT::pmixdiff(post_t, post_p, delta_interim, lower.tail = FALSE)
    delta_q <- RBesT::qmixdiff(post_t, post_p, c(0.025, 0.5, 0.975))

    go_interim <<- as.integer(prob_delta >= gamma_interim)

    # --- adaptive switching happens HERE ---
    if (mode == "adaptive") {
      old_plan <- get_plan_df(tr_timer)

      if (go_interim == 1L) {
        # Switch to expansion plan and add final MCP-Mod condition
        new_plan <- make_plan_ph2b_expand()

        tr_timer$timelist <- list()
        add_timepoints(tr_timer, new_plan)

        # Add final analysis only if GO
        tr_timer$add_condition(
          time == final_time,
          analysis = function(df, time) {

            dat2 <- df |>
              dplyr::filter(!is.na(enroll_time))

            summ <- dat2 |>
              dplyr::group_by(arm) |>
              dplyr::summarise(
                dose = unique(dose)[1],
                n = dplyr::n(),
                mu_hat = mean(y),
                .groups = "drop"
              ) |>
              dplyr::arrange(dose)

            n_vec <- summ$n |> as.double()
            mu_vec <- summ$mu_hat
            S_hat <- diag((sigma / sqrt(n_vec))^2)

            estimates_sim <- list(
              mu_hats = list(mu_vec),
              S_hats  = list(S_hat)
            )

            bmcp <- BayesianMCPMod::assessDesign(
              n_patients = n_vec,
              mods = mods,
              prior_list = prior_list_final,
              estimates_sim = estimates_sim,
              alpha_crit_val = alpha_final,
              n_sim = 1
            )

            data.frame(
              n_total_final = nrow(dat2),
              max_eff = attr(bmcp, "maxEff")
            )
          },
          name = "final_bayesian_mcpmod"
        )

      } else {
        # NO-GO: stop future enrollment
        new_plan <- make_plan_stop_after_interim(old_plan, interim_time = current_time)

        tr_timer$timelist <- list()
        add_timepoints(tr_timer, new_plan)
      }

      # Return plan summaries for transparency
      list(
        n_placebo = n_p,
        n_d20 = n_t,
        post_prob_delta = prob_delta,
        delta_q025 = delta_q[1],
        delta_q500 = delta_q[2],
        delta_q975 = delta_q[3],
        decision_go = go_interim,
        planned_before_switch = summarise_plan(old_plan),
        planned_after_switch = summarise_plan(get_plan_df(tr_timer))
      )

    } else {
      # Static mode: return interim stats only
      list(
        n_placebo = n_p,
        n_d20 = n_t,
        post_prob_delta = prob_delta,
        delta_q025 = delta_q[1],
        delta_q500 = delta_q[2],
        delta_q975 = delta_q[3],
        decision_go = go_interim
      )
    }
  })

  # Static mode adds final MCP-Mod upfront 
  if (mode == "static") {
    tr_timer$add_condition(
      time == final_time,
      analysis = function(df, time) {
        dat2 <- df |>
          dplyr::filter(!is.na(enroll_time))

        summ <- dat2 |>
          dplyr::group_by(arm) |>
          dplyr::summarise(
            dose = unique(dose)[1],
            n = dplyr::n(),
            mu_hat = mean(y),
            .groups = "drop"
          ) |>
          dplyr::arrange(dose)

        n_vec <- summ$n |> as.double()
        mu_vec <- summ$mu_hat
        S_hat <- diag((sigma / sqrt(n_vec))^2)

        estimates_sim <- list(
          mu_hats = list(mu_vec),
          S_hats  = list(S_hat)
        )

        bmcp <- BayesianMCPMod::assessDesign(
          n_patients = n_vec,
          mods = mods,
          prior_list = prior_list_final,
          estimates_sim = estimates_sim,
          alpha_crit_val = alpha_final,
          n_sim = 1
        )

        data.frame(
          n_total_final = nrow(dat2),
          max_eff = attr(bmcp, "maxEff")
        )
      },
      name = "final_bayesian_mcpmod"
    )
  }

  # Build and run Trial
  trial <- Trial$new(
    name = paste0("example_switching_", mode, "_", truth),
    seed = seed,
    timer = tr_timer,
    population = pops
  )

  trial$run()

  list(
    trial = trial,
    mode = mode,
    truth = truth,
    plan_initial = summarise_plan(plan0),
    plan_final = summarise_plan(get_plan_df(tr_timer)),
    enrolled_final = count_enrolled_by_arm(trial)
  )
}
```

---

# Run the comparisons

We run 4 trials:

- Static + GO truth
- Adaptive + GO truth
- Static + NO-GO truth
- Adaptive + NO-GO truth

```{r run-all}
res_static_go    <- run_trial(mode = "static",   truth = "go",   seed = 8808)
res_adapt_go     <- run_trial(mode = "adaptive", truth = "go",   seed = 8808)

res_static_nogo  <- run_trial(mode = "static",   truth = "nogo", seed = 8808)
res_adapt_nogo   <- run_trial(mode = "adaptive", truth = "nogo", seed = 8808)
```

---

# Inspect interim decision + plan switching effects

## GO truth: did adaptive switch expand?

```{r inspect-go}
key_interim <- paste0("time_", interim_time)
key_final   <- paste0("time_", final_time)

# Interim results (static vs adaptive)
res_static_go$trial$results[[key_interim]]
res_adapt_go$trial$results[[key_interim]]

# Final MCP-Mod results (static vs adaptive)
res_static_go$trial$results[[key_final]]
res_adapt_go$trial$results[[key_final]]
```

## NO-GO truth: did adaptive switch stop enrollment and skip final MCP-Mod?

```{r inspect-nogo}
# Interim results (static vs adaptive)
res_static_nogo$trial$results[[key_interim]]
res_adapt_nogo$trial$results[[key_interim]]

# Final results:
# Static has final MCP-Mod (because it's attached upfront).
# Adaptive only adds final MCP-Mod if GO; for NO-GO it should be missing.
names(res_static_nogo$trial$results[[key_final]])
res_adapt_nogo$trial$results[[key_final]]
```

---

# Compare planned enrollment (initial vs final plan)

```{r compare-plans}
bind_rows(
  res_static_go$plan_initial  |> mutate(run = "Static / GO",   stage = "initial"),
  res_static_go$plan_final    |> mutate(run = "Static / GO",   stage = "final"),
  res_adapt_go$plan_initial   |> mutate(run = "Adaptive / GO", stage = "initial"),
  res_adapt_go$plan_final     |> mutate(run = "Adaptive / GO", stage = "final"),

  res_static_nogo$plan_initial |> mutate(run = "Static / NO-GO",   stage = "initial"),
  res_static_nogo$plan_final   |> mutate(run = "Static / NO-GO",   stage = "final"),
  res_adapt_nogo$plan_initial  |> mutate(run = "Adaptive / NO-GO", stage = "initial"),
  res_adapt_nogo$plan_final    |> mutate(run = "Adaptive / NO-GO", stage = "final")
) |>
  arrange(run, stage, match(arm, arm_names))
```

---

# Compare actual enrollment (end of run)

```{r compare-enrolled}
bind_rows(
  res_static_go$enrolled_final  |> mutate(run = "Static / GO"),
  res_adapt_go$enrolled_final   |> mutate(run = "Adaptive / GO"),
  res_static_nogo$enrolled_final |> mutate(run = "Static / NO-GO"),
  res_adapt_nogo$enrolled_final  |> mutate(run = "Adaptive / NO-GO")
) |>
  arrange(run, match(arm, arm_names))
```

---

# Plan audit trail (adaptive runs)

If your updated `Trial` implementation records plan changes (e.g., `trial$plan_history`),
you can inspect them here.

```{r plan-history}
length(res_adapt_go$trial$plan_history)
res_adapt_go$trial$plan_history

length(res_adapt_nogo$trial$plan_history)
res_adapt_nogo$trial$plan_history
```

---

# Takeaways

1. **Static seamless**: The plan is fully defined upfront; Ph2a/Ph2b behavior is “encoded” by timepoints with zeros in the early period for new arms.
2. **Adaptive switching**: You can begin with a minimal Ph2a plan, then **switch the plan** at interim based on the Bayesian Go/No-Go result:
   - GO → expand to Ph2b multi-arm dose finding and attach the final Bayesian MCP-Mod analysis.
   - NO-GO → stop future enrollment and do not add final MCP-Mod.
3. This works because the updated `Trial$run()` loop re-reads the plan each step, so edits to `timer$timelist` affect subsequent timepoints in the same run.

``
