---
title: "Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 7: Two-arm | Fixed design | Single continuous endpoint | Bayesian Go/No-Go with placebo borrowing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Bayesian borrowing utilities
library(RBesT)

# Helpers
library(dplyr)

set.seed(777)
```

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single continuous endpoint** and a **single final analysis**.

- Placebo arm uses **historical borrowing** through an informative prior derived from historical placebo data and robustified with a non-informative component.
- Treatment arm uses a **non-informative prior**.
- Go/No-Go rule at final analysis:

\[
\text{Go if } P(\Delta > \delta \mid \text{data}) \ge \gamma,
\]

where \(\Delta = \mu_T - \mu_P\), with \(\delta = 0.1\) and \(\gamma = 0.8\).

```{r scenario}
# Trial design
sample_size <- 80
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Data-generating truth for this simulation example
mu_placebo_true <- 0.00
mu_treat_true   <- 0.30
sigma_known     <- 1.0

# Enrollment/dropout profile (fixed design timeline)
enrollment <- list(
  end_time = c(4, 8),
  rate     = c(6, 10)
)
dropout <- list(
  end_time = c(4, 8),
  rate     = c(0, 0)
)

# Bayesian decision thresholds
delta <- 0.1
gamma <- 0.8
```

## Historical placebo dataset and priors

We include a small historical placebo dataset as study-level means and sample sizes.

```{r priors}
hist_placebo <- data.frame(
  study = c("H1", "H2", "H3"),
  n     = c(24, 18, 20),
  mean  = c(-0.05, 0.02, 0.00)
)

hist_placebo

# Pooled historical summary used to derive an informative placebo prior
n_hist_total <- sum(hist_placebo$n)
m_hist_pooled <- weighted.mean(hist_placebo$mean, hist_placebo$n)

# Non-informative prior (very small pseudo-sample size in mn parametrization)
# Used as treatment prior and as base for historical update.
prior_noninf <- mixnorm(
  c(1, 0, 1e-6),
  sigma = sigma_known,
  param = "mn"
)

# Informative placebo prior from historical placebo summary
prior_placebo_inf <- postmix(
  prior_noninf,
  n = n_hist_total,
  m = m_hist_pooled
)

# Robustified placebo prior to protect against prior-data conflict
prior_placebo <- robustify(
  prior_placebo_inf,
  weight = 0.2,
  mean   = 0,
  n      = 1,
  sigma  = sigma_known
)

# Treatment prior remains non-informative
prior_treatment <- prior_noninf

summary(prior_placebo)
```

## Time points

Create trial time points and identify the single final analysis time.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

tr_timer <- Timer$new(name = "timer_example_7")
add_timepoints(tr_timer, timepoints)

final_time <- tr_timer$get_end_timepoint()
final_time
```

## Populations

```{r population}
n_per_arm <- sample_size / 2

pop_placebo <- Population$new(
  name = "placebo",
  data = data.frame(
    subject_id = seq_len(n_per_arm),
    y = rnorm(n_per_arm, mean = mu_placebo_true, sd = sigma_known)
  )
)

pop_treatment <- Population$new(
  name = "treatment",
  data = data.frame(
    subject_id = seq_len(n_per_arm),
    y = rnorm(n_per_arm, mean = mu_treat_true, sd = sigma_known)
  )
)
```

## Final analysis trigger (Bayesian Go/No-Go)

Only one trigger is added at `time == final_time`.

```{r trigger}
tr_timer$add_condition(
  time == final_time,
  analysis = function(df, time) {
    dat <- df |>
      dplyr::filter(!is.na(enroll_time)) |>
      dplyr::mutate(arm = factor(arm, levels = c("placebo", "treatment")))

    y_p <- dat$y[dat$arm == "placebo"]
    y_t <- dat$y[dat$arm == "treatment"]

    n_p <- length(y_p)
    n_t <- length(y_t)

    m_p <- mean(y_p)
    m_t <- mean(y_t)

    # Posterior for each arm mean
    post_placebo <- RBesT::postmix(prior_placebo, n = n_p, m = m_p)
    post_treat   <- RBesT::postmix(prior_treatment, n = n_t, m = m_t)

    # Posterior probability for treatment effect exceeding delta
    prob_delta <- RBesT::pmixdiff(post_treat, post_placebo, delta, lower.tail = FALSE)

    # Posterior summaries for Delta = mu_T - mu_P
    delta_ci <- RBesT::qmixdiff(post_treat, post_placebo, c(0.025, 0.5, 0.975))

    list(
      n_placebo = n_p,
      n_treatment = n_t,
      mean_placebo = m_p,
      mean_treatment = m_t,
      threshold_delta = delta,
      threshold_gamma = gamma,
      post_prob_delta = prob_delta,
      delta_q025 = delta_ci[1],
      delta_q500 = delta_ci[2],
      delta_q975 = delta_ci[3],
      decision_go = as.integer(prob_delta >= gamma)
    )
  },
  name = "final"
)
```

## Trial

```{r trial}
trial <- Trial$new(
  name = "example_7_bayes_two_arm_fixed",
  seed = 777,
  timer = tr_timer,
  population = list(pop_placebo, pop_treatment)
)
```

## Simulate

```{r simulate}
trial$run()
```

## Results

```{r results}
trial$results
```

```{r results-pretty}
prettify_results(trial$results)
```

`decision_go = 1` indicates **Go**, and `decision_go = 0` indicates **No-Go** under the criterion
\(P(\Delta > 0.1 \mid \text{data}) \ge 0.8\).
