---
title: "Example 11: Two arm | Fixed design | Bivariate TTE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 11: Two arm | Fixed design | Bivariate TTE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(survival)
library(MASS)   # for mvrnorm (simulate correlated endpoints)
```

Oncology and cardiovascular trials may specify two co‑primary time‑to‑event (TTE) endpoints, such as time to first disease progression and time to disease‑related death, or time to first hospitalization and time to cardiovascular death. Correlation between these TTE outcomes arises naturally from shared disease pathways. This example demonstrates simulating a joint structure for two correlated TTE endpoints and analyzing them at a single final analysis using log‑rank tests and Cox proportional hazards models for each endpoint. See [Example 4](https://boehringer-ingelheim.github.io/rxsim/articles/example-4.html) for the analogous correlated-continuous-tte setup.

## Scenario

We consider a **two-arm fixed design** with two **time-to-event (TTE)** endpoints per subject. The endpoints are **correlated** via a shared Gaussian latent structure. Enrollment is piece-wise linear; analysis happens **once at the final time**.

```{r scenario}
# Total target N (across both arms)
sample_size <- 1200

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)

# Correlation between continuous endpoint driver (Z1) and frailty for TTE (Z2)
rho <- 0.50

# TTE model parameters for first variable (Exponential PH with subject frailty exp(eta))
lambda0_1 <- 0.08    # baseline hazard (per time unit)
HR_trt_1  <- 0.70    # treatment hazard ratio (<1 favors treatment)

# TTE model parameters for second variable (Exponential PH with subject frailty exp(eta))
lambda0_2 <- 0.04  # baseline hazard for death (per time unit)
HR_trt_2  <- 0.75  # treatment hazard ratio (<1 favors treatment)

# Frailty scale (eta = sigma_frailty * Z2)
sigma_frailty <- 0.6

# Covariance of (Z1, Z2)
SigmaZ <- matrix(c(1, rho, rho, 1), nrow = 2)

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  rho = rho,
  lambda0_1 = lambda0_1,
  lambda0_2 = lambda0_2,
  hr_trt_true_1 = HR_trt_1,
  hr_trt_true_2 = HR_trt_2
)
```
`HR_trt_1 = 0.70` encodes a 30% hazard reduction in the treatment arm — a clinically meaningful effect in many oncology settings. `lambda0_1 = 0.08` is the baseline event rate per time unit; combined with n=1200, this ensures a substantial number of TTE events even with only 3 replicates. Additionally, `HR_trt_2 = 0.75` encodes a 25% hazard reduction in the treatment arm, and `lambda0_2 = 0.04` is the baseline event rate per time unit. `sigma_frailty = 0.6` controls the strength of the subject-level frailty term: larger values introduce more between-subject heterogeneity in TTE, making the frailty the shared biological driver linking the two TTE endpoints through the bivariate latent (Z1, Z2) with `rho = 0.50`.

## Time points

Generate the discrete timepoints and add them to a `Timer`.

```{r timepoints}
plan <- gen_plan(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment_fn,
  dropout     = dropout_fn
)
```

`gen_plan()` is called with `enrollment_fn` as a stochastic function, so each invocation by `replicate_trial()` generates a fresh per-replicate enrollment schedule. This differs from the piecewise-constant list approach in [Example 3](https://boehringer-ingelheim.github.io/rxsim/articles/example-3.html): here there is no deterministic timer pre-built — enrollment timing varies across replicates, which is the more realistic simulation regime for large trials.

## Arms (Populations)
Per arm we simulate subject-level latent bivariate normals `(Z1, Z2)` with correlation `rho`.  
- The **TTE** endpoints use a proportional-hazards exponential model with subject-specific frailty:  
  `hazard_i = lambda_arm_i * exp(eta_i)`, where `eta_i = sigma_frailty * Z_i` and `lambda_trt_i = lambda0_i * HR_trt_i`.
  Event time `T_i ~ Exponential(rate = hazard_i)`.

```{r population}
# Create generator function for arm-specific data
mk_population_generator <- function(lambda_arm_1, lambda_arm_2) {
  function(n) {
    Z <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = SigmaZ)
    Z1 <- Z[, 1]
    Z2 <- Z[, 2]
    eta_1 <- sigma_frailty * Z1
    eta_2 <- sigma_frailty * Z2
    rate_1 <- lambda_arm_1 * exp(eta_1)
    rate_2 <- lambda_arm_2 * exp(eta_2)
    tte_true_1 <- rexp(n, rate = rate_1)
    tte_true_2 <- rexp(n, rate = rate_2)

    data.frame(
      id = seq_len(n),
      tte_true_1 = tte_true_1,
      tte_true_2 = tte_true_2,
      readout_time = 1
    )
  }
}

lambda_pbo_1 <- lambda0_1
lambda_trt_1 <- lambda0_1 * HR_trt_1

lambda_pbo_2 <- lambda0_2
lambda_trt_2 <- lambda0_2 * HR_trt_2

population_generators <- list(
  pbo = mk_population_generator(lambda_arm_1 = lambda_pbo_1, lambda_arm_2 = lambda_pbo_2),
  trt = mk_population_generator(lambda_arm_1 = lambda_trt_1, lambda_arm_2 = lambda_trt_2)
)
```
Each subject gets a latent pair `(Z1, Z2)` drawn from a bivariate standard normal with correlation `rho`. Z1 drives TTE endpoint one through a multiplicative frailty: `eta_i = sigma_frailty * Z_i`, so `hazard_i = lambda_arm_i * exp(eta_i)`, and `T_i ~ Exponential(rate = hazard_i)`. The shared latent structure means subjects with higher Z2 (higher frailty, shorter TTE) also tend to have higher Z1 (higher frailty, shorter TTE), modelling a biological scenario where the TTE endpoints share a common underlying driver.

## Trigger & Analysis
This is a **fixed design**: analyze **once at the final timepoint**.
We run a **log-rank test** plus **Cox PH** for each TTE endpoint.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      df_e <- df[!is.na(df$enroll_time), , drop = FALSE]
      if (nrow(df_e) == 0) {
        return(data.frame(scenario, note = "no enrolled subjects", stringsAsFactors = FALSE))
      }

      # Build observed TTE with censoring
      censor_drop  <- ifelse(is.na(df_e$drop_time), Inf, pmax(0, df_e$drop_time - df_e$enroll_time))
      censor_admin <- pmax(0, max(df_e$enroll_time) + 100 - df_e$enroll_time)

      tte_true_1 <- df_e$tte_true_1
      t_obs_1    <- pmin(tte_true_1, censor_drop, censor_admin)
      status_1   <- as.integer(tte_true_1 <= pmin(censor_drop, censor_admin))
      
      tte_true_2 <- df_e$tte_true_2
      t_obs_2    <- pmin(tte_true_2, censor_drop, censor_admin)
      status_2   <- as.integer(tte_true_2 <= pmin(censor_drop, censor_admin))

      # Survival analysis: log-rank and Cox (first variable)
      surv_p_1 <- NA_real_
      hr_1 <- hr_lo_1 <- hr_hi_1 <- NA_real_
      events_1 <- sum(status_1)

      if (length(unique(df_e$arm)) == 2 && events_1 > 0 && all(t_obs_1 >= 0)) {
        S_1 <- Surv(time = t_obs_1, event = status_1)

        surv_p_1 <- tryCatch({
          sd_1 <- survdiff(S_1 ~ arm, data = df_e)
          1 - pchisq(sd_1$chisq, df = length(sd_1$n) - 1)
        }, error = function(e) NA_real_)

        cox_1 <- tryCatch(coxph(S_1 ~ arm, data = df_e), error = function(e) NULL)
        if (!is.null(cox_1)) {
          s_1 <- summary(cox_1)
          hr_1    <- unname(s_1$coef[1, "exp(coef)"])
          hr_lo_1 <- unname(s_1$conf.int[1, "lower .95"])
          hr_hi_1 <- unname(s_1$conf.int[1, "upper .95"])
        }
      }
      
      # Survival analysis: log-rank and Cox (second variable)
      surv_p_2 <- NA_real_
      hr_2 <- hr_lo_2 <- hr_hi_2 <- NA_real_
      events_2 <- sum(status_2)

      if (length(unique(df_e$arm)) == 2 && events_2 > 0 && all(t_obs_2 >= 0)) {
        S_2 <- Surv(time = t_obs_2, event = status_2)

        surv_p_2 <- tryCatch({
          sd_2 <- survdiff(S_2 ~ arm, data = df_e)
          1 - pchisq(sd_2$chisq, df = length(sd_2$n) - 1)
        }, error = function(e) NA_real_)

        cox_2 <- tryCatch(coxph(S_2 ~ arm, data = df_e), error = function(e) NULL)
        if (!is.null(cox_2)) {
          s_2 <- summary(cox_2)
          hr_2    <- unname(s_2$coef[1, "exp(coef)"])
          hr_lo_2 <- unname(s_2$conf.int[1, "lower .95"])
          hr_hi_2 <- unname(s_2$conf.int[1, "upper .95"])
        }
      }

      data.frame(
        scenario,
        n_total    = nrow(df_e),
        n_pbo      = sum(df_e$arm == "pbo"),
        n_trt      = sum(df_e$arm == "trt"),
        logrank_p_1  = surv_p_1,
        hr_cox_1     = hr_1,
        hr_ci_lo_1   = hr_lo_1,
        hr_ci_hi_1   = hr_hi_1,
        logrank_p_2  = surv_p_2,
        hr_cox_2     = hr_2,
        hr_ci_lo_2   = hr_lo_2,
        hr_ci_hi_2   = hr_hi_2,
        events_1     = events_1,
        events_2     = events_2,
        stringsAsFactors = FALSE
      )
    }
  )
)
```

`censor_drop` censors a subject at the time elapsed between enrollment and dropout (or infinity if no dropout). `censor_admin` provides administrative censoring: each subject is followed for 100 time units after the last enrollment, a common end-of-study rule. `t_obs` and `status` capture what would actually be observed. The `Surv()` object is then passed to `survdiff()` for the log-rank p-value and `coxph()` for the Cox HR and 95% CI. `tryCatch` guards against edge cases (e.g., one arm with no events) so the simulation continues rather than stopping on an error.

## Trial
Make multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "two_endpoints_bivariate_tte_fixed",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate
To simulate all replicates:

```{r sim}
run_trials(trials)
```

## Results

Bind one row per replicate into one data frame.

```{r results}
replicate_results <- do.call(rbind, lapply(seq_along(trials), function(i) {
  data.frame(
    replicate = i,
    trials[[i]]$results[[1]]$final,
    stringsAsFactors = FALSE
  )
}))

replicate_results
```

`collect_results()` returns a data.frame with `replicate`, `timepoint`, and `analysis` columns prepended to your analysis columns. The `events` column shows how many TTE events were observed — with n=1200 and `lambda0_1 = 0.04; lambda0_2 = 0.08`, this should be large. `logrank_p` tests the null of equal survival curves; `hr_cox` is the estimated hazard ratio from the Cox model (values < 1 favour treatment), and `hr_ci_lo/hr_ci_hi` are its 95% CI bounds. Because the TTE endpoints share the same latent drivers, replicates with a small `logrank_p_1` tend to also have a small `logrank_p_2`, though the correspondence is imperfect due to the different statistical tests and the censoring mechanism.
