---
title: "Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Helpers
library(dplyr)

set.seed(6606)
```

Subgroup analyses are a standard exploratory component of clinical trial
reporting. This example simulates a trial where the treatment effect differs
between two patient subgroups — a form of treatment effect heterogeneity. We
evaluate the overall treatment effect, subgroup-specific effects, and formally
test whether the arm-by-subgroup interaction is statistically significant.

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single
continuous endpoint** and one **final analysis**.

At final analysis, we report:

- **Overall analysis**: treatment effect across all subjects.
- **Exploratory subgroup analysis**: treatment effect within each subgroup +
  arm-by-subgroup interaction test.

Treatment effect heterogeneity arises when patients with different baseline
characteristics — such as disease subtype or biomarker status — respond
differently to treatment. Here, `delta_A = 0.40` and `delta_B = 0.15` encode
a larger benefit in subgroup A; treatment is effective in both subgroups, but
considerably more so in A. `subgroup_prob = c(0.6, 0.4)` means that on
average 60% of enrolled patients belong to subgroup A and 40% to subgroup B,
though actual proportions vary across replicates due to random assignment.

```{r scenario}
# Trial design
sample_size <- 120
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Subgroup settings
subgroup_levels <- c("A", "B")
subgroup_prob   <- c(0.6, 0.4)

# Data-generating truth (example values)
mu_placebo_A <- 0.00
mu_placebo_B <- 0.00
delta_A      <- 0.40  # true treatment effect in subgroup A
delta_B      <- 0.15  # true treatment effect in subgroup B
sigma        <- 1.0

# Enrollment/dropout profile
enrollment <- list(
  end_time = c(4, 8),
  rate     = c(8, 7)
)
dropout <- list(
  end_time = c(4, 8),
  rate     = c(0, 0)
)

scenario <- tidyr::expand_grid(
  sample_size  = sample_size,
  allocation   = list(allocation),
  delta_A      = delta_A,
  delta_B      = delta_B,
  p_subgroup_A = subgroup_prob[1],
  p_subgroup_B = subgroup_prob[2]
)
```

## Time points

`gen_timepoints()` uses a two-interval piecewise-constant schedule (rates 8 and
7 per time unit) to reach the target sample size of 120. Both dropout rates are
zero, so all enrolled subjects complete the trial and contribute to the final
analysis.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

tr_timer <- Timer$new(name = "timer_example_6")
add_timepoints(tr_timer, timepoints)

final_time <- tr_timer$get_end_timepoint()
final_time
```

## Populations

Create two populations and include a `subgroup` column in each population
dataset.

In each generator, subgroup membership is randomly drawn for each subject with
probabilities `c(0.6, 0.4)`, and the arm-specific mean is selected conditional
on that assignment. Because subgroup is sampled independently in each replicate,
realized subgroup counts — and hence per-subgroup sample sizes — vary across
replicates, reflecting a realistic pre-stratified (but not block-randomised)
design.

```{r populations}
# Create generator for arm-specific data with subgroups
population_generators <- list(
  placebo = function(n) {
    subgroup <- sample(subgroup_levels, size = n, replace = TRUE, prob = subgroup_prob)
    mu <- ifelse(subgroup == "A", mu_placebo_A, mu_placebo_B)
    data.frame(
      id = seq_len(n),
      subgroup = subgroup,
      y = rnorm(n, mean = mu, sd = sigma),
      readout_time = 1
    )
  },
  treatment = function(n) {
    subgroup <- sample(subgroup_levels, size = n, replace = TRUE, prob = subgroup_prob)
    mu <- ifelse(subgroup == "A", mu_placebo_A + delta_A, mu_placebo_B + delta_B)
    data.frame(
      id = seq_len(n),
      subgroup = subgroup,
      y = rnorm(n, mean = mu, sd = sigma),
      readout_time = 1
    )
  }
)
```

## Trial Parameters

`enrollment_fn` and `dropout_fn` supply the inter-arrival and time-to-dropout
distributions for individual subjects. These are passed to `replicate_trial()`,
which uses them when building each replicate's enrollment schedule.

```{r trial_params}
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
```

## Final analysis trigger

`lm(y ~ arm)` estimates the overall treatment effect across all subjects.
Per-subgroup effects are obtained by fitting `lm(y ~ arm)` separately within
each subgroup subset. The interaction model `lm(y ~ arm * subgroup)` includes an
`arm:subgroupB` coefficient that tests directly whether the treatment effect in
subgroup B differs from that in the reference subgroup A — a statistically
significant result indicates genuine treatment effect heterogeneity.

```{r analysis}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      dat <- df |>
        dplyr::filter(!is.na(enroll_time)) |>
        dplyr::mutate(
          arm = factor(arm, levels = c("placebo", "treatment")),
          subgroup = factor(subgroup, levels = subgroup_levels)
        )

      fit_overall <- lm(y ~ arm, data = dat)
      coef_overall <- summary(fit_overall)$coefficients

      overall_est <- unname(coef_overall["armtreatment", "Estimate"])
      overall_p   <- unname(coef_overall["armtreatment", "Pr(>|t|)"])

      subgroup_stats <- lapply(levels(dat$subgroup), function(sg) {
        dsg <- dat[dat$subgroup == sg, , drop = FALSE]
        fit_sg <- lm(y ~ arm, data = dsg)
        coef_sg <- summary(fit_sg)$coefficients

        c(
          n_total = nrow(dsg),
          estimate = unname(coef_sg["armtreatment", "Estimate"]),
          p_value = unname(coef_sg["armtreatment", "Pr(>|t|)"])
        )
      })
      names(subgroup_stats) <- levels(dat$subgroup)

      fit_interaction <- lm(y ~ arm * subgroup, data = dat)
      coef_int <- summary(fit_interaction)$coefficients
      int_term <- grep("^arm.*:subgroup", rownames(coef_int), value = TRUE)
      p_interaction <- if (length(int_term) > 0) {
        unname(coef_int[int_term[1], "Pr(>|t|)"])
      } else {
        NA_real_
      }

      data.frame(
        scenario,
        n_total = nrow(dat),
        overall_estimate = overall_est,
        overall_p_value = overall_p,
        subgroup_A_n = unname(subgroup_stats[["A"]]["n_total"]),
        subgroup_A_estimate = unname(subgroup_stats[["A"]]["estimate"]),
        subgroup_A_p_value = unname(subgroup_stats[["A"]]["p_value"]),
        subgroup_B_n = unname(subgroup_stats[["B"]]["n_total"]),
        subgroup_B_estimate = unname(subgroup_stats[["B"]]["estimate"]),
        subgroup_B_p_value = unname(subgroup_stats[["B"]]["p_value"]),
        interaction_p_value = p_interaction,
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Trial

```{r trial}
trials <- replicate_trial(
  trial_name = "example_6_two_arm_fixed_subgroup",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r simulate}
run_trials(trials)
```

## Results

`collect_results()` row-binds analysis outputs across all replicates and
prepends `replicate` (integer index), `timepoint` (calendar time at which the
analysis fired), and `analysis` (the analysis name) to each row.
`overall_estimate` and `overall_p_value` summarise the pooled treatment effect;
the per-subgroup columns reveal within-group signals. Note that with
approximately 24 patients per arm in subgroup B (40% of 60 treated),
subgroup-level power is limited and p-values will be highly variable — this is
intentional, since exploratory subgroup analyses are hypothesis-generating and
should not support confirmatory conclusions. A small `interaction_p_value`
suggests genuine treatment effect heterogeneity, but interpretation should be
made cautiously given the sample sizes. See [Getting
Started](getting-started.html) for simulation setup and [Core
Concepts](concepts.html) for background.

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```
