---
title: "Example 1: Two arm | Fixed design | Continuous"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 1: Two arm | Fixed design | Continuous}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(rxsim)
```

This example simulates a Phase II/III parallel-group trial evaluating a new
treatment against placebo on a continuous primary endpoint (e.g., a biomarker
score or symptom scale). With 100 subjects randomised 1:1, we run a two-sample
t-test at full enrollment and track the p-value and arm means across simulation
replicates. This is the simplest complete rxsim workflow and a good starting
point before moving to more complex designs. See
[Getting Started](getting-started.html) for an overview of the package.

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
sample_size <- 100
arms <- c("pbo", "trt")
allocation <- c(1,1)
true_delta <- 0.2
enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.01)
scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  true_delta = true_delta
)
```

`allocation = c(1, 1)` specifies balanced randomisation, equal expected numbers
per arm. Using `tidyr::expand_grid()` to build the `scenario` data.frame embeds
the design parameters directly into each analysis result row, making results
self-documenting and easy to compare across parameter sweeps.

## Populations
Define population generators.

```{r population}
population_generators <- list(
  pbo = function(n) data.frame(
    id = 1:n,
    value = rnorm(n),
    readout_time = 1
  ),
  trt = function(n) data.frame(
    id = 1:n,
    value = rnorm(n, true_delta),
    readout_time = 1
  )
)
```

Each population generator is a function of `n` that returns a `data.frame` with
one row per subject. `readout_time = 1` means the endpoint is observed exactly
1 time unit after a subject enrolls. The placebo arm has a mean of 0 and the
treatment arm a mean of `true_delta`, a small-to-moderate standardised effect.

## Triggers & Analysis
We want to do a t-test when `sample_size` subjects have been enrolled.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer){
      df_enrolled <- df |> subset(!is.na(enroll_time))
      tt <- t.test(value ~ arm, data = df_enrolled)
      data.frame(
        scenario,
        n_total = nrow(df_enrolled),
        mean_pbo = mean(df_enrolled$value[df_enrolled$arm == "pbo"]),
        mean_trt = mean(df_enrolled$value[df_enrolled$arm == "trt"]),
        p_value = unname(tt$p.value),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

The trigger fires when the cumulative count of enrolled subjects
(`sum(!is.na(enroll_time))`) reaches `sample_size`. Inside the analysis
function, `subset(!is.na(enroll_time))` drops subjects who have been allocated
but not yet enrolled. rxsim pre-generates the full allocation list so filtering
is essential. The two-sample t-test then compares endpoint values between the
two enrolled arms.

## Trial
Make multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "test_trial",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate
To simulate all replicates:

```{r sim}
run_trials(trials)
```

Bind one row per replicate into one data frame.
```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`collect_results()` returns a data.frame with `replicate`, `timepoint`, and
`analysis` columns prepended to your analysis columns. The `p_value` column
varies across replicates because each replicate draws fresh random data.