---
title: "Example 4: Two arm | Fixed design | Correlated continuous & time-to-event endpoints | t-test + survival analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 4: Two arm | Fixed design | Correlated continuous & time-to-event endpoints | t-test + survival analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, gen_timepoints, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)      # for mvrnorm (simulate correlated latent variables)
library(survival)  # for Surv, survdiff, coxph
```

## Scenario

We consider a **two-arm fixed design** with a **continuous endpoint (Y)** and a **time-to-event (TTE)** endpoint per subject.  
The endpoints are **correlated** via a shared Gaussian latent structure. Enrollment is piece-wise linear; analysis happens **once at the final time**.

```{r scenario}
# Total target N (across both arms)
sample_size <- 1200

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

# Piece-wise linear enrollment and dropout rates (per unit time)
# (Tune these to your development plan.)
enrollment <- list(
  end_time = c(6, 12, 18, 24),
  rate     = c(4,  8, 12, 16)
)
dropout <- list(
  end_time = c(6, 12, 18, 24),
  rate     = c(1,  2,  4,  6)
)

# Correlation between continuous endpoint driver (Z1) and frailty for TTE (Z2)
rho <- 0.50

# Continuous endpoint parameters
mu_pbo <- 0.0
mu_trt <- 0.3      # treatment mean shift
sd_y   <- 1.0

# TTE model parameters (Exponential PH with subject frailty exp(eta))
lambda0 <- 0.08    # baseline hazard (per time unit)
HR_trt  <- 0.70    # treatment hazard ratio (<1 favors treatment)

# Frailty scale (eta = sigma_frailty * Z2)
sigma_frailty <- 0.6

# Covariance of (Z1, Z2)
SigmaZ <- matrix(c(1, rho, rho, 1), nrow = 2)
```

## Time points

Generate the discrete timepoints and add them to a `Timer`.

```{r timepoints}
timepoints <- gen_timepoints(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = enrollment,
  dropout     = dropout
)

t <- Timer$new(name = "trial_timer_cont_tte_two_endpoints")
add_timepoints(t, timepoints)

final_time <- t$get_end_timepoint()
final_time
```

## Arms (Populations)

Per arm we simulate subject-level latent bivariate normals `(Z1, Z2)` with correlation `rho`.  
- The **continuous** endpoint is `Y = mu_arm + sd_y * Z1`.
- The **TTE** endpoint uses a proportional-hazards exponential model with subject-specific frailty:  
  `hazard_i = lambda_arm * exp(eta_i)`, where `eta_i = sigma_frailty * Z2` and `lambda_trt = lambda0 * HR_trt`.
  Event time `T_i ~ Exponential(rate = hazard_i)`.

```{r population}
# Create generator function for arm-specific data
mk_population_generator <- function(mu_y, lambda_arm) {
  function(n) {
    Z <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = SigmaZ)
    Z1 <- Z[, 1]
    Z2 <- Z[, 2]
    y  <- mu_y + sd_y * Z1
    eta <- sigma_frailty * Z2
    rate_i <- lambda_arm * exp(eta)
    tte_true <- rexp(n, rate = rate_i)

    data.frame(
      id = seq_len(n),
      y  = y,
      tte_true = tte_true,
      readout_time = 1
    )
  }
}

lambda_pbo <- lambda0
lambda_trt <- lambda0 * HR_trt

population_generators <- list(
  pbo = mk_population_generator(mu_y = mu_pbo, lambda_arm = lambda_pbo),
  trt = mk_population_generator(mu_y = mu_trt, lambda_arm = lambda_trt)
)
```

## Trigger & Analysis

This is a **fixed design**: analyze **once at the final timepoint**.
We run a **two-sample t-test** for the continuous endpoint and a **log-rank test** plus **Cox PH** for the TTE endpoint.

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      df_e <- df |> dplyr::filter(!is.na(enroll_time))
      if (nrow(df_e) == 0) return(list(note = "no enrolled subjects"))

      # Build observed TTE with censoring
      censor_drop  <- ifelse(is.na(df_e$drop_time), Inf, pmax(0, df_e$drop_time - df_e$enroll_time))
      censor_admin <- pmax(0, max(df_e$enroll_time) + 100 - df_e$enroll_time)

      tte_true <- df_e$tte_true
      t_obs    <- pmin(tte_true, censor_drop, censor_admin)
      status   <- as.integer(tte_true <= pmin(censor_drop, censor_admin))

      # Continuous endpoint t-test
      p_t <- tryCatch(
        t.test(y ~ arm, data = df_e)$p.value,
        error = function(e) NA_real_
      )

      # Survival analysis: log-rank and Cox
      surv_p <- NA_real_
      hr <- hr_lo <- hr_hi <- NA_real_
      events <- sum(status)

      if (length(unique(df_e$arm)) == 2 && events > 0 && all(t_obs >= 0)) {
        S <- Surv(time = t_obs, event = status)

        surv_p <- tryCatch({
          sd <- survdiff(S ~ arm, data = df_e)
          1 - pchisq(sd$chisq, df = length(sd$n) - 1)
        }, error = function(e) NA_real_)

        cox <- tryCatch(coxph(S ~ arm, data = df_e), error = function(e) NULL)
        if (!is.null(cox)) {
          s <- summary(cox)
          hr    <- unname(s$coef[1, "exp(coef)"])
          hr_lo <- unname(s$conf.int[1, "lower .95"])
          hr_hi <- unname(s$conf.int[1, "upper .95"])
        }
      }

      list(
        n_total    = nrow(df_e),
        n_pbo      = sum(df_e$arm == "pbo"),
        n_trt      = sum(df_e$arm == "trt"),
        p_ttest_y  = p_t,
        logrank_p  = surv_p,
        hr_cox     = hr,
        hr_ci_lo   = hr_lo,
        hr_ci_hi   = hr_hi,
        events     = events
      )
    }
  )
)
```

## Trial

Create and run multiple trial replicates.

```{r trial}
trials <- replicate_trial(
  trial_name = "two_endpoints_continuous_tte_fixed",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment_fn,
  dropout = dropout_fn,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

## Simulate

```{r sim}
run_trials(trials)
```

## Results

Analysis results from the first replicate:

```{r results}
trials[[1]]$results
```

Prettified results:

```{r results_pretty}
prettify_results(trials[[1]]$results)
```

## Locked data

Snapshot from the first replicate:

```{r locked}
lapply(trials[[1]]$locked_data, head)
```
