---
title: "Example Multi Run: Two arm | Interim | Continuous | t-test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Multi Run: Two arm | Interim | Continuous | t-test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rxsim)
```

## Scenario
Capture scenario parameters. We will assume piece-wise linear enrollment.

```{r scenario}
set.seed(123)
sample_size <- 6
arms        <- c("pbo", "trt")
allocation  <- c(1, 1)
enrollment  <- function(n) rexp(n, rate = 1)
dropout     <- function(n) rexp(n, rate = 0.01)

```

## Arms
Generate populations and endpoint data using generators (functions).

```{r population}
#update the generators to have sample size 50

population_generators <- list(
  pbo = function(n) vector_to_dataframe(stats::rnorm(n)),
  trt = function(n) vector_to_dataframe(stats::rnorm(n, mean = 2, sd = 3.14))
)

```

## Triggers & Analysis
We want to do a t-test when 100 subjects have been enrolled. 

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ), #can hold 1+ filter conditions
    analysis = function(df, timer) {
      df |> dplyr::group_by(arm) |> dplyr::summarise(avg = mean(data))
    }
  )
)
```

## Trial
Make the trial objects for when data is using generators.

```{r trial}
trials <- replicate_trial(
  trial_name    = "trial",
  sample_size   = sample_size,
  arms          = arms,
  allocation    = allocation,
  enrollment    = enrollment,
  dropout       = dropout,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n             = 3
)
```

## Simluate
To run a list of trial objects us the run_multiple_trial function

```{r sim}
 run_trials(trials)
```

Get results of first trial.
```{r results}
results <- trials[[1]]$results
```

Get locked data of first trial.
```{r locked}
lapply(trials[[1]]$locked_data, head)
```
