---
title: "Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 6: Two-arm | Fixed design | Single continuous endpoint | Overall + exploratory subgroup analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core simulation framework
library(rxsim)

# Helpers
library(dplyr)

set.seed(6606)
```

Subgroup analyses are a standard exploratory component of clinical trial
reporting. This example simulates a trial where the treatment effect differs
between two patient subgroups - a form of treatment effect heterogeneity. We
evaluate the overall treatment effect, subgroup-specific effects, and formally
test whether the arm-by-subgroup interaction is statistically significant.

**Unique focus:** treatment effect heterogeneity across subgroups, overall
vs subgroup-specific power, interaction test for heterogeneity.

## Scenario

A **two-arm, fixed design** trial (placebo vs treatment) with a **single
continuous endpoint** and one **final analysis**.

At final analysis, we report:

- **Overall analysis**: treatment effect across all subjects.
- **Exploratory subgroup analysis**: treatment effect within each subgroup +
  arm-by-subgroup interaction test.

Treatment effect heterogeneity arises when patients with different baseline
characteristics - such as disease subtype or biomarker status - respond
differently to treatment. Here, `delta_A = 0.40` and `delta_B = 0.15` encode
a larger benefit in subgroup A; treatment is effective in both subgroups, but
considerably more so in A. `subgroup_prob = c(0.6, 0.4)` means that on
average 60% of enrolled patients belong to subgroup A and 40% to subgroup B,
though actual proportions vary across replicates due to random assignment.

```{r scenario}
# Trial design
sample_size <- 120
allocation  <- c(1, 1)
arms        <- c("placebo", "treatment")

# Subgroup settings
subgroup_levels <- c("A", "B")
subgroup_prob   <- c(0.6, 0.4)

# Data-generating truth (example values)
mu_placebo_A <- 0.00
mu_placebo_B <- 0.00
delta_A      <- 0.40  # true treatment effect in subgroup A
delta_B      <- 0.15  # true treatment effect in subgroup B
sigma        <- 1.0

scenario <- tidyr::expand_grid(
  sample_size  = sample_size,
  allocation   = list(allocation),
  delta_A      = delta_A,
  delta_B      = delta_B,
  p_subgroup_A = subgroup_prob[1],
  p_subgroup_B = subgroup_prob[2]
)

enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn    <- function(n) rexp(n, rate = 0.01)
```

## Populations

Create two populations and include a `subgroup` column in each population
dataset.

In each generator, subgroup membership is randomly drawn for each subject with
probabilities `c(0.6, 0.4)`, and the arm-specific mean is selected conditional
on that assignment. Because subgroup is sampled independently in each replicate,
realized subgroup counts - and hence per-subgroup sample sizes - vary across
replicates, reflecting a realistic pre-stratified (but not block-randomised)
design.

```{r populations}
# Create generator for arm-specific data with subgroups
population_generators <- list(
  placebo = function(n) {
    subgroup <- sample(subgroup_levels, size = n, replace = TRUE, prob = subgroup_prob)
    mu <- ifelse(subgroup == "A", mu_placebo_A, mu_placebo_B)
    data.frame(
      id = seq_len(n),
      subgroup = subgroup,
      y = rnorm(n, mean = mu, sd = sigma),
      readout_time = 1
    )
  },
  treatment = function(n) {
    subgroup <- sample(subgroup_levels, size = n, replace = TRUE, prob = subgroup_prob)
    mu <- ifelse(subgroup == "A", mu_placebo_A + delta_A, mu_placebo_B + delta_B)
    data.frame(
      id = seq_len(n),
      subgroup = subgroup,
      y = rnorm(n, mean = mu, sd = sigma),
      readout_time = 1
    )
  }
)
```

## Final analysis trigger

`lm(y ~ arm)` estimates the overall treatment effect across all subjects.
Per-subgroup effects are obtained by fitting `lm(y ~ arm)` separately within
each subgroup subset. The interaction model `lm(y ~ arm * subgroup)` includes an
`arm:subgroupB` coefficient that tests directly whether the treatment effect in
subgroup B differs from that in the reference subgroup A - a statistically
significant result indicates genuine treatment effect heterogeneity.

```{r analysis}
analysis_generators <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      dat <- df |>
        dplyr::filter(!is.na(enroll_time)) |>
        dplyr::mutate(
          arm = factor(arm, levels = c("placebo", "treatment")),
          subgroup = factor(subgroup, levels = subgroup_levels)
        )

      fit_overall <- lm(y ~ arm, data = dat)
      coef_overall <- summary(fit_overall)$coefficients

      overall_est <- unname(coef_overall["armtreatment", "Estimate"])
      overall_p   <- unname(coef_overall["armtreatment", "Pr(>|t|)"])

      subgroup_stats <- lapply(levels(dat$subgroup), function(sg) {
        dsg <- dat[dat$subgroup == sg, , drop = FALSE]
        fit_sg <- lm(y ~ arm, data = dsg)
        coef_sg <- summary(fit_sg)$coefficients

        c(
          n_total = nrow(dsg),
          estimate = unname(coef_sg["armtreatment", "Estimate"]),
          p_value = unname(coef_sg["armtreatment", "Pr(>|t|)"])
        )
      })
      names(subgroup_stats) <- levels(dat$subgroup)

      fit_interaction <- lm(y ~ arm * subgroup, data = dat)
      coef_int <- summary(fit_interaction)$coefficients
      int_term <- grep("^arm.*:subgroup", rownames(coef_int), value = TRUE)
      p_interaction <- if (length(int_term) > 0) {
        unname(coef_int[int_term[1], "Pr(>|t|)"])
      } else {
        NA_real_
      }

      data.frame(
        scenario,
        n_total = nrow(dat),
        overall_estimate = overall_est,
        overall_p_value = overall_p,
        subgroup_A_n = unname(subgroup_stats[["A"]]["n_total"]),
        subgroup_A_estimate = unname(subgroup_stats[["A"]]["estimate"]),
        subgroup_A_p_value = unname(subgroup_stats[["A"]]["p_value"]),
        subgroup_B_n = unname(subgroup_stats[["B"]]["n_total"]),
        subgroup_B_estimate = unname(subgroup_stats[["B"]]["estimate"]),
        subgroup_B_p_value = unname(subgroup_stats[["B"]]["p_value"]),
        interaction_p_value = p_interaction,
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Simulate

```{r trial, results=FALSE}
set.seed(6)
trials <- replicate_trial(
  trial_name            = "example_6_two_arm_fixed_subgroup",
  sample_size           = sample_size,
  arms                  = arms,
  allocation            = allocation,
  enrollment            = enrollment_fn,
  dropout               = dropout_fn,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 5
)
run_trials(trials)
```

## Results

`collect_results()` row-binds analysis outputs across all replicates and
prepends `replicate` (integer index), `timepoint` (calendar time at which the
analysis fired), and `analysis` (the analysis name) to each row.
`overall_estimate` and `overall_p_value` summarise the pooled treatment effect;
the per-subgroup columns reveal within-group signals. Note that with
approximately 24 patients per arm in subgroup B (40% of 60 treated),
subgroup-level power is limited and p-values will be highly variable - this is
intentional, since exploratory subgroup analyses are hypothesis-generating and
should not support confirmatory conclusions. A small `interaction_p_value`
suggests genuine treatment effect heterogeneity, but interpretation should be
made cautiously given the sample sizes.

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

## Power curve

How does overall and subgroup-specific power vary with the subgroup A effect
size (`delta_A`), while holding `delta_B = 0.15` fixed?

```{r power_subgroup, cache=TRUE}
set.seed(66)
n_reps_pw  <- 200
deltaA_vals <- c(0.15, 0.25, 0.35, 0.45, 0.55)

pw_df6 <- do.call(rbind, lapply(deltaA_vals, function(dA) {
  pop_pw <- list(
    placebo = function(n) {
      sg <- sample(c("A", "B"), n, TRUE, subgroup_prob)
      data.frame(
        id = 1:n, subgroup = sg,
        y  = rnorm(n, 0, sigma), readout_time = 1
      )
    },
    treatment = local({
      da <- dA
      function(n) {
        sg <- sample(c("A", "B"), n, TRUE, subgroup_prob)
        mu <- ifelse(sg == "A", da, delta_B)
        data.frame(
          id = 1:n, subgroup = sg,
          y  = rnorm(n, mu, sigma), readout_time = 1
        )
      }
    })
  )
  an_pw <- list(final = list(
    trigger  = enroll_trigger(1.0, sample_size),
    analysis = function(df, ct) {
      dat <- dplyr::filter(df, !is.na(enroll_time)) |>
        dplyr::mutate(
          arm      = factor(arm, levels = c("placebo", "treatment")),
          subgroup = factor(subgroup, levels = c("A", "B"))
        )
      ov_p <- summary(lm(y ~ arm, data = dat))$coef["armtreatment", "Pr(>|t|)"]
      sg_A <- dplyr::filter(dat, subgroup == "A")
      sg_B <- dplyr::filter(dat, subgroup == "B")
      pA <- tryCatch(
        summary(lm(y ~ arm, data = sg_A))$coef["armtreatment", "Pr(>|t|)"],
        error = function(e) NA_real_
      )
      pB <- tryCatch(
        summary(lm(y ~ arm, data = sg_B))$coef["armtreatment", "Pr(>|t|)"],
        error = function(e) NA_real_
      )
      data.frame(overall_p = ov_p, p_A = pA, p_B = pB)
    }
  ))
  tr <- replicate_trial(
    "pw6", sample_size, arms, allocation,
    enrollment_fn, dropout_fn, an_pw, pop_pw, n_reps_pw
  )
  invisible(run_trials(tr))
  res <- collect_results(tr)
  data.frame(
    delta_A        = dA,
    power_overall  = mean(res$overall_p < 0.05, na.rm = TRUE),
    power_sgA      = mean(res$p_A       < 0.05, na.rm = TRUE),
    power_sgB      = mean(res$p_B       < 0.05, na.rm = TRUE)
  )
}))
```

```{r plot_power6}
matplot(
  pw_df6$delta_A,
  pw_df6[, c("power_overall", "power_sgA", "power_sgB")],
  type = "b", pch = 19, lty = 1,
  col  = c("black", "steelblue", "tomato"),
  xlab = "True delta_A (subgroup A effect size)",
  ylab = "Empirical power (alpha = 0.05)",
  main = "Overall and subgroup power (delta_B = 0.15 fixed)",
  ylim = c(0, 1)
)
legend("topleft",
       legend = c("Overall", "Subgroup A", "Subgroup B"),
       col = c("black", "steelblue", "tomato"), lty = 1, pch = 19)
abline(h = 0.80, lty = 2, col = "grey50")
```

## Next steps

- [Example 7](example-7.html) - Bayesian Go/No-Go decision rule with
  historical borrowing
- [Endpoint Types](endpoints.html) - modeling binary and TTE endpoints
