---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

This vignette walks through a complete rxsim simulation from start to finish:
defining a trial scenario, building arm populations, registering an analysis
trigger, running replicates, and collecting results. By the end you will have
seen the full rxsim workflow in one place and will be ready to adapt it to your
own design. For deeper explanations of each building block, see the
[Core Concepts](concepts.html) vignette.

## Load the package

```{r setup}
library(rxsim)
```

## Step 1 — Define the scenario

The `scenario` tibble captures the design parameters you want to track in your
results — sample size, allocation ratios, or any other factors you might vary
across simulation runs. Wrapping them in `tidyr::expand_grid()` produces a
one-row data frame that gets embedded in every analysis result. When you later
stack results across many scenarios or sensitivity analyses, this metadata keeps
each row traceable back to its design.

```{r scenario}
sample_size <- 30
arms        <- c("control", "treatment")
allocation  <- c(1, 1)
true_delta  <- 0.5

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation  = list(allocation),
  true_delta  = true_delta
)
```

## Step 2 — Define arm populations

Each trial arm is represented by a **generator function** — a plain R function
that takes `n` (the number of subjects planned for that arm) and returns a data
frame of subject-level endpoint data. rxsim calls these functions once per
replicate to draw a fresh population, so every replicate gets independent data.

The data frame must contain at least three columns:

- `id` — a unique integer identifier per subject
- `readout_time` — the time (in trial-time units) after enrollment when the
  endpoint is observed (use `0` for baseline or immediately-available data)
- at least one endpoint column

`vector_to_dataframe()` is a convenience helper that wraps a numeric vector
into this standard format with `id`, `data`, and `readout_time = 0`.

Here we simulate a simple continuous endpoint: the control arm is
standard-normal, the treatment arm has a mean shift of `true_delta`.

```{r population}
population_generators <- list(
  control   = function(n) vector_to_dataframe(rnorm(n)),
  treatment = function(n) vector_to_dataframe(rnorm(n, mean = true_delta))
)
```

## Step 3 — Define enrollment and dropout

Enrollment and dropout are modelled as random processes. Each function takes
`n` and returns a vector of **inter-event times** — the waiting times between
successive enrollments (or dropouts). These times are drawn independently for
every replicate, giving each trial its own enrollment trajectory.

`rexp(n, rate = 1)` generates exponentially-distributed inter-arrival times
with a mean gap of 1 time unit between enrolments — a common approximation for
Poisson arrivals. A lower dropout rate (`0.05`) reflects a trial where most
subjects complete the study.

```{r trial_params}
enrollment <- function(n) rexp(n, rate = 1.0)
dropout    <- function(n) rexp(n, rate = 0.05)
```

## Step 4 — Define analysis triggers

Analysis triggers are the heart of rxsim. Each trigger pairs a **condition**
(when to fire) with an **analysis function** (what to compute when it fires).

The condition is written as a `dplyr`-style boolean expression inside
`rlang::exprs()`. It is evaluated against a snapshot of all currently enrolled
subjects at each timepoint. Here the condition fires once full enrollment is
reached, i.e., `sample_size` subjects have accumulated a non-`NA` `enroll_time`.

The `!!` operator (pronounced "bang-bang") **injects the current value** of
`sample_size` into the expression at definition time, rather than looking it up
at evaluation time. This is necessary because the expression is stored and
evaluated later inside the simulation loop.

When the condition is met, rxsim calls the analysis function with two arguments:

- `df`: a data frame snapshot of all enrolled subjects at the triggering
  timepoint, with columns from the population data plus `enroll_time`,
  `drop_time`, and `arm`
- `time`: the current trial clock time at which the trigger fired

The function should return a data frame (one row per trigger event is the
conventional pattern).

```{r triggers}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) {
      enrolled <- subset(df, !is.na(enroll_time))
      data.frame(
        scenario,
        n_enrolled  = nrow(enrolled),
        mean_ctrl   = mean(enrolled$data[enrolled$arm == "control"]),
        mean_trt    = mean(enrolled$data[enrolled$arm == "treatment"]),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

## Step 5 — Create replicates and run

`replicate_trial()` generates `n` fully independent `Trial` objects. For each
replicate it:

1. Samples a fresh enrollment/dropout timeline from your functions
2. Calls each population generator to draw new subject-level data
3. Registers your analysis triggers on the trial's internal timer

`run_trials()` then executes every replicate's simulation loop in sequence.

```{r trial}
set.seed(42)

trials <- replicate_trial(
  trial_name            = "getting_started",
  sample_size           = sample_size,
  arms                  = arms,
  allocation            = allocation,
  enrollment            = enrollment,
  dropout               = dropout,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 5
)

run_trials(trials)
```

## Step 6 — Collect and interpret results

### Analysis results

After running, each `Trial` object exposes a `results` list. It is indexed
first by the **timepoint** at which an analysis fired (e.g., `"time_30.4"`),
then by the **analysis name** you gave it (here `"final"`). The value is
whatever your analysis function returned.

The `collect_results()` helper gathers every analysis from every timepoint
across all replicates into one tidy data frame:

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

Each row is one replicate. The `mean_ctrl` and `mean_trt` columns show the
arm-level sample means at the moment the final analysis fired. The `timepoint`
and `analysis` columns identify when and which analysis produced each row —
essential when a trial has both interim and final analyses. Variation across
replicates reflects both the stochastic endpoint data and the randomness in
enrollment timing; exactly what operating characteristic simulations are
designed to characterise.

### Locked data snapshots

Every time an analysis fires, rxsim also saves a **locked data snapshot**, the
full subject-level data frame at that timepoint. You can inspect it directly to
debug your analysis function, audit which subjects were enrolled, or compute
additional statistics post-hoc.

```{r locked_data}
# One snapshot per timepoint that fired in replicate 1
names(trials[[1]]$locked_data)

# First six rows of the snapshot
head(trials[[1]]$locked_data[[1]])
```

The locked data contains the population columns (`id`, `data`, `arm`,
`readout_time`) plus three columns added by rxsim:

| Column | Meaning |
|--------|---------|
| `enroll_time` | Calendar time the subject was enrolled (`NA` if not yet enrolled) |
| `drop_time` | Calendar time the subject dropped out (`NA` if still active) |
| `time` | The trial clock time at which this snapshot was taken |

## Next steps

Now that you have seen the full workflow, here is where to go next:

- **[Core Concepts](concepts.html)** — understand how `Population`, `Timer`,
  and `Trial` compose, and how to write more advanced trigger expressions
- **[Enrollment & Dropout Modeling](enrollment-dropout.html)** — choose between
  stochastic (`gen_plan`) and piecewise-constant (`gen_timepoints`) schedules
- **[Example 1](example-1.html)** through **[Example 7](example-7.html)** —
  progressively complex designs: correlated endpoints, time-to-event, multi-arm
  dose-finding, subgroup analyses, and Bayesian Go/No-Go rules
