---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup package
```{r setup}
library(rxsim)
```

## Scenario
Capture essential trial parameters. These will form scenarios.

```{r scenario}
sample_size <- 15
```

## Arm / Population / Cohort
Let's assume a continuous endpoint that follows a normal distribution.

```{r population}
control <- Population$new(
  name = "control",
  data = vector_to_dataframe(rnorm(sample_size))
)
```

## Time points
Record all time points where we expect important events to occur like:

- enrollment
- dropout
- interim analysis
- final analysis
- futility
- measurements 

```{r timepoints}
timepoints <- tibble::tribble(
 ~time,      ~arm, ~enroller, ~dropper,
     1, "control",         5L,        0L,
     2, "control",         5L,        1L,
     3, "control",         5L,        1L
)

t <- Timer$new(name = "timer")
add_timepoints(t, timepoints)
```

## Event Triggers
Specify event triggers using helper functions. For example, analyze at time = 3:

```{r timer}
trigger_by_calendar(3, t, analysis = function(df, current_time) mean(df$data))
```

Or trigger when 50% of target (15) subjects are enrolled:

```{r alt_trigger}
trigger_by_fraction(0.5, t, sample_size, analysis = function(df, current_time) mean(df$data))
```

## Collect & Run
Collect everything in the `Trial` object:

```{r trial}
trial <- Trial$new(
  name = "test_trial",
  timer = t,
  population = list(control)
)
```

That's it now we can run a single replicate using

```{r run}
trial$run()
prettify_results(trial$results)
```

You can of course also get the raw data at the specified milestone.

```{r data}
trial$locked_data
```
