---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup package
```{r setup}
library(rxsim)
```

## Scenario
Capture essential trial parameters. These will form scenarios.

```{r scenario}
sample_size <- 15
```

## Arm / Population / Cohort
Let's assume a continuous endpoint that follows a normal distribution.

```{r population}
population_generators <- list(
  control = function(n) vector_to_dataframe(rnorm(n))
)
```

## Trial Parameters
Define enrollment and dropout profiles.

```{r trial_params}
arms <- c("control")
allocation <- c(1)
enrollment <- function(n) rexp(n, rate = 1)
dropout <- function(n) rexp(n, rate = 0.1)
```

## Event Triggers
Define analysis conditions.

```{r triggers}
analysis_generators <- list(
  final = list(
    trigger = rlang::exprs(
      sum(!is.na(enroll_time)) >= !!sample_size
    ),
    analysis = function(df, timer) mean(df$data)
  )
)
```

## Collect & Run
Create multiple replicates of the trial:

```{r trial}
trials <- replicate_trial(
  trial_name = "test_trial",
  sample_size = sample_size,
  arms = arms,
  allocation = allocation,
  enrollment = enrollment,
  dropout = dropout,
  analysis_generators = analysis_generators,
  population_generators = population_generators,
  n = 3
)
```

Run all replicates:

```{r run}
run_trials(trials)
prettify_results(trials[[1]]$results)
```

You can also get the raw data from any replicate:

```{r data}
lapply(trials[[1]]$locked_data, head)
```
