---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache   = TRUE
)
```

This vignette builds a complete two-arm trial simulation step by step using
the direct API: `Population`, `Timer`, `Condition`, and `Trial` are all
constructed explicitly. This is the most transparent way to learn how
rxsim works. For the generator shortcut (`replicate_trial`) and a
side-by-side comparison of both styles, see
[Two API Styles](api-styles.html).

```{r setup}
library(rxsim)
set.seed(42)
```

## Step 1 - Design parameters

Capture the scenario parameters that you want to appear in every result
row. Storing them up front in a named list (or data frame) makes it
simple to sweep over multiple designs later.

```{r scenario}
sample_size <- 40
arms        <- c("pbo", "trt")
allocation  <- c(1, 1)          # balanced: 20 per arm
delta       <- 0.5              # true treatment effect (SD units)

scenario <- data.frame(
  sample_size = sample_size,
  delta       = delta
)
```

## Step 2 - Populations

A `Population` holds subject-level data for one arm. The data frame must
contain `id`, `readout_time`, and at least one endpoint column. Here
`readout_time = 0` means the endpoint is observed at the moment of
enrollment (instantaneous readout).

```{r populations}
n_per_arm <- sample_size / 2   # 20 subjects per arm

pbo_data <- data.frame(
  id           = seq_len(n_per_arm),
  response     = rnorm(n_per_arm, mean = 0,     sd = 1),
  readout_time = 0
)

trt_data <- data.frame(
  id           = seq_len(n_per_arm),
  response     = rnorm(n_per_arm, mean = delta, sd = 1),
  readout_time = 0
)

pop_pbo <- Population$new(name = "pbo", data = pbo_data)
pop_trt <- Population$new(name = "trt", data = trt_data)

pop_pbo$n   # 20 subjects
pop_trt$n
```

## Step 3 - Timer

The `Timer` defines the trial clock: a list of (time, arm, enroll, drop)
events. `gen_plan()` draws a stochastic enrollment schedule from
inter-arrival time distributions, and `add_timepoints()` registers it on
the Timer.

```{r timer}
tmr <- Timer$new(name = "gs_timer")

plan <- gen_plan(
  sample_size = sample_size,
  arms        = arms,
  allocation  = allocation,
  enrollment  = function(n) rexp(n, rate = 1),   # ~1 subject/time unit
  dropout     = function(n) rexp(n, rate = 0.02) # low dropout
)

add_timepoints(tmr, plan)

tmr$get_end_timepoint()   # last event time
tmr$get_n_arms()
```

## Step 4 - Condition

A `Condition` pairs a trigger expression with an analysis function. The
trigger here uses `enroll_trigger(1.0, sample_size)`: the condition fires
once all 40 subjects have been enrolled. The analysis runs a two-sample
t-test and returns a one-row data frame.

```{r condition}
final_cond <- Condition$new(
  where    = enroll_trigger(1.0, sample_size),
  name     = "final",
  analysis = function(df, current_time) {
    enrolled <- subset(df, !is.na(enroll_time))
    tt <- t.test(response ~ arm, data = enrolled)
    data.frame(
      scenario,
      n_pbo    = sum(enrolled$arm == "pbo"),
      n_trt    = sum(enrolled$arm == "trt"),
      mean_pbo = mean(enrolled$response[enrolled$arm == "pbo"]),
      mean_trt = mean(enrolled$response[enrolled$arm == "trt"]),
      p_value  = unname(tt$p.value)
    )
  }
)
```

## Step 5 - Trial

Assemble the three pieces into a `Trial` and call `run()`. The `seed`
argument ensures this specific replicate is reproducible.

```{r trial}
trial <- Trial$new(
  name       = "getting_started_1",
  seed       = 101,
  timer      = tmr,
  population = list(pop_pbo, pop_trt),
  conditions = list(final_cond)
)

trial$run()
```

Inspect the result. `trial$results` is a named list keyed by timepoint
(`"time_<t>"`). Inside each entry is the named analysis result.

```{r inspect}
prettify_results(trial$results)
```

## Step 6 - Replication

A single run is one stochastic draw. To estimate operating
characteristics (power, type I error, expected study duration), clone the
trial and run many replicates. `clone_trial(trial, n)` produces `n` deep
copies, each with an independent Timer and Population.

Note: the clone inherits the original data. For truly independent
replicates we recreate populations with fresh random draws inside the
loop. The direct API lets us do this explicitly:

```{r replicate}
n_reps <- 300

replicates <- lapply(seq_len(n_reps), function(i) {
  n_arm <- sample_size / 2

  # Fresh populations
  p_pbo <- Population$new(
    name = "pbo",
    data = data.frame(
      id = seq_len(n_arm),
      response = rnorm(n_arm, mean = 0,     sd = 1),
      readout_time = 0
    )
  )
  p_trt <- Population$new(
    name = "trt",
    data = data.frame(
      id = seq_len(n_arm),
      response = rnorm(n_arm, mean = delta, sd = 1),
      readout_time = 0
    )
  )

  # Fresh timer (new stochastic enrollment schedule)
  t <- Timer$new(name = paste0("timer_", i))
  add_timepoints(t, gen_plan(
    sample_size = sample_size,
    arms        = arms,
    allocation  = allocation,
    enrollment  = function(n) rexp(n, rate = 1),
    dropout     = function(n) rexp(n, rate = 0.02)
  ))

  # Condition (shares the same analysis function)
  cond <- Condition$new(
    where    = enroll_trigger(1.0, sample_size),
    name     = "final",
    analysis = function(df, current_time) {
      enrolled <- subset(df, !is.na(enroll_time))
      tt <- t.test(response ~ arm, data = enrolled)
      data.frame(
        scenario,
        n_pbo    = sum(enrolled$arm == "pbo"),
        n_trt    = sum(enrolled$arm == "trt"),
        mean_pbo = mean(enrolled$response[enrolled$arm == "pbo"]),
        mean_trt = mean(enrolled$response[enrolled$arm == "trt"]),
        p_value  = unname(tt$p.value)
      )
    }
  )

  Trial$new(
    name       = paste0("gs_", i),
    seed       = i,
    timer      = t,
    population = list(p_pbo, p_trt),
    conditions = list(cond)
  )
})
```

```{r run_reps, results=FALSE}
run_trials(replicates)
```

```{r collect}
results_df <- collect_results(replicates)
head(results_df)
```

## Interpreting results

`collect_results()` prepends `replicate`, `timepoint`, and `analysis`
columns to your result data. Each row is one analysis from one replicate.
The `timepoint` column is the calendar time at which the analysis fired.

Key summaries:

```{r summary}
# Empirical power at alpha = 0.05 (two-sided)
power <- mean(results_df$p_value < 0.05)
cat("Estimated power:", round(power, 3), "\n")

# Distribution of study duration (last enrollment time)
cat("Mean study duration:", round(mean(results_df$timepoint), 1), "\n")
```

## Power curve

Repeating the simulation across a range of delta values shows how power
scales with effect size. Below we sweep delta from 0 to 1 in steps of
0.2 using 200 replicates per value.

```{r power_curve, cache=TRUE}
deltas <- seq(0, 1, by = 0.2)
n_reps_sweep <- 200

power_curve <- lapply(deltas, function(d) {
  reps <- lapply(seq_len(n_reps_sweep), function(i) {
    n_arm <- sample_size / 2
    p_pbo <- Population$new(
      name = "pbo",
      data = data.frame(
        id = seq_len(n_arm),
        response = rnorm(n_arm, 0, 1),
        readout_time = 0
      )
    )
    p_trt <- Population$new(
      name = "trt",
      data = data.frame(
        id = seq_len(n_arm),
        response = rnorm(n_arm, d, 1),
        readout_time = 0
      )
    )
    t <- Timer$new(name = paste0("t_", i))
    add_timepoints(t, gen_plan(
      sample_size = sample_size, arms = arms,
      allocation  = allocation,
      enrollment  = function(n) rexp(n, rate = 1),
      dropout     = function(n) rexp(n, rate = 0.02)
    ))
    sc <- data.frame(sample_size = sample_size, delta = d)
    cond <- Condition$new(
      where    = enroll_trigger(1.0, sample_size),
      name     = "final",
      analysis = function(df, current_time) {
        enrolled <- subset(df, !is.na(enroll_time))
        tt <- t.test(response ~ arm, data = enrolled)
        data.frame(sc, p_value = unname(tt$p.value))
      }
    )
    Trial$new(name = paste0("sweep_", i), seed = i,
              timer = t,
              population = list(p_pbo, p_trt),
              conditions = list(cond))
  })
  invisible(run_trials(reps))
  res <- collect_results(reps)
  data.frame(delta = d, power = mean(res$p_value < 0.05))
})

power_df <- do.call(rbind, power_curve)
power_df
```

```{r plot_power_curve}
plot(
  power_df$delta, power_df$power,
  type  = "b", pch = 19,
  xlab  = "True effect (delta, SD units)",
  ylab  = "Empirical power",
  main  = paste0("Power curve (n = ", sample_size, " total, alpha = 0.05)"),
  ylim  = c(0, 1)
)
abline(h = 0.8, lty = 2, col = "grey50")
abline(h = 0.05, lty = 3, col = "grey50")
```

## Next steps

- [Two API Styles](api-styles.html) — see the generator API shortcut
  (`replicate_trial`) that handles Timer construction automatically
- [Core Concepts](concepts.html) — architecture overview and class
  summary table
- [Population](class-population.html), [Timer](class-timer.html),
  [Condition](class-condition.html), [Trial](class-trial.html) - class
  deep dives
- [Example 1](example-1.html) — same design via the generator API
