---
title: "Adding Interim Analyses to Fixed Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding Interim Analyses to Fixed Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(rxsim)
set.seed(42)
```

A fixed design has one analysis at the end of the trial. An interim design adds
one or more earlier looks at the same trial as enrollment progresses.

In rxsim, each look is a `Condition` object. A one-interim design therefore has
two conditions: one for the interim and one for the final. The trigger
expression decides when each condition fires, `max_triggers` controls how many
times it can fire, and `cooldown` enforces a minimum gap between repeated fires.

This vignette keeps the same two-arm design throughout:

- placebo vs treatment: `arms = c("pbo", "trt")`
- true effect: `delta = 0.5`
- total sample size: `60`
- balanced allocation: `c(1, 1)`
- stochastic enrollment: `rexp(n, rate = 1)`
- stochastic dropout: `rexp(n, rate = 0.02)`
- continuous endpoint with `readout_time = 0`
- analysis: a two-sample t-test returning arm sizes, mean difference, and
  p-value

```{r helpers}
sample_size <- 60
arms <- c("pbo", "trt")
allocation <- c(1, 1)
delta <- 0.5

make_population_generators <- function(effect = delta) {
  list(
    pbo = function(n) {
      data.frame(
        id = seq_len(n),
        value = rnorm(n, mean = 0, sd = 1),
        readout_time = 0
      )
    },
    trt = function(n) {
      data.frame(
        id = seq_len(n),
        value = rnorm(n, mean = effect, sd = 1),
        readout_time = 0
      )
    }
  )
}

enrollment_fn <- function(n) rexp(n, rate = 1)
dropout_fn <- function(n) rexp(n, rate = 0.02)

ttest_analysis <- function(df, current_time) {
  dat <- subset(df, !is.na(enroll_time))
  tt <- t.test(value ~ arm, data = dat)
  mean_pbo <- mean(dat$value[dat$arm == "pbo"])
  mean_trt <- mean(dat$value[dat$arm == "trt"])

  data.frame(
    n_pbo = sum(dat$arm == "pbo"),
    n_trt = sum(dat$arm == "trt"),
    mean_diff = mean_trt - mean_pbo,
    p_value = unname(tt$p.value)
  )
}

make_analysis_generators <- function(triggers) {
  stats::setNames(
    lapply(triggers, function(trigger) {
      list(trigger = trigger, analysis = ttest_analysis)
    }),
    names(triggers)
  )
}

add_review_times <- function(plan, calendar_times) {
  if (is.null(calendar_times) || length(calendar_times) == 0L) {
    return(plan)
  }

  extra <- expand.grid(
    time = calendar_times,
    arm = arms,
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )
  extra$enroller <- 0L
  extra$dropper <- 0L

  plan <- rbind(plan, extra[, c("time", "arm", "enroller", "dropper")])
  plan <- stats::aggregate(
    cbind(enroller, dropper) ~ time + arm,
    data = plan,
    FUN = sum
  )
  plan[order(plan$time, plan$arm), ]
}

make_single_trial <- function(
  conditions,
  calendar_times = NULL,
  effect = delta
) {
  plan <- gen_plan(
    sample_size = sample_size,
    arms = arms,
    allocation = allocation,
    enrollment = enrollment_fn,
    dropout = dropout_fn
  )
  plan <- add_review_times(plan, calendar_times)
  arm_sizes <- tapply(plan$enroller, plan$arm, sum)
  pop_gens <- make_population_generators(effect)

  timer <- Timer$new(name = "manual_timer")
  add_timepoints(timer, plan)

  Trial$new(
    name = "manual_trial",
    timer = timer,
    population = lapply(names(pop_gens), function(name) {
      gen_population(
        name = name,
        generator = pop_gens[[name]],
        sample_size = as.integer(arm_sizes[[name]])
      )
    }),
    conditions = conditions
  )
}

make_replicates <- function(triggers, n_reps, effect = delta) {
  replicate_trial(
    trial_name = "interim_trial",
    sample_size = sample_size,
    arms = arms,
    allocation = allocation,
    enrollment = enrollment_fn,
    dropout = dropout_fn,
    analysis_generators = make_analysis_generators(triggers),
    population_generators = make_population_generators(effect),
    n = n_reps
  )
}
```

## Fixed design baseline

Start with the usual fixed design: one `Condition` at full enrollment. With the
generator API, that means one named entry in `analysis_generators`.

```{r fixed-trials}
fixed_triggers <- list(
  final = enroll_trigger(1.0, sample_size)
)

trials_fixed <- make_replicates(fixed_triggers, n_reps = 300L)
```

```{r fixed-run, results=FALSE}
run_trials(trials_fixed)
```

```{r fixed-results}
fixed_results <- collect_results(trials_fixed)
str(fixed_results)

head(
  fixed_results[, c(
    "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
    "mean_diff", "p_value"
  )]
)
```

The result has one row per replicate and one analysis label, `"final"`.

```{r fixed-power}
mean(fixed_results$p_value < 0.05)
```

That estimate is the fixed-design power from 300 replicates under `delta = 0.5`.

## Adding one interim

To add an interim to a fixed design, keep the final `Condition` and add one more
for the earlier look. Here we use the direct `Trial` API so the two conditions
are explicit in `conditions = list(interim_cond, final_cond)`.

```{r one-interim-trial}
interim_cond <- Condition$new(
  where = enroll_trigger(0.5, sample_size),
  analysis = ttest_analysis,
  name = "interim"
)

final_cond <- Condition$new(
  where = enroll_trigger(1.0, sample_size),
  analysis = ttest_analysis,
  name = "final"
)

trial_one_interim <- make_single_trial(
  conditions = list(interim_cond, final_cond)
)
```

Both conditions use the default `max_triggers = 1L`, so each look fires once.

```{r one-interim-run, results=FALSE}
trial_one_interim$run()
```

```{r one-interim-results}
one_interim_results <- collect_results(trial_one_interim)
one_interim_results[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)]
```

`collect_results()` now returns one row for `"interim"` and one row for
`"final"`.

```{r one-interim-final-only}
collect_results(trial_one_interim, analysis = "final")[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)]
```

Filtering by `analysis = "final"` is useful when a multi-look design should be
summarised as a fixed design at the last look.

## Three looks: 33%, 67%, 100%

For more looks, add more named trigger entries. Each name becomes a value in the
`analysis` column of `collect_results()`.

```{r three-look-trials}
three_look_triggers <- list(
  look_1 = enroll_trigger(1 / 3, sample_size),
  look_2 = enroll_trigger(2 / 3, sample_size),
  final = enroll_trigger(1.0, sample_size)
)

trials_three_looks <- make_replicates(three_look_triggers, n_reps = 8L)
```

```{r three-look-run, results=FALSE}
run_trials(trials_three_looks)
```

```{r three-look-results}
three_look_results <- collect_results(trials_three_looks)
head(
  three_look_results[, c(
    "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
    "mean_diff", "p_value"
  )],
  9
)

table(three_look_results$replicate)
```

Each replicate contributes three rows: `look_1`, `look_2`, and `final`.

## Calendar-triggered interim

Calendar looks are useful when a review must happen on a date rather than at a
sample fraction. Here the interim fires at calendar time 8, while the final
still waits for full enrollment.

For calendar looks, the calendar times need to appear in the trial clock. The
helper `add_review_times()` adds zero-event timepoints so the condition can be
evaluated at time 8.

```{r calendar-trial}
calendar_cond <- Condition$new(
  where = calendar_trigger(8),
  analysis = ttest_analysis,
  name = "interim"
)

calendar_final_cond <- Condition$new(
  where = enroll_trigger(1.0, sample_size),
  analysis = ttest_analysis,
  name = "final"
)

trial_calendar <- make_single_trial(
  conditions = list(calendar_cond, calendar_final_cond),
  calendar_times = 8
)
```

```{r calendar-run, results=FALSE}
trial_calendar$run()
```

```{r calendar-results}
calendar_results <- collect_results(trial_calendar)
calendar_results[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)]
```

Because enrollment is stochastic, the calendar interim may happen before or
after 50% of subjects are enrolled. The trigger is based on time, not on the
enrollment fraction.

## Combining triggers

You can combine trigger primitives with `&` when both conditions must be true.
The expression below waits for 50% enrollment and also requires the trial clock
to be at one of the scheduled review times.

```{r combined-trial}
combined_cond <- Condition$new(
  where = enroll_trigger(0.5, sample_size) &
    calendar_trigger(c(4, 8, 12, 16)),
  analysis = ttest_analysis,
  name = "interim"
)

combined_final_cond <- Condition$new(
  where = enroll_trigger(1.0, sample_size),
  analysis = ttest_analysis,
  name = "final"
)

trial_combined <- make_single_trial(
  conditions = list(combined_cond, combined_final_cond),
  calendar_times = c(4, 8, 12, 16)
)
```

```{r combined-run, results=FALSE}
trial_combined$run()
```

```{r combined-results}
collect_results(trial_combined)[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)]
```

This pattern protects against very early looks caused by unusually fast
enrollment. In this design, the review times 4, 8, 12, and 16 are usually too
early for 50% enrollment, so the interim often does not fire. The same pattern
is still useful when the review grid overlaps the expected enrollment window.

## Monitoring with cooldown

A monitoring condition can re-fire many times. Set `max_triggers = Inf` and use
`cooldown` to enforce a minimum gap between successive evaluations.

Here the monitor starts once any subjects are enrolled and is checked on a
regular review grid. The analysis returns the condition's `trigger_count` so you
can see how often it fired.

```{r monitoring-trial}
monitor_cond <- NULL
monitor_cond <- Condition$new(
  where = count_trigger("enroll_time", ">=", 1) &
    calendar_trigger(seq(2, 40, by = 2)),
  analysis = function(df, current_time) {
    data.frame(
      n_enrolled = sum(!is.na(df$enroll_time)),
      trigger_count = monitor_cond$trigger_count + 1L
    )
  },
  name = "monitor",
  cooldown = 2,
  max_triggers = Inf
)

trial_monitor <- make_single_trial(
  conditions = list(monitor_cond),
  calendar_times = seq(2, 40, by = 2)
)
```

```{r monitoring-run, results=FALSE}
trial_monitor$run()
```

```{r monitoring-results}
monitor_results <- collect_results(trial_monitor)
head(
  monitor_results[, c(
    "replicate", "timepoint", "analysis", "n_enrolled", "trigger_count"
  )]
)

tail(monitor_results$trigger_count, 1)
```

This is a simple template for ongoing monitoring. In a richer design, the same
pattern can support safety checks, data quality checks, or repeated futility
assessments.

## Collecting and filtering results

The most common post-processing step is to collect all looks together and then
filter by `analysis`.

```{r compare-trials}
compare_triggers <- list(
  interim = enroll_trigger(0.5, sample_size),
  final = enroll_trigger(1.0, sample_size)
)

trials_compare <- make_replicates(compare_triggers, n_reps = 300L)
```

```{r compare-run, results=FALSE}
run_trials(trials_compare)
```

```{r compare-results}
compare_all <- collect_results(trials_compare)
compare_final <- collect_results(trials_compare, analysis = "final")
compare_interim <- collect_results(trials_compare, analysis = "interim")

head(compare_all[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)])

head(compare_final[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)])

head(compare_interim[, c(
  "replicate", "timepoint", "analysis", "n_pbo", "n_trt",
  "mean_diff", "p_value"
)])
```

```{r compare-power}
data.frame(
  analysis = c("interim", "final"),
  power = c(
    mean(compare_interim$p_value < 0.05),
    mean(compare_final$p_value < 0.05)
  )
)
```

This makes it easy to compare operating characteristics by look without
rewriting the analysis function.

## Power curve

Finally, use the same interim and final design across a grid of treatment
effects. The only thing that changes is the treatment arm mean.

```{r power-curve-data, results=FALSE}
delta_grid <- seq(0, 1, by = 0.25)

estimate_power_curve <- function(effect) {
  trials <- make_replicates(compare_triggers, n_reps = 500L, effect = effect)
  run_trials(trials)
  res <- collect_results(trials)
  out <- aggregate(p_value ~ analysis, data = res, function(x) mean(x < 0.05))
  data.frame(
    delta = effect,
    analysis = out$analysis,
    power = out$p_value
  )
}

power_curve <- do.call(
  rbind,
  lapply(delta_grid, estimate_power_curve)
)
```

```{r power-curve-plot}
power_curve

ggplot2::ggplot(
  power_curve,
  ggplot2::aes(x = delta, y = power, color = analysis)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_continuous(breaks = delta_grid) +
  ggplot2::labs(
    x = "True treatment effect (delta)",
    y = "Power",
    color = "Look"
  )
```

The final look is more powerful because it uses the full sample. The interim
look is earlier, so it trades information for speed.

## Next steps

- [Condition](class-condition.html): trigger definitions, `cooldown`, and
  `max_triggers`
- [Example 2](example-2.html): a compact two-look worked example
- [Two API Styles](api-styles.html): direct `Trial$new()` versus
  `replicate_trial()`
